### 2017-01-27 manuscript/phylotyper.tex
Starting manuscript branch for tracking and editing

\documentclass{bioinfo}
\copyrightyear{2017} \pubyear{2017}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Application Note}

\usepackage{url}

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[short Title]{Phylotyper: In silico predictor of bacterial strain types from gene sequences}
\author[Whiteside \textit{et~al}.]{Matthew D. Whiteside\,$^{\text{\sfb 1,}*}$, Chad R. Laing\,$^{\text{\sfb 1}}$ and Victor P.J. Gannon\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$National Microbiology Laboratory, Public Health Agency of Canada, Lethbridge, AB, Canada, T1J 3Z4}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text
Text Text Text Text Text.\\
\textbf{Results:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}

http://genomemedicine.biomedcentral.com/articles/10.1186/s13073-014-0090-6

https://holtlab.net/2014/12/27/behind-the-paper-srst2-for-short-read-sequence-typing-of-bacterial-pathogens/

http://jcm.asm.org/content/53/8/2410.full

http://mgen.microbiologyresearch.org/content/journal/mgen/10.1099/mgen.0.000064

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4876609/

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4508412/

Transition to WGS.  Need rapid / accurate prediction of typing

Subtyping schemes have been used as as indicators. As a survillence tool, subtypes

Current sequence-based predictors of bacterial strain types use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known types.
Type assignment is based on sequence similarity with the type annotation from the top match above a pre-selected sequence similarity score threshold being used to assign a type to the unknown query.
In current in silico type tools, the broader context and rate of sequence mutation within and between types is not built into the predictive framework.
Sequence similarity scores are not direct indicators of the level of uncertainty assocated with a type prediction.

Phylotyper is a novel in silico predictor of subtypes from sequence data. 
It builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences.  
Using phylogenetic ancestral reconstruction to assign likelihoods of each type to the branch points in the tree and also compute the transition rates between subtypes, Phylotyper can assign an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
Subtype information is mainly used as a proxy for evolutionarily-related bacterial strain groups or to infer phenotypes.
A phylogenetic framework for predicting subtypes is more consistent with the uses of subtype information.

\section{The Phylotyper Approach}

The core of Phylotyper, is an ancestral state reconstruction method that has been adapted for hidden state prediction.
In phylogenetics, ancestral state reconstruction involves the prediction of traits of ancestors from extant descendants.
This methodology can be extended to predict properties in existing strains under investigation for which the a properties' state is unknown.
Ancestral state reconstruction (ASR) has been successfully applied to hidden state prediction in the field of microbial metagenomics; the tool PICRUSt uses ASR to estimate the gene family content contributed by a bacteria to a metagenomic sample.

In Phylotyper, the \texttt{rerootingMethod} function from the phytools R package is used to perform the ASR.
It is based on a method originally described in Yang \emph{et al}. (1995) for estimating the marginal likelihood of a discrete set of states for the internal nodes in a phylogenetic tree.
The phytools \texttt{rerootingMethod} function was selected over alternatives, such as APE ace, because it can handle extant tip nodes with unkown states in the phylogenetic tree and also compute posterior probability for those nodes, hence, making applicable for hidden state prediction.
The maximum marginal likelihood (also called emperical Bayesian posterior probability) calculated for unknown tip nodes reflects the most likely state for the node given the emperically estimated evolution model and phylogeny.
It captures the uncertainty associated with phylogenetic branch lengths and rates of change associated with the states.
In the context of Phylotyper, the posterior probabilities provides a confidence value associated with a predicted subtype.
The evolutionary model used in \texttt{rerootingMethod} is the $\text{M}_{\text{k}}$ model for discrete states (a continuous-time Markov process model)).  
It is estimated from the data, so in Phylotyper, the reference set of genes used to build a phylogenetic tree and run the ASR, is key to the performance of the predictor.
For subtype schemes packaged in Phylotyper, the annotated genes will be maintained and routinely updated to provide a robust and comprehensive reference set.
The other key assumption is that gene phylogeny is correlated with the evolution of subtype state.
Because of these assumptions are central to the method, all new subtype schemes added into Phylotyper are evaluated for their predictive performance.

Phylotyper is developed in python and R. 
Outlined below are the steps and tools used in the Phylotyper pipeline:

\begin{enumerate}
\item Identify subtype gene loci in input genomes using blastn or blastx.
\item Align input genes with unknown sunbtype against a pre-aligned set of reference genes using the tool MAFFT's --add feature.
\item If multiple loci are involved, concatenate individual alignments into superalignment.
\item Generate phylogenetic tree of aligned genes with FastTree.
\item Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes. 
Genes with unknown subtype are assigned a flat prior.
\item Identify the subtype with maximum marginal likelihood for the unknown genes and report to user.
Users are also provided with a image of the phylogenetic tree that shows the position of the unknown genes. 
An example is shown in Figure~1\vphantom{\ref{fig:01}}.
\end{enumerate}

Phylotyper is available for download from: \url{https://github.com/superphy/insilico-subtyping}.

\section{Functionality in the Phylotyper Tool}

Phylotyper was designed to be incorporated into a whole genome sequencing public health workflow.  
The main input into Phylotyper is assembled (but not necessarily closed) genome sequences.  
Putative loci needed for the selected subtype scheme are identified in the input genomes using BLAST.
The discovered loci are then sent to the Phylotyper subtype prediction module.
It is possible to bypass the loci search step and provide genes directly to Phylotyper using the subtype run mode.

The Phylotyper approach can be used to predict any biological property that can be inferred from the phylogenetic distribution of a nominal set of genomic loci.
Currently, subtype schemes for \emph{Escherichia coli} Shiga-toxin 1 and 2 subtypes, Serotype O and H-types and Intimin subtypes are available in the Phylotyper package.
Details about the current schemes available in the Phylotyper package is provided in table XXX.
However, included in the Phylotyper software, is the capability to add new subtype schemes. 
Creating a new subtype scheme will save the required reference files in a data directory, allowing the newly added schemes to be easily run from Phylotyper. 
We also encourage users to contribute their subtype schemes to the main software repository (please send us subtype schemes you have developed and wish to share).

Checks are built-in new subtype scheme pipeline to ensure the main assumptions of the Phylotyper approach are not violated.  
A scan of the reference phylogenetic tree is conducted to search for tightly clustered subclades in the tree that have divergent subtypes.
We first estimate the distribution of the inter-patristic distances of genes with the same subtype.  
The distribution is used to dynamically select a distance threshold, which we employ to flag subclades wherein the inter-patristic distance is less than  the threshold but contains multiple distinct subtypes.
We also conduct a leave-one-out cross validation for each gene in the new subtype scheme to assess the scheme's predictive capability.
Users are alerted if their subtype schemes that have a F1-score below 0.9 from the cross-validation analysis.

Phylotyper also offers flexibility in the parameterization of evolution model used in the ASR step.
A component of the underlying ASR framework, is an Mk or markov model of subtype evolution, for which an emperical transition rate matrix is estimated from the data.
The transition matrix is used to calculate the expected number of subtype state changes given a distance in the phylogenetic tree.  
Different model parameterizations can be defined for the transition rate matrix. 
The simplest parameterization available in Phylotyper is the equal rates model; all subtypes have the same forward and reverse rate.  
The most complex parameterization available im Phylotyper is the symmetric model, wherein each forward and reverse rate for a given pair of subtypes are assigned a separate parameter.
Frequently, the number of subtypes makes the symmetric model too computationally prohibitive.
To offer more flexible models in these situations with reduced numbers of free parameters, two custom parameterization approaches were developed.
The custom approaches both use a binning strategy that attempts to identify sets of subtypes that would have similar rates and assign them a single parameter as a set.
These approaches are described in detail in the Supplementary Information XXX.
Each of these model parameterizations; equal, symmetric and the two custom models are tested and evaluated in new subtype pipeline.
The parameterization that has highest accuracy (based on a LOOCV analysis) is selected.
In the case of ties, the model with the fewest parameters is given precidence (the symmetric model is not tested when the number of subtypes is over 10)

In testing the predictive capability of Phylotyper, we found that in some cases, jointly using multiple loci increased the subtype prediction accuracy.  
And so, in Phylotyper, multiple loci can be used in creating subtype schemes.
The loci will be independently identified in the genome using BLAST and then independently aligned.  
The individual loci alignments are then concatenated to form a single superalignment that is used to build the predictive phylogenetic tree.  
The Stx1 and Stx2 schemes are examples of a multi-loci subtype schemes.  
The A and B subunit genes that make up the holo-toxin in Shiga-toxin are the inputs in these subtype schemes.

\section{Comparison to a Sequence Similarity Approach}

To assess the performance of the phylogenetic-based method used in Phylotyper and compare it to a direct sequence similarity-based approach, we ran a leave-one-out cross validation analysis. 
Each reference gene with experimentally validated subtype was removed from the training dataset and used as a test input into the Phylotyper program. 
Similarily, using a mock sequence-similarity based tool we developed for assessment purposes, each gene was witheld from the BLAST reference database and used as input (Source code is available in TODO).
The sequence-similarity tool assigns putative subtypes using a BLAST sequence search.  
The predicted subtype annotation for the test input is assigned from top BLAST hit, provided the top hit passes a percent identity, e-value and alignment coverage minimum thresholds.  
The threshold values are listed in TODO.
This approach of transfering annotations from genes with highest sequence similarity is the core of most current subtying methods.

\section{Discussion}

Phylotyper can...
New subtypes,
Performance test highly accurate.

Sequence does not develop a model of subtype evolution. 
Subtypes can evolve and different rates. 
Strict one-size fits all cutoffs will not work.  
The phylogenetic framework provides natural statistical interpretation of confidence values.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}

Text Text text\vspace*{-10pt}

\section*{Acknowledgements}

text text text\vspace*{-12pt}

\section*{Funding}

This work has been supported by the... \vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\begin{thebibliography}{}

\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

\bibitem[Bag {\it et~al}., 2001]{Bag01}
Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
in nigral dopaminergic neurnol cell: correlation with the known
pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
Res.}, \textbf{110}(Suppl. 1), 76--84.

\bibitem[Lehmann, 1986]{Leh86}
Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

\bibitem[Crenshaw and Jones, 2003]{Cre03}
Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
cancer management: one tumor, one chip. \textit{Bioinformatics},
doi:10.1093/bioinformatics/btn000.

\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
???--???.

\bibitem[Bardet, 1920]{Bar20}
Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
polydactylie et retinite pigmentaire (contribution a l'etude des
formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
institution, Paris, France.

\end{thebibliography}
\end{document}

### 2017-01-27 manuscript/phylotyper.tex
Working on intro

@@
 
-\title[short Title]{Phylotyper: In silico predictor of bacterial strain types from gene sequences}
+\title[short Title]{Phylotyper: In silico predictor of bacterial subtypes from gene sequences}
 \author[Whiteside \textit{et~al}.]{Matthew D. Whiteside\,$^{\text{\sfb 1,}*}$, Chad R. Laing\,$^{\text{\sfb 1}}$ and Victor P.J. Gannon\,$^{\text{\sfb 1,}*}$}
@@
 Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
-\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
-\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
+\textbf{Availability:} Phylotyper is available for download from: \url{https://github.com/superphy/insilico-subtyping}.\\
+\textbf{Contact:} \href{matthew.whiteside@phac-aspc.gc.ca}{matthew.whiteside@phac-aspc.gc.ca}\\
 \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
@@
 
-Transition to WGS.  Need rapid / accurate prediction of typing
+Whole-genome sequencing (WGS) is a transformative technology.
+In public health, the speed, discriminatory power and broad utility of WGS data promises to improve surviellence and outbreak analysis.
+Adoption of WGS in the public health, however, requires bridging of historical data and systems with new the technologies and data.
+One of the workhorse methods used in public health are phenotypic subtyping (such as serotyping).
+As a survillence tool, subtypes provide an straight-forward designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
+A WGS-based approach to subtyping would have several benefits over current subtype systems; improved discrimination, cheaper and easier to maintain and routinely run, and would be faster.
+Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data.
 
-Subtyping schemes have been used as as indicators. As a survillence tool, subtypes
-
-Current sequence-based predictors of bacterial strain types use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known types.
-Type assignment is based on sequence similarity with the type annotation from the top match above a pre-selected sequence similarity score threshold being used to assign a type to the unknown query.
-In current in silico type tools, the broader context and rate of sequence mutation within and between types is not built into the predictive framework.
-Sequence similarity scores are not direct indicators of the level of uncertainty assocated with a type prediction.
+The current \textit{in silico} predictors of bacterial subtypes use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known subtypes.
+Subtype assignment is based on sequence similarity with the subtype annotation from the top match above a pre-selected sequence similarity threshold being used to assign a subtype to the unknown query.
+These tools do not examine the phylogenetic context or consider rate of sequence mutation within and between subtypes.
+Sequence similarity scores are not direct indicators of the level of uncertainty assocated with a subtype prediction.
 
@@
 It builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences.  
-Using phylogenetic ancestral reconstruction to assign likelihoods of each type to the branch points in the tree and also compute the transition rates between subtypes, Phylotyper can assign an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
+Using phylogenetic ancestral reconstruction to assign likelihoods of each type to the branch points in the tree and also compute the transition rates between subtypes, Phylotyper assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
 Subtype information is mainly used as a proxy for evolutionarily-related bacterial strain groups or to infer phenotypes.
-A phylogenetic framework for predicting subtypes is more consistent with the uses of subtype information.
+A phylogenetic framework for predicting subtypes is more consistent with the main uses of subtype information.
 
@@
 \item Identify subtype gene loci in input genomes using blastn or blastx.
-\item Align input genes with unknown sunbtype against a pre-aligned set of reference genes using the tool MAFFT's --add feature.
+\item Align input genes with unknown sunbtype against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature.
 \item If multiple loci are involved, concatenate individual alignments into superalignment.
@@
 \end{enumerate}
-
-Phylotyper is available for download from: \url{https://github.com/superphy/insilico-subtyping}.
 

### 2017-01-30 manuscript/phylotyper.tex
Added intro, discussion and .bib file

@@
 
-\title[short Title]{Phylotyper: In silico predictor of bacterial subtypes from gene sequences}
+\title[short Title]{Phylotyper: In silico predictor of molecular subtypes from gene sequences}
 \author[Whiteside \textit{et~al}.]{Matthew D. Whiteside\,$^{\text{\sfb 1,}*}$, Chad R. Laing\,$^{\text{\sfb 1}}$ and Victor P.J. Gannon\,$^{\text{\sfb 1,}*}$}
@@
 
-\abstract{\textbf{Motivation:} Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text
-Text Text Text Text Text.\\
-\textbf{Results:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
+\abstract{\textbf{Motivation:} Whole genome sequencing is being adopted in public health for improved surviellence and outbreak analysis.
+Molecular subtyping has been used in public health to infer phenotypes and flag high-risk bacterial strain groups.
+In silico tools that predict molecular subtypes from WGS data are needed to transition historical data and systems to WGS-based protocols.\\
+\textbf{Results:} Phylotyper is a novel solution for in silico molecular subtype prediction from WGS data.
+It is a general prediction tool that can be applied to most molecular subtype schemes.
+Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
+The phylogenic framework in Phylotyper, improves accuracy, provides useful contextual feedback and confidence scores, and is more capable at identifying novel subtypes over in silico approaches based on sequence similarity.\\
 \textbf{Availability:} Phylotyper is available for download from: \url{https://github.com/superphy/insilico-subtyping}.\\
@@
 
-http://genomemedicine.biomedcentral.com/articles/10.1186/s13073-014-0090-6
-
-https://holtlab.net/2014/12/27/behind-the-paper-srst2-for-short-read-sequence-typing-of-bacterial-pathogens/
-
-http://jcm.asm.org/content/53/8/2410.full
-
-http://mgen.microbiologyresearch.org/content/journal/mgen/10.1099/mgen.0.000064
-
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4876609/
-
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4508412/
-
-Whole-genome sequencing (WGS) is a transformative technology.
-In public health, the speed, discriminatory power and broad utility of WGS data promises to improve surviellence and outbreak analysis.
+Whole-genome sequencing (WGS) is transforming the public health field by providing a highly efficent tool for interogating bacterial populations.
+The speed, discriminatory power and broad utility of WGS data can improve surviellence and outbreak analysis.
 Adoption of WGS in the public health, however, requires bridging of historical data and systems with new the technologies and data.
-One of the workhorse methods used in public health are phenotypic subtyping (such as serotyping).
+One of the workhorse methods used in public health are molecular subtyping (such as serotyping).
 As a survillence tool, subtypes provide an straight-forward designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
-A WGS-based approach to subtyping would have several benefits over current subtype systems; improved discrimination, cheaper and easier to maintain and routinely run, and would be faster.
+A WGS-based approach to subtyping would have several benefits over current subtype systems; would be faster, have improved discrimination and would be cheaper and easier to maintain and routinely run.
 Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data.
 
-The current \textit{in silico} predictors of bacterial subtypes use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known subtypes.
+Many of the current \textit{in silico} predictors of bacterial subtypes use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known subtypes.
 Subtype assignment is based on sequence similarity with the subtype annotation from the top match above a pre-selected sequence similarity threshold being used to assign a subtype to the unknown query.
@@
 \begin{enumerate}
-\item Identify subtype gene loci in input genomes using blastn or blastx.
-\item Align input genes with unknown sunbtype against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature.
+\item Identify subtype gene loci in input genomes using blastn or blastx %\citep{Camacho2009}.
+\item Align input genes with unknown sunbtype against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature \citep{Katoh2013}.
 \item If multiple loci are involved, concatenate individual alignments into superalignment.
@@
 \end{enumerate}
+
+%\begin{figure}[!tpb]%figure1
+%%\centerline{\includegraphics{fig01.eps}}
+%\caption{Stx1 subtype phylogenetic tree. The posterior probability of each subtype as estimated by the Phylotyper method is displayed at each node as a pie chart.}\label{fig:01}
+%\end{figure}
 
@@
 
+\begin{table}[!t]
+\processtable{Subtype schemes available in Phyloyper\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule Name &
+Description & Species & Loci\\\midrule
+stx1 & Shiga-toxin 1 subtype & {\it E. coli} & 2 \\
+stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
+eae & Intimin subtype & {\it E. coli} & 1\\
+flic & H-serotype based on flagellin gene & {\it E. coli} & 1\\
+wz & O-serotype based on wzx and wzy genes & {\it E. coli} & 2\\\botrule
+\end{tabular}}{}
+\end{table}
+
 \section{Comparison to a Sequence Similarity Approach}
@@
 Each reference gene with experimentally validated subtype was removed from the training dataset and used as a test input into the Phylotyper program. 
-Similarily, using a mock sequence-similarity based tool we developed for assessment purposes, each gene was witheld from the BLAST reference database and used as input (Source code is available in TODO).
+Similarily, using a mock sequence-similarity based tool we developed for assessment purposes, each gene was witheld from the BLAST reference database and used as input (Source code is also available in the phyloyper git repository).
 The sequence-similarity tool assigns putative subtypes using a BLAST sequence search.  
@@
 
+% \begin{table}[!t]
+% \processtable{Results of a Leave-one-out cross validation analysis of Phylotyper and a sequence-similarity approach to predicting subtypes for four different schemes \label{Tab:01}} {
+% \begin{tabular}{@{}lllllll@{}}
+% \toprule 
+% \multirow{2}{*}{Scheme} & \multicolumn{3}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
+% & Precision & Recall & F$_{1}$ Score & Precision & Recall & F$_{1}$ Score \\
+% \midrule
+% {\it E. coli} Stx1 & row1 & row1 & row1 & row1 & row1 & row1\\
+% {\it E. coli} Stx2 & row2 & row2 & row2 & row2 & row2 & row2\\
+% {\it E. coli} Intimin & row3 & row3 & row3 & row3 & row3 & row3\\
+% {\it E. coli} H-serotype & row4 & row4 & row4 & row4 & row4 & row4\\
+% {\it E. coli} O-serotype & row4 & row4 & row4 & row4 & row4 & row4\\\botrule
+% \end{tabular}}{This is a footnote}
+% \end{table}
+
 \section{Discussion}
 
-Phylotyper can...
-New subtypes,
-Performance test highly accurate.
+From assembled WGS data, Phylotyper can assign unclassifed strains a molecular subtype.
+Currently the Phylotyper software offers several subtyping schemes for \textit{E. coli}.
+It can be applied to most molecular subtype schemes, however.
+To add a new scheme, it requires a reference set of loci sequences that describe the subtype scheme profile.
+The Phylotyper suite includes functionality to build and run new schemes by users.
+Performance testing showed that the Phylotyper method is highly accurate.
 
-Sequence does not develop a model of subtype evolution. 
-Subtypes can evolve and different rates. 
-Strict one-size fits all cutoffs will not work.  
-The phylogenetic framework provides natural statistical interpretation of confidence values.
+Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences. 
+The model can accomodate subtypes that evolve at different rates; a feature that would be difficult in a sequence similarity top-hit approach. 
+By estimating the phylogenetic distribution of each subtype in a particular scheme, Phylotyper more easily identify a novel subtype or accurately classify a novel sequence allele.
+The performance of sequence similarity approaches is more dependent on the reference database containing the exact allele.  
+They do not provide straight-forward mechanism to interpret novel alleles that are not in the database.
+The phylogenetic framework and empirical evolution model provides a statistical confidence value for interpreting novel alleles.
+
+To exand the Phylotyper program suite, we are working on developing functionality that will automatically identify loci in a whole genome sequence that is highly evolutionarily correlated with a target subtype.
 
@@
 
-%\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
+\bibliographystyle{natbib}
+% \bibliographystyle{achemnat}
+% \bibliographystyle{plainnat}
+% \bibliographystyle{abbrv}
 %\bibliographystyle{bioinformatics}
-%
+
 %\bibliographystyle{plain}
-%
-%\bibliography{Document}
 
-\begin{thebibliography}{}
+\bibliography{phylotyper}
 
-\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
+%\begin{thebibliography}{}
 
-\bibitem[Bag {\it et~al}., 2001]{Bag01}
-Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
+% \bibitem[Inouye \textit{et~al}., 2014]{Inouye14}
+% Inouye,M. \textit{et~al}. (2014) SRST2: Rapid genomic surveillance for public health and hospital microbiology labs. \textit{Genome Med.}, \textbf{6}, 90.
 
-\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-in nigral dopaminergic neurnol cell: correlation with the known
-pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-Res.}, \textbf{110}(Suppl. 1), 76--84.
+% \bibitem[Joensen \textit{et~al}., 2015]{Joensen15}
+% Joensen,K.G. \textit{et~al}. (2015) Rapid and Easy In Silico Serotyping of Escherichia coli Isolates by Use of Whole-Genome Sequencing Data. \textit{J. Clin. Microbiol.}, \textbf{53}, 2410-26.
 
-\bibitem[Lehmann, 1986]{Leh86}
-Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
+% \bibitem[Ingle \textit{et~al}., 2016]{Ingle16}
+% Ingle,D.J.
+\textit{et~al}.
+(2016) In silico serotyping of E.
+coli from short read data identifies limited novel O-loci but extensive diversity of O:H serotype combinations within and between pathogenic lineages.
+\textit{Microb.
+Genom.}, \textbf{2}, 7.
 
-\bibitem[Crenshaw and Jones, 2003]{Cre03}
-Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-cancer management: one tumor, one chip. \textit{Bioinformatics},
-doi:10.1093/bioinformatics/btn000.
+% \bibitem[Jenkins, 2016]{Jenkins15}
+% Jenkins C. (2015) Whole-Genome Sequencing Data for Serotyping Escherichia coli—It’s Time for a Change! \textit{J. Clin. Microbiol.},\textbf{53}, 2402-2403.
 
-\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-???--???.
+% \bibitem[Lindsey \textit{et~al}., 2016]{Lindsey16}
+% Lindsey,R.L.
+\textit{et~al}.
+(2016) Implementation of Whole Genome Sequencing (WGS) for Identification and Characterization of Shiga Toxin-Producing Escherichia coli (STEC) in the United States.
+\textit{Front.
+Microbiol.}, \textbf{7}, 766.
 
-\bibitem[Bardet, 1920]{Bar20}
-Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-polydactylie et retinite pigmentaire (contribution a l'etude des
-formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-institution, Paris, France.
-
-\end{thebibliography}
+%\end{thebibliography}
 \end{document}

### 2017-02-02 manuscript/phylotyper.tex
Fixed tables

@@
 \usepackage{url}
+\usepackage{multirow}
 
@@
 
-\title[short Title]{Phylotyper: In silico predictor of molecular subtypes from gene sequences}
+\title[short Title]{Phylotyper: {\it In silico} predictor of molecular subtypes from gene sequences}
 \author[Whiteside \textit{et~al}.]{Matthew D. Whiteside\,$^{\text{\sfb 1,}*}$, Chad R. Laing\,$^{\text{\sfb 1}}$ and Victor P.J. Gannon\,$^{\text{\sfb 1,}*}$}
@@
 
-\abstract{\textbf{Motivation:} Whole genome sequencing is being adopted in public health for improved surviellence and outbreak analysis.
+\abstract{\textbf{Summary:} Whole genome sequencing is being adopted in public health for improved surveillance and outbreak analysis.
 Molecular subtyping has been used in public health to infer phenotypes and flag high-risk bacterial strain groups.
-In silico tools that predict molecular subtypes from WGS data are needed to transition historical data and systems to WGS-based protocols.\\
-\textbf{Results:} Phylotyper is a novel solution for in silico molecular subtype prediction from WGS data.
+\textit{In silico} tools that predict molecular subtypes from WGS data are needed to transition historical data and systems to WGS-based protocols.
+Phylotyper is a novel solution for \textit{in silico} molecular subtype prediction from WGS data.
 It is a general prediction tool that can be applied to most molecular subtype schemes.
 Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
-The phylogenic framework in Phylotyper, improves accuracy, provides useful contextual feedback and confidence scores, and is more capable at identifying novel subtypes over in silico approaches based on sequence similarity.\\
-\textbf{Availability:} Phylotyper is available for download from: \url{https://github.com/superphy/insilico-subtyping}.\\
+The phylogenic framework in Phylotyper, improves accuracy, provides useful contextual feedback and confidence scores, and is more capable at identifying novel subtypes over \textit{in silico} approaches based on sequence similarity.\\
+\textbf{Availability and Implementation:} Phylotyper is a python package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
 \textbf{Contact:} \href{matthew.whiteside@phac-aspc.gc.ca}{matthew.whiteside@phac-aspc.gc.ca}\\
@@
 
-Whole-genome sequencing (WGS) is transforming the public health field by providing a highly efficent tool for interogating bacterial populations.
-The speed, discriminatory power and broad utility of WGS data can improve surviellence and outbreak analysis.
-Adoption of WGS in the public health, however, requires bridging of historical data and systems with new the technologies and data.
-One of the workhorse methods used in public health are molecular subtyping (such as serotyping).
-As a survillence tool, subtypes provide an straight-forward designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
-A WGS-based approach to subtyping would have several benefits over current subtype systems; would be faster, have improved discrimination and would be cheaper and easier to maintain and routinely run.
-Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data.
+Whole-genome sequencing (WGS) is transforming the public health field by providing a highly efficient tool for surveying bacterial populations.
+The speed, discriminatory power and broad utility of WGS data can improve surveillance and outbreak analysis.
+Adoption of WGS in the public health, however, requires bridging of historical data and systems with new the technologies and data \citep{Jenkins2015}.
+One of the workhorse methods used in public health are molecular subtyping (such as stereotyping).
+As a surveillance tool, subtypes provide an straight-forward designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens \citep{Jenkins2015}.
+A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain and routinely run \citep{Jenkins2015}.
+Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Inouye2014,Joensen2015,Ingle2016,Lindsey2016,CARRILLO2016}.
 
-Many of the current \textit{in silico} predictors of bacterial subtypes use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known subtypes.
+Many of the current \textit{in silico} predictors of bacterial subtypes use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known subtypes \citep{Joensen2015,CARRILLO2016,Ingle2016}.
 Subtype assignment is based on sequence similarity with the subtype annotation from the top match above a pre-selected sequence similarity threshold being used to assign a subtype to the unknown query.
-These tools do not examine the phylogenetic context or consider rate of sequence mutation within and between subtypes.
-Sequence similarity scores are not direct indicators of the level of uncertainty assocated with a subtype prediction.
+These tools do not examine the phylogeny context or consider rate of sequence mutation within and between subtypes.
+Sequence similarity scores are not direct indicators of the level of uncertainty associated with a subtype prediction.
 
-Phylotyper is a novel in silico predictor of subtypes from sequence data. 
+Phylotyper is a novel \textit{in silico} predictor of subtypes from sequence data. 
 It builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences.  
@@
 This methodology can be extended to predict properties in existing strains under investigation for which the a properties' state is unknown.
-Ancestral state reconstruction (ASR) has been successfully applied to hidden state prediction in the field of microbial metagenomics; the tool PICRUSt uses ASR to estimate the gene family content contributed by a bacteria to a metagenomic sample.
+Ancestral state reconstruction (ASR) has been successfully applied to hidden state prediction in the field of microbial metagenomics; the tool PICRUSt uses ASR to estimate the gene family content contributed by a bacteria to a metagenomics sample.
 
-In Phylotyper, the \texttt{rerootingMethod} function from the phytools R package is used to perform the ASR.
+In Phylotyper, the \texttt{rerootingMethod} function from the phytools R package is used to perform the ASR \citep{Revell2011}.
 It is based on a method originally described in Yang \emph{et al}. (1995) for estimating the marginal likelihood of a discrete set of states for the internal nodes in a phylogenetic tree.
-The phytools \texttt{rerootingMethod} function was selected over alternatives, such as APE ace, because it can handle extant tip nodes with unkown states in the phylogenetic tree and also compute posterior probability for those nodes, hence, making applicable for hidden state prediction.
-The maximum marginal likelihood (also called emperical Bayesian posterior probability) calculated for unknown tip nodes reflects the most likely state for the node given the emperically estimated evolution model and phylogeny.
+The phytools \texttt{rerootingMethod} function was selected over alternatives, such as APE ace, because it can handle extant tip nodes with unkown states in the phylogenetic tree and also compute posterior probability for those nodes, hence, making applicable for hidden state prediction \citep{Revell2011}.
+The maximum marginal likelihood (also called empirical Bayesian posterior probability) calculated for unknown tip nodes reflects the most likely state for the node given the empirically estimated evolution model and phylogeny.
 It captures the uncertainty associated with phylogenetic branch lengths and rates of change associated with the states.
 In the context of Phylotyper, the posterior probabilities provides a confidence value associated with a predicted subtype.
-The evolutionary model used in \texttt{rerootingMethod} is the $\text{M}_{\text{k}}$ model for discrete states (a continuous-time Markov process model)).  
+The evolutionary model used in \texttt{rerootingMethod} is the $\text{M}_{\text{k}}$ model for discrete states (a continuous-time Markov process model)) \citep{Revell2011}.  
 It is estimated from the data, so in Phylotyper, the reference set of genes used to build a phylogenetic tree and run the ASR, is key to the performance of the predictor.
@@
 \item Identify subtype gene loci in input genomes using blastn or blastx %\citep{Camacho2009}.
-\item Align input genes with unknown sunbtype against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature \citep{Katoh2013}.
+\item Align input genes with unknown subtype against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature \citep{Katoh2013}.
 \item If multiple loci are involved, concatenate individual alignments into superalignment.
-\item Generate phylogenetic tree of aligned genes with FastTree.
+\item Generate phylogenetic tree of aligned genes with FastTree \citep{Price2010}.
 \item Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes. 
-Genes with unknown subtype are assigned a flat prior.
+Genes with unknown subtype are assigned a flat prior \citep{Revell2011}.
 \item Identify the subtype with maximum marginal likelihood for the unknown genes and report to user.
@@
 The main input into Phylotyper is assembled (but not necessarily closed) genome sequences.  
-Putative loci needed for the selected subtype scheme are identified in the input genomes using BLAST.
+Putative loci needed for the selected subtype scheme are identified in the input genomes using BLAST \citep{Camacho2009}.
 The discovered loci are then sent to the Phylotyper subtype prediction module.
@@
 The Phylotyper approach can be used to predict any biological property that can be inferred from the phylogenetic distribution of a nominal set of genomic loci.
-Currently, subtype schemes for \emph{Escherichia coli} Shiga-toxin 1 and 2 subtypes, Serotype O and H-types and Intimin subtypes are available in the Phylotyper package.
-Details about the current schemes available in the Phylotyper package is provided in table XXX.
+Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) Shiga-toxin 1 and 2 subtypes, Serotype O and H-types and Intimin subtypes are available in the Phylotyper package.
+Details about the current schemes available in the Phylotyper package is provided in Table~1\vphantom{\ref{tab:01}}.
 However, included in the Phylotyper software, is the capability to add new subtype schemes. 
@@
 We also conduct a leave-one-out cross validation for each gene in the new subtype scheme to assess the scheme's predictive capability.
-Users are alerted if their subtype schemes that have a F1-score below 0.9 from the cross-validation analysis.
+Users are alerted if their subtype schemes that have a F$_\text{1}$-score below 0.9 from the cross-validation analysis.
 
 Phylotyper also offers flexibility in the parameterization of evolution model used in the ASR step.
-A component of the underlying ASR framework, is an Mk or markov model of subtype evolution, for which an emperical transition rate matrix is estimated from the data.
+A component of the underlying ASR framework, is an $\text{M}_{\text{k}}$ or Markov model of subtype evolution, for which an empirical transition rate matrix is estimated from the data.
 The transition matrix is used to calculate the expected number of subtype state changes given a distance in the phylogenetic tree.  
@@
 The simplest parameterization available in Phylotyper is the equal rates model; all subtypes have the same forward and reverse rate.  
-The most complex parameterization available im Phylotyper is the symmetric model, wherein each forward and reverse rate for a given pair of subtypes are assigned a separate parameter.
+The most complex parameterization available in Phylotyper is the symmetric model, wherein each forward and reverse rate for a given pair of subtypes are assigned a separate parameter \citep{Revell2011}.
 Frequently, the number of subtypes makes the symmetric model too computationally prohibitive.
@@
 The parameterization that has highest accuracy (based on a LOOCV analysis) is selected.
-In the case of ties, the model with the fewest parameters is given precidence (the symmetric model is not tested when the number of subtypes is over 10)
+In the case of ties, the model with the fewest parameters is given precedence (the symmetric model is not tested when the number of subtypes is over 10)
 
@@
 The predicted subtype annotation for the test input is assigned from top BLAST hit, provided the top hit passes a percent identity, e-value and alignment coverage minimum thresholds.  
-The threshold values are listed in TODO.
 This approach of transfering annotations from genes with highest sequence similarity is the core of most current subtying methods.
 
-% \begin{table}[!t]
-% \processtable{Results of a Leave-one-out cross validation analysis of Phylotyper and a sequence-similarity approach to predicting subtypes for four different schemes \label{Tab:01}} {
-% \begin{tabular}{@{}lllllll@{}}
-% \toprule 
-% \multirow{2}{*}{Scheme} & \multicolumn{3}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
-% & Precision & Recall & F$_{1}$ Score & Precision & Recall & F$_{1}$ Score \\
-% \midrule
-% {\it E. coli} Stx1 & row1 & row1 & row1 & row1 & row1 & row1\\
-% {\it E. coli} Stx2 & row2 & row2 & row2 & row2 & row2 & row2\\
-% {\it E. coli} Intimin & row3 & row3 & row3 & row3 & row3 & row3\\
-% {\it E. coli} H-serotype & row4 & row4 & row4 & row4 & row4 & row4\\
-% {\it E. coli} O-serotype & row4 & row4 & row4 & row4 & row4 & row4\\\botrule
-% \end{tabular}}{This is a footnote}
-% \end{table}
+\begin{table}[!t]
+\processtable{Results of a Leave-one-out cross validation analysis of Phylotyper and a sequence-similarity approach to predicting subtypes for four different schemes \label{Tab:01}} {
+\begin{tabular}{@{\extracolsep{2pt}}lllllll@{}}
+\toprule 
+\multirow{2}{*}{Scheme} & \multicolumn{3}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
+\cline{2-4}\cline{5-7}
+& Precision & Recall & F$_{1}$ Score & Precision & Recall & F$_{1}$ Score \\
+\midrule
+{\it E. coli} Stx1 & row1 & row1 & row1 & row1 & row1 & row1\\
+{\it E. coli} Stx2 & row2 & row2 & row2 & row2 & row2 & row2\\
+{\it E. coli} Intimin & row3 & row3 & row3 & row3 & row3 & row3\\
+{\it E. coli} H-serotype & row4 & row4 & row4 & row4 & row4 & row4\\
+{\it E. coli} O-serotype & row4 & row4 & row4 & row4 & row4 & row4\\\botrule
+\end{tabular}}{This is a footnote}
+\end{table}
 
@@
 To exand the Phylotyper program suite, we are working on developing functionality that will automatically identify loci in a whole genome sequence that is highly evolutionarily correlated with a target subtype.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Conclusion}
-
-Text Text text\vspace*{-10pt}
+This function will help streamline the process of developing new subtype schemes in Phylotyper.\vspace*{-10pt}
 
@@
 
-This work has been supported by the... \vspace*{-12pt}
+This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}
 
@@
 % \bibliographystyle{achemnat}
-% \bibliographystyle{plainnat}
+%\bibliographystyle{plainnat}
 % \bibliographystyle{abbrv}
@@
 
-%\begin{thebibliography}{}
-
-% \bibitem[Inouye \textit{et~al}., 2014]{Inouye14}
-% Inouye,M. \textit{et~al}. (2014) SRST2: Rapid genomic surveillance for public health and hospital microbiology labs. \textit{Genome Med.}, \textbf{6}, 90.
-
-% \bibitem[Joensen \textit{et~al}., 2015]{Joensen15}
-% Joensen,K.G. \textit{et~al}. (2015) Rapid and Easy In Silico Serotyping of Escherichia coli Isolates by Use of Whole-Genome Sequencing Data. \textit{J. Clin. Microbiol.}, \textbf{53}, 2410-26.
-
-% \bibitem[Ingle \textit{et~al}., 2016]{Ingle16}
-% Ingle,D.J.
-\textit{et~al}.
-(2016) In silico serotyping of E.
-coli from short read data identifies limited novel O-loci but extensive diversity of O:H serotype combinations within and between pathogenic lineages.
-\textit{Microb.
-Genom.}, \textbf{2}, 7.
-
-% \bibitem[Jenkins, 2016]{Jenkins15}
-% Jenkins C. (2015) Whole-Genome Sequencing Data for Serotyping Escherichia coli—It’s Time for a Change! \textit{J. Clin. Microbiol.},\textbf{53}, 2402-2403.
-
-% \bibitem[Lindsey \textit{et~al}., 2016]{Lindsey16}
-% Lindsey,R.L.
-\textit{et~al}.
-(2016) Implementation of Whole Genome Sequencing (WGS) for Identification and Characterization of Shiga Toxin-Producing Escherichia coli (STEC) in the United States.
-\textit{Front.
-Microbiol.}, \textbf{7}, 766.
-
-%\end{thebibliography}
 \end{document}

### 2017-02-05 manuscript/phylotyper.tex
Added supplementary info

@@
 \usepackage{url}
-\usepackage{multirow}
 
@@
 
-\abstract{\textbf{Summary:} Whole genome sequencing is being adopted in public health for improved surveillance and outbreak analysis.
+\abstract{\textbf{Summary:} Whole genome sequencing (WGS) is being adopted in public health for improved surveillance and outbreak analysis.
 Molecular subtyping has been used in public health to infer phenotypes and flag high-risk bacterial strain groups.
-\textit{In silico} tools that predict molecular subtypes from WGS data are needed to transition historical data and systems to WGS-based protocols.
-Phylotyper is a novel solution for \textit{in silico} molecular subtype prediction from WGS data.
-It is a general prediction tool that can be applied to most molecular subtype schemes.
+\textit{In silico} tools that predict molecular subtypes from gene sequences are needed to transition historical data and systems to WGS-based protocols.
+Phylotyper is a novel solution for \textit{in silico} molecular subtype prediction from gene sequences.
+Designed for incorporation into WGS pipelines, it is a general prediction tool that can be applied to most molecular subtype schemes.
 Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
-The phylogenic framework in Phylotyper, improves accuracy, provides useful contextual feedback and confidence scores, and is more capable at identifying novel subtypes over \textit{in silico} approaches based on sequence similarity.\\
+The phylogenic framework in Phylotyper improves accuracy, provides useful contextual feedback, and is more capable at identifying novel subtypes over approaches based solely on sequence similarity.\\
 \textbf{Availability and Implementation:} Phylotyper is a python package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
@@
 
-Whole-genome sequencing (WGS) is transforming the public health field by providing a highly efficient tool for surveying bacterial populations.
+Whole-genome sequencing (WGS) is transforming the public health field by providing a highly efficient method for surveying bacterial populations.
 The speed, discriminatory power and broad utility of WGS data can improve surveillance and outbreak analysis.
-Adoption of WGS in the public health, however, requires bridging of historical data and systems with new the technologies and data \citep{Jenkins2015}.
-One of the workhorse methods used in public health are molecular subtyping (such as stereotyping).
-As a surveillance tool, subtypes provide an straight-forward designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens \citep{Jenkins2015}.
+Adoption of WGS in the public health, however, requires bridging of historical data with the new technologies \citep{Jenkins2015}.
+One of the workhorse methods in public health is molecular subtyping (such as stereotyping).
+As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens \citep{Jenkins2015}.
 A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain and routinely run \citep{Jenkins2015}.
@@
 
-Many of the current \textit{in silico} predictors of bacterial subtypes use a similar approach; genes or genome sequences with unknown type are compared against a reference database of sequences with known subtypes \citep{Joensen2015,CARRILLO2016,Ingle2016}.
-Subtype assignment is based on sequence similarity with the subtype annotation from the top match above a pre-selected sequence similarity threshold being used to assign a subtype to the unknown query.
-These tools do not examine the phylogeny context or consider rate of sequence mutation within and between subtypes.
-Sequence similarity scores are not direct indicators of the level of uncertainty associated with a subtype prediction.
+Phylotyper is a novel \textit{in silico} predictor of subtypes from sequence data. 
+Phylotyper is unique in that it builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences to help inform subtype prediction. 
+Using phylogenetic ancestral reconstruction to assign the likelihood of each subtype to the branch points in the tree, Phylotyper assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
 
-Phylotyper is a novel \textit{in silico} predictor of subtypes from sequence data. 
-It builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences.  
-Using phylogenetic ancestral reconstruction to assign likelihoods of each type to the branch points in the tree and also compute the transition rates between subtypes, Phylotyper assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
-Subtype information is mainly used as a proxy for evolutionarily-related bacterial strain groups or to infer phenotypes.
-A phylogenetic framework for predicting subtypes is more consistent with the main uses of subtype information.
+\section{Implementation}
 
-\section{The Phylotyper Approach}
-
-The core of Phylotyper, is an ancestral state reconstruction method that has been adapted for hidden state prediction.
-In phylogenetics, ancestral state reconstruction involves the prediction of traits of ancestors from extant descendants.
-This methodology can be extended to predict properties in existing strains under investigation for which the a properties' state is unknown.
-Ancestral state reconstruction (ASR) has been successfully applied to hidden state prediction in the field of microbial metagenomics; the tool PICRUSt uses ASR to estimate the gene family content contributed by a bacteria to a metagenomics sample.
+The core of Phylotyper, is an ancestral state reconstruction (ASR) method that has been adapted for hidden state prediction.
+In phylogenetic analysis, ancestral state reconstruction involves the prediction of traits of ancestors from existent descendants.
+This methodology can be extended to also predict properties in a limited number of existing strains.
 
 In Phylotyper, the \texttt{rerootingMethod} function from the phytools R package is used to perform the ASR \citep{Revell2011}.
-It is based on a method originally described in Yang \emph{et al}. (1995) for estimating the marginal likelihood of a discrete set of states for the internal nodes in a phylogenetic tree.
-The phytools \texttt{rerootingMethod} function was selected over alternatives, such as APE ace, because it can handle extant tip nodes with unkown states in the phylogenetic tree and also compute posterior probability for those nodes, hence, making applicable for hidden state prediction \citep{Revell2011}.
-The maximum marginal likelihood (also called empirical Bayesian posterior probability) calculated for unknown tip nodes reflects the most likely state for the node given the empirically estimated evolution model and phylogeny.
-It captures the uncertainty associated with phylogenetic branch lengths and rates of change associated with the states.
-In the context of Phylotyper, the posterior probabilities provides a confidence value associated with a predicted subtype.
-The evolutionary model used in \texttt{rerootingMethod} is the $\text{M}_{\text{k}}$ model for discrete states (a continuous-time Markov process model)) \citep{Revell2011}.  
-It is estimated from the data, so in Phylotyper, the reference set of genes used to build a phylogenetic tree and run the ASR, is key to the performance of the predictor.
-For subtype schemes packaged in Phylotyper, the annotated genes will be maintained and routinely updated to provide a robust and comprehensive reference set.
+This function calculates the maximum marginal likelihood for unknown tip nodes in a phylogenetic tree.
+The likelihood reflects the most likely state for the node given the empirically estimated evolution model and phylogeny.
+It captures the uncertainty associated with phylogenetic branch lengths and rates of change between states.
+In the context of Phylotyper, the marginal likelihood provides a confidence value associated with a predicted subtype.
+The evolutionary model used in Phylotyper, is estimated from the data, so the reference set of genes used to build a phylogenetic tree and run the ASR, is key to the accuracy.
 The other key assumption is that gene phylogeny is correlated with the evolution of subtype state.
-Because of these assumptions are central to the method, all new subtype schemes added into Phylotyper are evaluated for their predictive performance.
+Because of these assumptions are central to the method, all new subtype schemes are evaluated for their predictive performance.
 
@@
 \begin{enumerate}
-\item Identify subtype gene loci in input genomes using blastn or blastx %\citep{Camacho2009}.
-\item Align input genes with unknown subtype against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature \citep{Katoh2013}.
+\item Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}.
+\item Align input genes against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature \citep{Katoh2013,CapellaGutierrez2009}.
 \item If multiple loci are involved, concatenate individual alignments into superalignment.
-\item Generate phylogenetic tree of aligned genes with FastTree \citep{Price2010}.
-\item Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes. 
-Genes with unknown subtype are assigned a flat prior \citep{Revell2011}.
+\item Generate maximum likelihood phylogenetic tree of aligned genes with FastTree \citep{Price2010}.
+\item Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes \citep{Revell2011}.
 \item Identify the subtype with maximum marginal likelihood for the unknown genes and report to user.
-Users are also provided with a image of the phylogenetic tree that shows the position of the unknown genes. 
-An example is shown in Figure~1\vphantom{\ref{fig:01}}.
+Users are provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g. Figure~1\vphantom{\ref{fig:01}}).
 \end{enumerate}
 
-%\begin{figure}[!tpb]%figure1
-%%\centerline{\includegraphics{fig01.eps}}
-%\caption{Stx1 subtype phylogenetic tree. The posterior probability of each subtype as estimated by the Phylotyper method is displayed at each node as a pie chart.}\label{fig:01}
-%\end{figure}
+\begin{figure}[!tpb]%figure1
+\centerline{\includegraphics{fig01.eps}}
+\caption{Phylogenetic tree for select Stx2 genes. 
+The subtype marginal likelihood is displayed at each node as a pie chart.
+The full Stx2 subtype tree is displayed in Supplementary Figure S1.}\label{fig:01}
+\end{figure}
 
-\section{Functionality in the Phylotyper Tool}
+Phylotyper was designed to be incorporated into a WGS public health workflow.  
+The main input into Phylotyper is assembled genome sequences.  
+Putative loci needed for the subtype scheme are identified in the input genomes using BLAST \citep{Camacho2009}.
+The identified loci are then sent to the Phylotyper subtype prediction module.
+It is possible in Phylotyper to use multiple loci for subtype prediction.
+Individual loci alignments are concatenated to form a single superalignment that is used to build the phylogenetic tree.  
+The Stx1 and Stx2 schemes are examples of a multi-loci subtype schemes, with toxin A and B subunit genes are the inputs in these schemes.
 
-Phylotyper was designed to be incorporated into a whole genome sequencing public health workflow.  
-The main input into Phylotyper is assembled (but not necessarily closed) genome sequences.  
-Putative loci needed for the selected subtype scheme are identified in the input genomes using BLAST \citep{Camacho2009}.
-The discovered loci are then sent to the Phylotyper subtype prediction module.
-It is possible to bypass the loci search step and provide genes directly to Phylotyper using the subtype run mode.
+Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package (listed in Supplementary Table~1).
+However, included in the Phylotyper software, is the capability to add new subtype schemes. 
+Creating a new subtype scheme will save the required reference files in a data directory, allowing newly added schemes to be easily re-run from Phylotyper.
+Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.
 
-The Phylotyper approach can be used to predict any biological property that can be inferred from the phylogenetic distribution of a nominal set of genomic loci.
-Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) Shiga-toxin 1 and 2 subtypes, Serotype O and H-types and Intimin subtypes are available in the Phylotyper package.
-Details about the current schemes available in the Phylotyper package is provided in Table~1\vphantom{\ref{tab:01}}.
-However, included in the Phylotyper software, is the capability to add new subtype schemes. 
-Creating a new subtype scheme will save the required reference files in a data directory, allowing the newly added schemes to be easily run from Phylotyper. 
-We also encourage users to contribute their subtype schemes to the main software repository (please send us subtype schemes you have developed and wish to share).
+\section{Results}
 
-Checks are built-in new subtype scheme pipeline to ensure the main assumptions of the Phylotyper approach are not violated.  
-A scan of the reference phylogenetic tree is conducted to search for tightly clustered subclades in the tree that have divergent subtypes.
-We first estimate the distribution of the inter-patristic distances of genes with the same subtype.  
-The distribution is used to dynamically select a distance threshold, which we employ to flag subclades wherein the inter-patristic distance is less than  the threshold but contains multiple distinct subtypes.
-We also conduct a leave-one-out cross validation for each gene in the new subtype scheme to assess the scheme's predictive capability.
-Users are alerted if their subtype schemes that have a F$_\text{1}$-score below 0.9 from the cross-validation analysis.
-
-Phylotyper also offers flexibility in the parameterization of evolution model used in the ASR step.
-A component of the underlying ASR framework, is an $\text{M}_{\text{k}}$ or Markov model of subtype evolution, for which an empirical transition rate matrix is estimated from the data.
-The transition matrix is used to calculate the expected number of subtype state changes given a distance in the phylogenetic tree.  
-Different model parameterizations can be defined for the transition rate matrix. 
-The simplest parameterization available in Phylotyper is the equal rates model; all subtypes have the same forward and reverse rate.  
-The most complex parameterization available in Phylotyper is the symmetric model, wherein each forward and reverse rate for a given pair of subtypes are assigned a separate parameter \citep{Revell2011}.
-Frequently, the number of subtypes makes the symmetric model too computationally prohibitive.
-To offer more flexible models in these situations with reduced numbers of free parameters, two custom parameterization approaches were developed.
-The custom approaches both use a binning strategy that attempts to identify sets of subtypes that would have similar rates and assign them a single parameter as a set.
-These approaches are described in detail in the Supplementary Information XXX.
-Each of these model parameterizations; equal, symmetric and the two custom models are tested and evaluated in new subtype pipeline.
-The parameterization that has highest accuracy (based on a LOOCV analysis) is selected.
-In the case of ties, the model with the fewest parameters is given precedence (the symmetric model is not tested when the number of subtypes is over 10)
-
-In testing the predictive capability of Phylotyper, we found that in some cases, jointly using multiple loci increased the subtype prediction accuracy.  
-And so, in Phylotyper, multiple loci can be used in creating subtype schemes.
-The loci will be independently identified in the genome using BLAST and then independently aligned.  
-The individual loci alignments are then concatenated to form a single superalignment that is used to build the predictive phylogenetic tree.  
-The Stx1 and Stx2 schemes are examples of a multi-loci subtype schemes.  
-The A and B subunit genes that make up the holo-toxin in Shiga-toxin are the inputs in these subtype schemes.
-
-\begin{table}[!t]
-\processtable{Subtype schemes available in Phyloyper\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule Name &
-Description & Species & Loci\\\midrule
-stx1 & Shiga-toxin 1 subtype & {\it E. coli} & 2 \\
-stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
-eae & Intimin subtype & {\it E. coli} & 1\\
-flic & H-serotype based on flagellin gene & {\it E. coli} & 1\\
-wz & O-serotype based on wzx and wzy genes & {\it E. coli} & 2\\\botrule
-\end{tabular}}{}
-\end{table}
-
-\section{Comparison to a Sequence Similarity Approach}
-
-To assess the performance of the phylogenetic-based method used in Phylotyper and compare it to a direct sequence similarity-based approach, we ran a leave-one-out cross validation analysis. 
-Each reference gene with experimentally validated subtype was removed from the training dataset and used as a test input into the Phylotyper program. 
-Similarily, using a mock sequence-similarity based tool we developed for assessment purposes, each gene was witheld from the BLAST reference database and used as input (Source code is also available in the phyloyper git repository).
-The sequence-similarity tool assigns putative subtypes using a BLAST sequence search.  
-The predicted subtype annotation for the test input is assigned from top BLAST hit, provided the top hit passes a percent identity, e-value and alignment coverage minimum thresholds.  
-This approach of transfering annotations from genes with highest sequence similarity is the core of most current subtying methods.
-
-\begin{table}[!t]
-\processtable{Results of a Leave-one-out cross validation analysis of Phylotyper and a sequence-similarity approach to predicting subtypes for four different schemes \label{Tab:01}} {
-\begin{tabular}{@{\extracolsep{2pt}}lllllll@{}}
-\toprule 
-\multirow{2}{*}{Scheme} & \multicolumn{3}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
-\cline{2-4}\cline{5-7}
-& Precision & Recall & F$_{1}$ Score & Precision & Recall & F$_{1}$ Score \\
-\midrule
-{\it E. coli} Stx1 & row1 & row1 & row1 & row1 & row1 & row1\\
-{\it E. coli} Stx2 & row2 & row2 & row2 & row2 & row2 & row2\\
-{\it E. coli} Intimin & row3 & row3 & row3 & row3 & row3 & row3\\
-{\it E. coli} H-serotype & row4 & row4 & row4 & row4 & row4 & row4\\
-{\it E. coli} O-serotype & row4 & row4 & row4 & row4 & row4 & row4\\\botrule
-\end{tabular}}{This is a footnote}
-\end{table}
+To assess the performance of the phylogenetic-based method used in Phylotyper and compare it to a sole sequence similarity-based approach, we ran a leave-one-out cross validation analysis.
+For this assessment, we developed a mock sequence-similarity based tool that assigns putative subtypes using BLAST.
+The leave-one-out analysis examined the four subtype schemes available in Phylotyper.
+The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than in a top-BLAST hit approach (see Supplementary Table S2 for results).
+The methods have similar recall rates.
 
@@
 From assembled WGS data, Phylotyper can assign unclassifed strains a molecular subtype.
-Currently the Phylotyper software offers several subtyping schemes for \textit{E. coli}.
-It can be applied to most molecular subtype schemes, however.
-To add a new scheme, it requires a reference set of loci sequences that describe the subtype scheme profile.
-The Phylotyper suite includes functionality to build and run new schemes by users.
-Performance testing showed that the Phylotyper method is highly accurate.
+Currently the Phylotyper software offers subtyping schemes for \textit{E. coli}.
+It can, however, be applied to most molecular subtype schemes and Phylotyper includes functionality to build new schemes.
 
+Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
 Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
-The model can accomodate subtypes that evolve at different rates; a feature that would be difficult in a sequence similarity top-hit approach. 
-By estimating the phylogenetic distribution of each subtype in a particular scheme, Phylotyper more easily identify a novel subtype or accurately classify a novel sequence allele.
-The performance of sequence similarity approaches is more dependent on the reference database containing the exact allele.  
-They do not provide straight-forward mechanism to interpret novel alleles that are not in the database.
-The phylogenetic framework and empirical evolution model provides a statistical confidence value for interpreting novel alleles.
+By estimating the phylogenetic distribution of each subtype, Phylotyper can more easily identify a novel subtype or accurately classify a novel sequence allele.
+The phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting novel alleles.
+In comparison, the performance of a sequence similarity approach is highly dependent on the reference database containing the exact allele.
+In a direct sequence search approach, there is no inherient mechanism to interpret novel alleles.
 
-To exand the Phylotyper program suite, we are working on developing functionality that will automatically identify loci in a whole genome sequence that is highly evolutionarily correlated with a target subtype.
-This function will help streamline the process of developing new subtype schemes in Phylotyper.\vspace*{-10pt}
-
-\section*{Acknowledgements}
-
-text text text\vspace*{-12pt}
+Subtypes are mainly used as a proxy for evolutionarily-related bacterial strain groups or to infer phenotypes.
+Recent analysis of serotype data revealed several inconsistencies between molecular subtype and genomic data \citep{DebRoy2016}.
+Subtypes predicted using a phylogenetic framework is more consistent with these main uses of subtype information, so 
+Phylotyper is uniquely capable to transition historical subtype data to new WGS systems.\vspace*{-10pt}
 
@@
 
-This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}
+This work is funded in part by the Public Health Agency of Canada and a grant from the Genomics Research and Development Initiative\vspace*{-12pt}
 
 \bibliographystyle{natbib}
-% \bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-% \bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-
-%\bibliographystyle{plain}
 

### 2017-02-05 manuscript/phylotyper_suppl_info.tex
Added supplementary info

\documentclass[12pt,letterpaper]{article}
%\usepackage[T1]{fontenc}
%\usepackage[utf8]{inputenc}
\usepackage{ragged2e,booktabs,multirow}
\usepackage{caption,capt-of}
\usepackage{graphicx}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

\RequirePackage[LY1]{fontenc}

\renewcommand{\rmdefault}{ptm}
\renewcommand{\sfdefault}{phb}
\renewcommand{\ttdefault}{pcr}

%\RequirePackage[LY1,mtbold]{mathtime}
\def\helvetica{\fontfamily{phv}\selectfont}
\def\helveticaitalic{\fontfamily{phv}\itshape\selectfont}
\def\helveticabold{\fontfamily{phv}\bfseries\selectfont}
\def\helveticabolditalic{\fontfamily{phv}\bfseries\itshape\selectfont}
\def\helveticacn{\fontfamily{phv}\fontseries{mc}\fontshape{n}\selectfont}
\def\helveticacnitalic{\fontfamily{phv}\fontseries{mc}\fontshape{sl}\selectfont}
\def\helveticacnbold{\fontfamily{phv}\fontseries{bc}\fontshape{n}\selectfont}
\def\helveticacnbolditalic{\fontfamily{phv}\fontseries{bc}\fontshape{sl}\selectfont}

\title{Phylotyper: {\it In silico} predictor of molecular subtypes from gene sequences \\ \large Supplementary Information}
\author{Matthew D. Whiteside, Chad R. Laing and Victor P.J. Gannon}
\date{ }

\begin{document}

\maketitle

\tableofcontents

\section{Supplementary Figures}

\subsection{Supplementary Figure S1}

\begin{figure}[h!]
\centering
%\includegraphics{fig01.eps}
\caption{Stx2 Phylogenetic tree showing the Phylotyper marginal likelihoods as pie charts. This output is provided to the user.}
\end{figure}

\section{Supplementary Tables}

\subsection{Supplementary Table S1}

\begin{minipage}{\linewidth}
\captionof{table}{Subtype Schemes in Phylotyper}
\medskip
\begin{tabular}{@{}llll@{}}\toprule Name &
Description & Species & Loci\\\midrule
stx1 & Shiga-toxin 1 subtype & {\it E. coli} & 2 \\
stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
eae & Intimin subtype & {\it E. coli} & 1\\
flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
wzy & O-serotype based on wzy gene & {\it E. coli} & 1\\\bottomrule
\end{tabular}\par
\bigskip

\end{minipage}

\bigskip

\subsection{Supplementary Table S2}

Table S2 contains performance metrics from a leave-one-out cross-validation analysis comparing Phylotyper and a top-BLAST hit approach.  The analysis examines the four \textit{E. coli} schemes available in Phylotyper. In this multi-class analysis, precision, recall and F$_{1}$ score are calculated for each individual class provided that at least one instance of the class is in the training set.  The individual class positive and negatives are summed to calculate an overall precision, recall and F$_{1}$ score for the scheme.

\begin{minipage}{\linewidth}
\captionof{table}{Leave-One-Out Cross Validation Results}
\medskip
\begin{tabular}{@{\extracolsep{2pt}}lllllll@{}}
\toprule 
\multirow{2}{*}{Scheme} & \multicolumn{3}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
\cline{2-4}\cline{5-7}
& Precision & Recall & F$_{1}$ Score & Precision & Recall & F$_{1}$ Score \\
\midrule
{\it E. coli} Stx1 & 1.00 & 0.94 & 0.97 & 0.94 & 0.94 & 0.94\\
{\it E. coli} Stx2 & 1.00 & 0.99 & 0.99 & 0.93 & 0.93 & 0.93\\
{\it E. coli} Intimin & 1.00 & 0.98 & 0.99 & 0.99 & 0.98 & 0.99\\
{\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 0.96 & 0.85 & 0.90\\
{\it E. coli} O-serotype & row4 & row4 & row4 & row4 & row4 & row4\\\bottomrule
\end{tabular}\par
\bigskip
Formula:
\begin{enumerate}
\item $Precision = TP / (TP + FP)$
\item $Recall = TP / (TP + FN)$
\item $F_{1}~score = 2*Precision*Recall/(Precision + Recall)$
\end{enumerate}
$TP = True~Positive,~~FP = False~Positive,~~FN = False~Negative$

\end{minipage}

\end{document}

0.989795918367347       0.97979797979798

### 2017-02-06 manuscript/phylotyper.tex
GRammer fixes

@@
 Molecular subtyping has been used in public health to infer phenotypes and flag high-risk bacterial strain groups.
-\textit{In silico} tools that predict molecular subtypes from gene sequences are needed to transition historical data and systems to WGS-based protocols.
+\textit{In silico} tools that predict molecular subtypes from gene sequences are needed to transition historical data to WGS-based protocols.
 Phylotyper is a novel solution for \textit{in silico} molecular subtype prediction from gene sequences.
@@
 Whole-genome sequencing (WGS) is transforming the public health field by providing a highly efficient method for surveying bacterial populations.
-The speed, discriminatory power and broad utility of WGS data can improve surveillance and outbreak analysis.
-Adoption of WGS in the public health, however, requires bridging of historical data with the new technologies \citep{Jenkins2015}.
-One of the workhorse methods in public health is molecular subtyping (such as stereotyping).
+The speed, discriminatory power and broad utility of WGS can improve surveillance and outbreak analysis.
+Adoption of WGS in public health, however, requires transitioning of historical data with the new methods \citep{Jenkins2015}.
+One of the workhorse methods in public health is molecular subtyping (such as serotyping).
 As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens \citep{Jenkins2015}.
@@
 
-The core of Phylotyper, is an ancestral state reconstruction (ASR) method that has been adapted for hidden state prediction.
+The core of Phylotyper is an ancestral state reconstruction (ASR) method that has been adapted for hidden state prediction.
 In phylogenetic analysis, ancestral state reconstruction involves the prediction of traits of ancestors from existent descendants.
@@
 In the context of Phylotyper, the marginal likelihood provides a confidence value associated with a predicted subtype.
-The evolutionary model used in Phylotyper, is estimated from the data, so the reference set of genes used to build a phylogenetic tree and run the ASR, is key to the accuracy.
+The evolutionary model used in Phylotyper is estimated from the data, so the reference set of genes used to build a phylogenetic tree and run the ASR, is key to the accuracy.
 The other key assumption is that gene phylogeny is correlated with the evolution of subtype state.
@@
 Individual loci alignments are concatenated to form a single superalignment that is used to build the phylogenetic tree.  
-The Stx1 and Stx2 schemes are examples of a multi-loci subtype schemes, with toxin A and B subunit genes are the inputs in these schemes.
+The Stx1 and Stx2 schemes are examples of a multi-loci subtype scheme; the Stx toxin A and B subunit genes are the inputs in these schemes.
 
@@
 
-To assess the performance of the phylogenetic-based method used in Phylotyper and compare it to a sole sequence similarity-based approach, we ran a leave-one-out cross validation analysis.
+To assess the performance of the phylogenetic-based method used in Phylotyper and compare it to a sequence similarity-based approach, we ran a leave-one-out cross validation analysis.
 For this assessment, we developed a mock sequence-similarity based tool that assigns putative subtypes using BLAST.

### 2017-02-06 manuscript/phylotyper.tex
Added supplementary figure

@@
 The leave-one-out analysis examined the four subtype schemes available in Phylotyper.
-The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than in a top-BLAST hit approach (see Supplementary Table S2 for results).
-The methods have similar recall rates.
+The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than in a top-BLAST hit approach (see Supplementary Table S2 for performance metrics).
 

### 2017-02-13 manuscript/phylotyper.tex
Small edits / comments. Looks great!

@@
 The other key assumption is that gene phylogeny is correlated with the evolution of subtype state.
-Because of these assumptions are central to the method, all new subtype schemes are evaluated for their predictive performance.
+Because these assumptions are central to the method, all new subtype schemes are evaluated for their predictive performance.
 
@@
 \item Identify the subtype with maximum marginal likelihood for the unknown genes and report to user.
-Users are provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g. Figure~1\vphantom{\ref{fig:01}}).
+Users are provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g.
+Figure~1\vphantom{\ref{fig:01}}).
+{Is it possible to print out the type directly, without looking at the tree?}
 \end{enumerate}
@@
 
-Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package (listed in Supplementary Table~1).
+Currently, subtype schemes for \emph{Escherichia coli} (\textit{E.
+coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (listed in Supplementary Table~1).
 However, included in the Phylotyper software, is the capability to add new subtype schemes. 
@@
 The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than in a top-BLAST hit approach (see Supplementary Table S2 for performance metrics).
+{Is there room for some mention of time required / batch capability?}
 

### 2017-02-20 phylotyper/genome/loci.py
Added tests for flic and eae. Fixed missing loci detecition

@@ -217,7 +217,6 @@ class LociSearch(object):
 
             self._sequence_lengths[desc] = int(length)
 
-
         None
 
 
@@ -232,6 +231,9 @@ class LociSearch(object):
 
         Raises Exception if blast search fails
 
+        Returns tuple with:
+            fasta_prefix, number of hits
+
         """
 
         loci = {}
@@ -242,7 +244,6 @@ class LociSearch(object):
 
         if len(output) != self.nloci:
             raise Exception('Missing output file. Need {} files for each loci.'.format(self.nloci))
-
 
         if not fasta_prefix:
             # If an output fasta header ID prefix is not provided
@@ -265,7 +266,7 @@ class LociSearch(object):
             opts = self._blast_options
             opts['query'] = genome
             opts['out'] = tmpfh.name
-
+           
             cmd = "{blastcmd} -evalue {evalue} -outfmt {outfmt} -db {db} -query {query} -out {out}".format(
                 blastcmd=self._blast_exe, **opts)
             if self._blast_specific_options:
@@ -283,7 +284,7 @@ class LociSearch(object):
             
             with open(tmpfh.name, 'r') as blast_handle:
                 blast_records = NCBIXML.parse(blast_handle)
-
+               
                 # Iterate through results
                 for blast_record in blast_records:
                     for alignment in blast_record.alignments:
@@ -358,6 +359,7 @@ class LociSearch(object):
                     mode = 'a'
 
                 locus = 0
+                nhits = 0
                 for ls in loci_sets:
                     genome_copy = 1
                     outp = output[locus]
@@ -366,7 +368,11 @@ class LociSearch(object):
                             outfh.write("\n>{}pt_allele{} {}\n{}".format(fasta_prefix, genome_copy, h, s))
                             genome_copy += 1
 
+                    nhits += 1
                     locus += 1
+
+
+        return (fasta_prefix, nhits)
 
 
     def locad(self, contig, start, stop):

### 2017-02-20 phylotyper/run.py

@@ -7,7 +7,7 @@ Script for running various phylotyper functions
 Examples:
     To run subtyping routine on Stx1 genes:
 
-        $ python main.py subtype ../phylotyper_example.ini ecoli_stx1 test/ecoli_stx1.ffn output/test/
+        $ python -m phylotyper subtype stx1 test/ecoli_stx1.ffn 
 
 """
 
@@ -187,21 +187,38 @@ def subtype_pipeline(options, config):
     logger.info('Settings:\n%s' % (pprint.pformat(options)))
     logger.info('Config:\n%s' % (config.pformat()))
 
-    if genome_search:
-        # Identify loci in input genomes
-        seqtype = 'prot' if options['seq'] == 'aa' else 'nucl'
-        detector = LociSearch(config, options['search_database'], None, seqtype, options['nloci'])
-
-        for genome in options['genomes']:
-            detector.search(genome, options['input'], append=True)
 
     # Predict subtypes
     with open(options['result'], 'w') as resfile:
         assignments = csv.DictWriter(resfile, fieldnames=results_header(), delimiter='\t', quoting=csv.QUOTE_MINIMAL)
         assignments.writeheader()
 
+        loci_found = 0
+        if genome_search:
+            # Identify loci in input genomes
+            seqtype = 'prot' if options['seq'] == 'aa' else 'nucl'
+            detector = LociSearch(config, options['search_database'], None, seqtype, options['nloci'])
+
+            for genome in options['genomes']:
+                genome_label, nhits = detector.search(genome, options['input'], append=True)
+
+                if nhits == 0:
+                     assignments.writerow({
+                        'genome': genome_label,
+                        'tree_label': 'not applicable',
+                        'subtype': 'not applicable',
+                        'probability': 'not applicable',
+                        'phylotyper_assignment': 'Subtype loci not found in genome',
+                        'loci': 'not applicable'
+                    })
+                else:
+                    loci_found += nhits
+              
+
         # Check if sequences match known subtyped sequences
-        remaining = identical_sequences(options, assignments)
+        remaining = []
+        if loci_found > 0:
+            remaining = identical_sequences(options, assignments)
 
         # Run phylotyper on remaining untyped input sequences
         for genome in remaining:
@@ -222,16 +239,12 @@ def subtype_pipeline(options, config):
                 if len(remaining[genome]) == 1:
                     tree_label = genome
 
-                # Only one gene for one genome, don't need to split up input
-                if options['ngenomes'] == 1 and len(remaining[genome]) == 1: 
[…]

### 2017-02-20 phylotyper/test/test_subtype.py

@@ -25,7 +25,8 @@ class SubtypeTests(unittest.TestCase):
 
     def tearDown(self):
         # Remove previous directories created
-        shutil.rmtree(self.test_dir)
+        #shutil.rmtree(self.test_dir)
+        pass
 
 
     def init(self, scheme):
@@ -62,6 +63,30 @@ class SubtypeTests(unittest.TestCase):
         self.subtype_options['ngenomes'] = 4
 
 
+    def thegenome(self):
+        test_genomes = ['test_stx2_genome2.fasta']
+        self.subtype_options['genomes'] = [os.path.join(self.data_dir, f) for f in test_genomes]
+        self.subtype_options['ngenomes'] = 1
+
+
+    def theothergenome(self):
+        test_genomes = ['test_eae_genome1.fasta']
+        self.subtype_options['genomes'] = [os.path.join(self.data_dir, f) for f in test_genomes]
+        self.subtype_options['ngenomes'] = 1
+
+
+    def theotherothergenome(self):
+        test_genomes = ['test_eae_genome2.fasta']
+        self.subtype_options['genomes'] = [os.path.join(self.data_dir, f) for f in test_genomes]
+        self.subtype_options['ngenomes'] = 1
+
+
+    def theflicgenome(self):
+        test_genomes = ['test_flic_genome1.fasta']
+        self.subtype_options['genomes'] = [os.path.join(self.data_dir, f) for f in test_genomes]
+        self.subtype_options['ngenomes'] = 1
+
+
     def testStx2Unknowns(self):
 
         self.init('stx2')
@@ -70,7 +95,7 @@ class SubtypeTests(unittest.TestCase):
         subtype_pipeline(self.subtype_options, self.configObj)
 
         # Check predictions
-        with open(os.path.join(self.test_dir, 'subtype_predictions.csv'), 'rb') as csvfile:
+        with open(os.path.join(self.test_dir, 'subtype_predictions.tsv'), 'rb') as csvfile:
             csvreader = csv.reader(filter(lambda row: row[0]!='#', csvfile), delimiter='\t')
             csvreader.next() # Header
             row1 = csvreader.next()
@@ -82,6 +107,143 @@ class SubtypeTests(unittest.TestCase):
         self.assertTrue(all([row1[4] == 'non-significant/undetermined', row2[4] == 'c', in_range1, above_value2]))
 
 
+    def testStx2(self):
+
+        self.init('stx2')
+        self.thegenome()
+
[…]

### 2017-02-21 manuscript/phylotyper.tex
Add suggestions including runtimes for each subtype scheme

@@
 \item Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes \citep{Revell2011}.
-\item Identify the subtype with maximum marginal likelihood for the unknown genes and report to user.
-Users are provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g.
-Figure~1\vphantom{\ref{fig:01}}).
-{Is it possible to print out the type directly, without looking at the tree?}
+\item Identify the subtype with maximum marginal likelihood for the unknown genes and report to user in text output file.
+Users are also provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g. Figure~1\vphantom{\ref{fig:01}}).
 \end{enumerate}
@@
 The leave-one-out analysis examined the four subtype schemes available in Phylotyper.
-The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than in a top-BLAST hit approach (see Supplementary Table S2 for performance metrics).
-{Is there room for some mention of time required / batch capability?}
+The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than in a top-BLAST hit approach (see Supplementary Table S2 for performance metrics and run-times).
 

### 2017-02-21 manuscript/phylotyper_suppl_info.tex
Updated subtype table
Added supplementary figure
Add suggestions including runtimes for each subtype scheme

@@
-\documentclass[12pt,letterpaper]{article}
+\documentclass[10pt,letterpaper]{article}
 %\usepackage[T1]{fontenc}
@@
 \usepackage{graphicx}
+\usepackage{array}
 \renewcommand{\thetable}{S\arabic{table}}
 \renewcommand{\thefigure}{S\arabic{figure}}
+
+\newcolumntype{M}[1]{>{\begin{varwidth}[t]{#1}}l<{\end{varwidth}}}
 
@@
 
-\maketitle
+%\maketitle
 
 \tableofcontents
+
+\clearpage
 
@@
 \centering
-%\includegraphics{fig01.eps}
+\includegraphics[scale=0.5]{sfig01.png}
 \caption{Stx2 Phylogenetic tree showing the Phylotyper marginal likelihoods as pie charts. This output is provided to the user.}
 \end{figure}
+
+\clearpage
 
@@
 flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
-wzy & O-serotype based on wzy gene & {\it E. coli} & 1\\\bottomrule
+wz & O-serotype based on the wzy and wzx genes & {\it E. coli} & 2\\\bottomrule
 \end{tabular}\par
@@
 \begin{minipage}{\linewidth}
+\setlength{\tabcolsep}{4pt}
 \captionof{table}{Leave-One-Out Cross Validation Results}
 \medskip
-\begin{tabular}{@{\extracolsep{2pt}}lllllll@{}}
+\begin{tabular}{@{\extracolsep{4pt}}llll>{\centering}m{1.6cm}lll@{}}
 \toprule 
-\multirow{2}{*}{Scheme} & \multicolumn{3}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
-\cline{2-4}\cline{5-7}
-& Precision & Recall & F$_{1}$ Score & Precision & Recall & F$_{1}$ Score \\
+\multirow{2}{*}{Scheme} & \multicolumn{4}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
+\cline{2-5}\cline{6-8}
+& Precision & Recall & F$_{1}$ Score & Run-time (s) $^{1}$ & Precision & Recall & F$_{1}$ Score \\
 \midrule
-{\it E. coli} Stx1 & 1.00 & 0.94 & 0.97 & 0.94 & 0.94 & 0.94\\
-{\it E. coli} Stx2 & 1.00 & 0.99 & 0.99 & 0.93 & 0.93 & 0.93\\
-{\it E. coli} Intimin & 1.00 & 0.98 & 0.99 & 0.99 & 0.98 & 0.99\\
-{\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 0.96 & 0.85 & 0.90\\
-{\it E. coli} O-serotype & row4 & row4 & row4 & row4 & row4 & row4\\\bottomrule
+{\it E. coli} Stx1 & 1.00 & 0.94 & 0.97 & 6 & 0.94 & 0.94 & 0.94\\
+{\it E. coli} Stx2 & 1.00 & 0.99 & 0.99 & 32 & 0.93 & 0.93 & 0.93\\
+{\it E. coli} Intimin & 1.00 & 0.98 & 0.99 & 17 & 0.99 & 0.98 & 0.99\\
+{\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 16 & 0.96 & 0.85 & 0.90\\
+{\it E. coli} O-serotype & 1.00 & 0.61 & 0.75 & 278 & 1.00 & 0.36 & 0.53\\\bottomrule
 \end{tabular}\par
@@
 $TP = True~Positive,~~FP = False~Positive,~~FN = False~Negative$
+\par
+\bigskip
+$^{1}$ Run-time is for an {\it Escherichia coli} genome containing a non-identical subtype gene which triggers the full phylotyper pipeline.
 
@@
 \end{document}
-
-0.989795918367347       0.97979797979798

### 2017-02-22 manuscript/phylotyper.tex
Cut down to two pages

@@
 Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
-The phylogenic framework in Phylotyper improves accuracy, provides useful contextual feedback, and is more capable at identifying novel subtypes over approaches based solely on sequence similarity.\\
+The phylogenic framework in Phylotyper improves accuracy, provides useful contextual feedback, and is more capable of identifying novel subtypes over approaches based solely on sequence similarity.\\
 \textbf{Availability and Implementation:} Phylotyper is a python package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
@@
 
-Whole-genome sequencing (WGS) is transforming the public health field by providing a highly efficient method for surveying bacterial populations.
+Whole-genome sequencing (WGS) is transforming the public health field by providing an efficient method for surveying bacterial populations.
 The speed, discriminatory power and broad utility of WGS can improve surveillance and outbreak analysis.
@@
 One of the workhorse methods in public health is molecular subtyping (such as serotyping).
-As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens \citep{Jenkins2015}.
-A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain and routinely run \citep{Jenkins2015}.
-Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Inouye2014,Joensen2015,Ingle2016,Lindsey2016,CARRILLO2016}.
+As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
+A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain\citep{Jenkins2015}.
+Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
 
@@
 Phylotyper is unique in that it builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences to help inform subtype prediction. 
-Using phylogenetic ancestral reconstruction to assign the likelihood of each subtype to the branch points in the tree, Phylotyper assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
+Using phylogenetic ancestral state reconstruction to assign the likelihood of each subtype to the tree branch points, Phylotyper assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
 
@@
 This function calculates the maximum marginal likelihood for unknown tip nodes in a phylogenetic tree.
-The likelihood reflects the most likely state for the node given the empirically estimated evolution model and phylogeny.
-It captures the uncertainty associated with phylogenetic branch lengths and rates of change between states.
+The likelihood reflects the most likely state for the node given the empirically estimated subtype evolution model and phylogeny.
 In the context of Phylotyper, the marginal likelihood provides a confidence value associated with a predicted subtype.
-The evolutionary model used in Phylotyper is estimated from the data, so the reference set of genes used to build a phylogenetic tree and run the ASR, is key to the accuracy.
-The other key assumption is that gene phylogeny is correlated with the evolution of subtype state.
-Because these assumptions are central to the method, all new subtype schemes are evaluated for their predictive performance.
 
 Phylotyper is developed in python and R. 
-Outlined below are the steps and tools used in the Phylotyper pipeline:
-
-\begin{enumerate}
-\item Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}.
-\item Align input genes against a pre-aligned set of reference genes using the tool MAFFT's \texttt{--add} feature \citep{Katoh2013,CapellaGutierrez2009}.
-\item If multiple loci are involved, concatenate individual alignments into superalignment.
-\item Generate maximum likelihood phylogenetic tree of aligned genes with FastTree \citep{Price2010}.
-\item Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes \citep{Revell2011}.
-\item Identify the subtype with maximum marginal likelihood for the unknown genes and report to user in text output file.
+The steps in the Phylotyper pipeline are: 
+(1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}.
+(2) Align input genes against a pre-aligned set of reference genes using MAFFT's \texttt{--add} feature \citep{Katoh2013}.
+(3) If multiple loci are involved, concatenate individual alignments into superalignment.
+(4) Generate maximum likelihood phylogenetic tree of aligned genes with FastTree \citep{Price2010}.
+(5) Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes \citep{Revell2011}.
+(6) Identify the subtype with maximum marginal likelihood for the unknown genes and report to user in text output file.
 Users are also provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g. Figure~1\vphantom{\ref{fig:01}}).
-\end{enumerate}
 
@@
 The subtype marginal likelihood is displayed at each node as a pie chart.
-The full Stx2 subtype tree is displayed in Supplementary Figure S1.}\label{fig:01}
+The full Stx2 tree is displayed in Supplementary Figure S1.}\label{fig:01}
 \end{figure}
 
-Phylotyper was designed to be incorporated into a WGS public health workflow.  
+Phylotyper was designed to be incorporated into a WGS workflow.  
 The main input into Phylotyper is assembled genome sequences.  
@@
 Individual loci alignments are concatenated to form a single superalignment that is used to build the phylogenetic tree.
-The Stx1 and Stx2 schemes are examples of a multi-loci subtype scheme; the Stx toxin A and B subunit genes are the inputs in these schemes.
 
-Currently, subtype schemes for \emph{Escherichia coli} (\textit{E.
-coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (listed in Supplementary Table~1).
-However, included in the Phylotyper software, is the capability to add new subtype schemes. 
-Creating a new subtype scheme will save the required reference files in a data directory, allowing newly added schemes to be easily re-run from Phylotyper.
+Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (Supplementary Table~1).
+However, the Phylotyper software aslo has the capability to add new subtype schemes. 
+Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
 Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.
@@
 
-To assess the performance of the phylogenetic-based method used in Phylotyper and compare it to a sequence similarity-based approach, we ran a leave-one-out cross validation analysis.
+To compare the performance of Phylotyper to a sequence similarity-based approach, we ran a leave-one-out cross validation analysis.
 For this assessment, we developed a mock sequence-similarity based tool that assigns putative subtypes using BLAST.
 The leave-one-out analysis examined the four subtype schemes available in Phylotyper.
-The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than in a top-BLAST hit approach (see Supplementary Table S2 for performance metrics and run-times).
+The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than a top-BLAST hit approach (Supplementary Table S2).
 
@@
 The phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting novel alleles.
-In comparison, the performance of a sequence similarity approach is highly dependent on the reference database containing the exact allele.
-In a direct sequence search approach, there is no inherient mechanism to interpret novel alleles.
+In comparison, the performance of a sequence similarity approach is highly dependent on the reference database.
+This type of approach, there is no inherient mechanism to identify novel alleles.
 
@@
 Recent analysis of serotype data revealed several inconsistencies between molecular subtype and genomic data \citep{DebRoy2016}.
-Subtypes predicted using a phylogenetic framework is more consistent with these main uses of subtype information, so 
+Subtypes predicted using a phylogenetic framework is more consistent with the main uses of subtype information, so 
 Phylotyper is uniquely capable to transition historical subtype data to new WGS systems.\vspace*{-10pt}

### 2017-02-22 phylotyper/test/test_new.py
Changed Wz test genome (previously contained 4 alleles).

@@ -335,7 +335,7 @@ class SubtypeIndexTests(unittest.TestCase):
             touch(os.path.join(self.root_dir, f))
 
         # Blast files
-        touch(os.path.join(self.root_dir, pre_options['search_database']+'.psq'))
+        touch(os.path.join(self.root_dir, pre_options['search_database']+'.nsq'))
 
         post = SubtypeConfig(self.yamlfile)
         post_options = post.get_subtype_config(scheme)

### 2017-02-22 phylotyper/test/test_subtype.py

@@ -69,6 +69,12 @@ class SubtypeTests(unittest.TestCase):
         self.subtype_options['ngenomes'] = 1
 
 
+    def thatgenome(self):
+        test_genomes = ['ecoli_mgy_modified.fasta']
+        self.subtype_options['genomes'] = [os.path.join(self.data_dir, f) for f in test_genomes]
+        self.subtype_options['ngenomes'] = 1
+
+
     def theothergenome(self):
         test_genomes = ['test_eae_genome1.fasta']
         self.subtype_options['genomes'] = [os.path.join(self.data_dir, f) for f in test_genomes]
@@ -186,8 +192,6 @@ class SubtypeTests(unittest.TestCase):
             csvreader.next() # Header
             row = csvreader.next()
 
-            print row
-
         above_value = True if float(row[3]) > .95 else False
         self.assertTrue(all([row[2] == 'epsilon-1', above_value]))
 
@@ -226,7 +230,7 @@ class SubtypeTests(unittest.TestCase):
         self.assertTrue(all([row[2] == 'h12', above_value]))
 
 
-    def testWz(self):
+    def testWzMulti(self):
 
         self.init('wz')
         self.thegenome()
@@ -237,11 +241,27 @@ class SubtypeTests(unittest.TestCase):
         with open(os.path.join(self.test_dir, 'subtype_predictions.tsv'), 'rb') as csvfile:
             csvreader = csv.reader(filter(lambda row: row[0]!='#', csvfile), delimiter='\t')
             csvreader.next() # Header
+            
+            tests = [float(row[3]) > .8 and row[2] == 'o22' for row in csvreader.next()]
+            self.assertTrue(all(tests))
+
+
+    def testWz(self):
+
+        self.init('wz')
+        self.thatgenome()
+
+        subtype_pipeline(self.subtype_options, self.configObj)
+
+        # Check predictions
+        with open(os.path.join(self.test_dir, 'subtype_predictions.tsv'), 'rb') as csvfile:
+            csvreader = csv.reader(filter(lambda row: row[0]!='#', csvfile), delimiter='\t')
+            csvreader.next() # Header
+
             row = csvreader.next()
-
-        above_value = True if float(row[3]) > .85 else False
-        self.assertTrue(all([row[2] == 'a', above_value]))
-
+            
+            above_value = True if float(row[3]) > .95 else False
+            self.assertTrue(all([row[2] == 'o16', above_value]))
 
 
     def testStx2Knowns(self):

### 2017-02-23 phylotyper/test/test_subtype.py
Removed testWzMulti -- too slow

@@ -230,22 +230,6 @@ class SubtypeTests(unittest.TestCase):
         self.assertTrue(all([row[2] == 'h12', above_value]))
 
 
-    def testWzMulti(self):
-
-        self.init('wz')
-        self.thegenome()
-
-        subtype_pipeline(self.subtype_options, self.configObj)
-
-        # Check predictions
-        with open(os.path.join(self.test_dir, 'subtype_predictions.tsv'), 'rb') as csvfile:
-            csvreader = csv.reader(filter(lambda row: row[0]!='#', csvfile), delimiter='\t')
-            csvreader.next() # Header
-            
-            tests = [float(row[3]) > .8 and row[2] == 'o22' for row in csvreader.next()]
-            self.assertTrue(all(tests))
-
-
     def testWz(self):
 
         self.init('wz')

### 2017-04-13 compile_readme.sh
Fixing multi-allele bug

@@ -1,2 +1,2 @@
-pandoc README.md -t json | python pandocfilter-include.py | pandoc -f json -o README.md
+pandoc README.md -t json | python2.7 pandocfilter-include.py | pandoc -f json -o README.md
 

### 2017-04-13 phylotyper/R/performance_testing_roc.R

@@ -1,7 +1,7 @@
 #!/usr/bin/env Rscript
 
 ##########################################################
-## File: performance_tests_for_false_positives.R
+## File: performance_tests_roc.R
 ##
 ##  Remove entire subtype from training data, count number of 
 ##  test set that are incorrectly classified as other subtype
@@ -62,8 +62,31 @@ rs = loadSubtype(treefile,subtypefile)
 tree = rs$tree; subtypes = rs$subtypes
 
 # Run cross-validation
-pp = kfcv(tree, subtypes, scheme=5, model='ER')
+res = kfcv(tree, subtypes, scheme=5, model=modelname)
+
+# Compute FPR / TPR vs cutoff
+pred = prediction(res$all$prediction, res$all$label)
+fpr = performance(pred, 'fpr')
+tpr = performance(pred, 'tpr')
+
+plot(fpr, avg='vertical', lwd=1, col='red', spread.estimate="stderror", 
+	ylab='Average rate across subtypes', xlab='Phylotyper posterior probability cutoff')
+plot(tpr, avg='vertical', lwd=1, col='blue', spread.estimate="stderror", add=TRUE)
+legend(0.6,0.6,c('FPR','TPR'),col=c('red','blue'),lwd=1)
+
+# Compute PPV / TPR for max scheme
+mx.pred = prediction(res$max$prediction, res$max$label)
+plot(performance(mx.pred, 'ppv'), col='red', ylab='Precision', xlab='Phylotyper posterior probability cutoff')
+plot(performance(mx.pred, 'tpr'), col='red', ylab='Recall', xlab='Phylotyper posterior probability cutoff')
+plot(performance(mx.pred, 'f'), col='red', ylab='F1-Statistic', xlab='Phylotyper posterior probability cutoff')
 	
+# Compute the averate FPR when entire subtype is removed
+fp = lsocv(tree, subtypes, scheme=5, model=modelname, threshold=0.85)
+
+lso.fpr = mean(fp[,1]/fp[,2])
+
+
+
 
 
 

### 2017-04-13 phylotyper/R/phylotyper.R

@@ -157,7 +157,7 @@ phylotyper$loadInstallLibraries <- function(libloc="~/R/", repo="http://cran.sta
 	
 	# Install libraries from CRAN
 	cran.libs = c("devtools", "ape", "phangorn", "RColorBrewer", "ggplot2", "optparse", "mclust",
-		"robustbase", "fitdistrplus", "igraph")
+		"robustbase", "fitdistrplus", "igraph", "ROCR")
 	for(x in cran.libs) {
 		if (!require(x,character.only = TRUE)) {
 	  		install.packages(x,dep=TRUE)
@@ -231,6 +231,8 @@ phylotyper$loadSubtype <- function(treefile, stfile=NULL, do.root=TRUE, resolve.
 	return(res)
 }
 
+phylotyper$getPalette <- colorRampPalette(brewer.pal(8, "Set1"))
+
 phylotyper$mypalette <- function(subtypes) {
 	# Generate a set of colors representing subtypes
 	#
@@ -248,16 +250,15 @@ phylotyper$mypalette <- function(subtypes) {
 	
 	states = levels(subtypes)
 	n = length(states)
-	cols = rainbow(n, s=0.6,v=1)
+	cols = phylotyper$getPalette(n)
 
 	if(n > length(cols)) {
 		stop("Number of subtypes exceeds available colors in palette. Please defined your own color palette.")
 	}
 
-	pal = cols[sample(1:n,n)]
-	names(pal) = states
-
-	pal = c(pal,'other'="#D3D3D3")
+	#pal = cols[sample(1:n,n)]
+	names(cols) = states
+	pal = c(cols,'other'="#D3D3D3")
 
 	return(pal)
 }

### 2017-04-13 phylotyper/R/test_functions.R

@@ -180,8 +180,9 @@ kfcv = function(tree, subtypes, scheme=5, model='ER') {
 	results = list()
 	results$predictions = vector("list", length(subtype_classes))
 	results$labels = vector("list", length(subtype_classes))
-	names(results$predictions) <- subtype_classes
-	names(results$labels) <- subtype_classes
+	names(results$predictions) <- c(subtype_classes)
+	names(results$labels) <- c(subtype_classes)
+	pickmax.results = list(predictions=NULL, labels=NULL)
 
 	# Flat prior
 	flat = matrix(1/nstates,1,nstates)
@@ -207,14 +208,19 @@ kfcv = function(tree, subtypes, scheme=5, model='ER') {
 				results$predictions[[c]] <- c(results$predictions[[c]], scores[j,c])
 				results$labels[[c]] <- c(results$labels[[c]], if(c == actual) 1 else 0)
 			}
-		
+
+			mx = which.max(scores[j,])
+			c = colnames(scores)[mx]
+			score = scores[j,mx]
+			
+			pickmax.results$predictions = c(pickmax.results$predictions, score)
+			pickmax.results$labels = c(pickmax.results$labels, if(c == actual) 1 else 0)
 		}
 
 		cat("Completed test iteration",i,"...\n")
-		break;
 	}
 
-	return(results)
+	return(list(all=results,max=pickmax.results))
 }
 
 

### 2017-04-13 phylotyper/performance/cv.py

@@ -152,6 +152,7 @@ def kfcv(config, options, k=10):
 
     # Result objects
     results = { k: [] for k in subtype_classes }
+    maxresults = []
     ng = len(genomes)
     niter = 100
     
@@ -188,25 +189,45 @@ def kfcv(config, options, k=10):
 
             if g in hits:
                 
+                besthit = None
                 for h in hits[g]:
+
+                    if besthit:
+                        if h.bitscore > besthit.bitscore:
+                            besthit = h
+                    else:
+                        besthit = h
+
                     label = 0
                     if truevalue == h.sprop:
                         label = 1
 
                     results[truevalue].append([h.pident, label])
 
+                label = 0
+                if truevalue == besthit.sprop:
+                    label = 1
+                maxresults.append([besthit.pident, label])
+
             else:
                 # Blast returned no results,
                 # so get a FN right out of the gates
                 results[truevalue].append([0, 1])
+                maxresults.append([0, 1])
                 logger.debug('No BLAST result for {}'.format(g))
 
+        i+=1
         logger.debug('Iteration {}'.format(i))
 
 
-    outputfile = os.path.join(options.results, 'performance_results.csv')
+    outputfile = os.path.join(options.results, 'performance_result_tbh_kfcv.csv')
     with open(outputfile, 'w') as csvfile:
         csvwriter =  csv.writer(csvfile, delimiter='\t', quoting=csv.QUOTE_MINIMAL)
+
+        scores = [ r[0] for r in maxresults ]
+        labels = [ r[1] for r in maxresults ]
+        csvwriter.writerow(scores)
+        csvwriter.writerow(labels)
         
         for st in results:
             scores = [ r[0] for r in results[st] ]
@@ -214,9 +235,10 @@ def kfcv(config, options, k=10):
             csvwriter.writerow(st)
             csvwriter.writerow(scores)
             csvwriter.writerow(labels)
+
     
 
-def lco(config, options):
+def lco(config, options, pident_cutoff=90.0):
     """Leave-Class-Out Cross Validation
 
     Args:
@@ -241,10 +263,11 @@ def lco(config, options):
 
     # Leave each subtype class out
     results = []
-    n = 0
-    failed_blasts = []
+
+    print subtype_classes
 
     for c in subtype_classes:
+        logger.debug('Running {}'.format(c))
 
         testg = []
         for g in genomes:
[…]

### 2017-04-13 phylotyper/run.py

@@ -575,7 +575,6 @@ def main():
     genome_parser.add_argument('output', action='store', help='Directory for subtype predictions')
     genome_parser.add_argument('inputs', nargs='+', help='Fasta input for genomes')
     genome_parser.add_argument('--index', help='Specify non-default location of YAML-formatted file index for pre-built subtype schemes')
-    genome_parser.add_argument('--aa', action='store_true', help='Amino acid sequences')
     genome_parser.add_argument('--noplots', action='store_true', help='Do not generate tree image file')
     genome_parser.add_argument('--config', action='store', help='Phylotyper config options file')
     genome_parser.set_defaults(which='genome')

### 2017-04-13 phylotyper/tree/seq.py

@@ -462,6 +462,7 @@ class LociConcat(object):
 
                 # Record allele for this genome
                 sequences[genome].add(i, str(record.seq), str(record.description))
+                print('Added: {} in position {}'.format(str(record.description), i))
 
             for name in uniq:
                 if uniq[name] != i:
@@ -582,6 +583,7 @@ class TypingSequence(object):
         self.positions = {}
         self.nloci = 0
         self.typing_sequences = [[]]
+        self.combinations = []
        
 
     def __iter__(self):
@@ -614,6 +616,9 @@ class TypingSequence(object):
         newallele = Allele(alleleid, sequence, header)
         self.alleles.append(newallele)
 
+        print(self.typing_sequences)
+        print(self.combinations)
+
         if not isinstance(loci, str):
             loci = str(loci)
 
@@ -625,26 +630,59 @@ class TypingSequence(object):
             for ts in self.typing_sequences:
                 ts.append(alleleid)
 
+            if self.nloci < len(self.combinations):
+                self.combinations[self.nloci].append(alleleid)
+            else:
+                self.combinations.append([alleleid])
+
             self.nloci += 1
+            print('extending ts')
 
         else:
             # Add allele to existing loci
-            # Splits typing sequence into multiple
-            # versions; one for each allele
-
             pos = self.positions[loci]
             newseqs = []
 
-            for ts in self.typing_sequences:
-                # make copy and change allele
-                copy = self._copy(ts)
-                copy[pos] = alleleid
-                newseqs.append(copy)
+            # Generate all combinations
+            for c in xrange(len(self.combinations)):
+                print(c)
+                print(newseqs)
+                if c == pos:
+                    if newseqs:
+                        for s in newseqs:
+                            copy = self._copy(s)
+                            print(copy)
+                            print(c)
+                            copy[c] = alleleid
+                            newseqs.append(copy)
+                    else:
+                        newseqs.append([alleleid])
+
+                else:
+                    for a in self.combinations[c]:
+                        if newseqs:
+                            for s in newseqs:
+                                copy = self._copy(s)
+                                copy[c] = a
+                                newseqs.append(copy)
+                        else:
+                            newseqs.append([a])
+                            
+            print('Adding variations:')
+            print(newseqs)
+            # for ts in self.typing_sequences:
+            #     # make copy and change allele
+            #     copy = self._copy(ts)
+                
+            #     newseqs.append(copy)
 
[…]

### 2017-04-13 manuscript/response_to_reviewers.txt
Forgot the new files

response_to_reviewers.txt

----------------------------------------

Here are the comments of the reviewers:

----------------------------------------
Reviewer: 1

Comments to the Author
This is a great little tool that I can see being integrated into numerous pipelines. It moves away from the conventional MLST scheme and is a little more precise than the large scale core genome typing methods being used. I was able to successfully download, install and run the application. I also found the output files very useful and clear, something that doesn't happen enough with many bioinformatics applications.

I have two minor comments that are more "food for thought" than required revisions. I am good with the note as is, everything is clear, well cited and concise. For more information it is possible to visit the GitHub page.

"Food for thought" comments:

Although the GitHub page is well laid out and easy enough to follow, please consider including some sort of example to follow along with. As a newcomer to the application it took me a little while to figure out how to get everything exactly right for an initial run. I think a very small example to follow and run that is located early in the GitHub instructions would really help to just get things going.

*Response*:
We agree with the reviewer's suggestion that a working example provided for new users will enhance the adoption and usability of Phylotyper.  We have added an "Example" section to the README.md file (which is also used as the landing page for Phylotyper's GitHub repository: https://github.com/superphy/insilico-subtyping/) along with example data that demonstrates how to 1) predict subtypes for schemes already packaged in Phylotyper and 2) how to add and run user-defined subtype schemes.

There lies a slight issue in the "user generated sub-type schemes" in that the results that you get with Phylotyper are highly dependant on the quality of the sub-type scheme provided. Since this tool will likely be used in outbreak situations and analyses, do you have any plans to address the quality of the sub-type scheme that was used? Do you plan on having a section in the GitHub page that allows users to send new sub-type schemes for validation and posting?

*Response*:
The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype scheme. We recognized this dependency and incorporated a series of checks into the new subtype build pipeline.  To verify that the phylogeny of the provided sequence is predictive of the subtype, .  To further

Reviewer: 2

Comments to the Author
General comments:
Whiteside et al present here a novel python package for the prediction of bacterial molecular subtypes that are widely used in public health surveillance and outbreak analysis. The release of such a bioinformatic tools is highly valuable and timely with the democratization of WGS in microbiology laboratories worldwide. The article would benefit from some rewriting to provide more details to the reader regarding the use of Phylotyper and how to extend Phylotyper use to other bacteria than Escherichia coli.  Furthermore, a comparison of Phylotyper performance with other existing tools that the authors mention in the introduction would be welcome to assess the strength and weaknesses of this interesting phylogenetic approach.

Major comments:
Introduction:
1.      The authors should elaborate on the "current subtype systems" for readers that would not be familiar with clinical microbiology standards.

Implementation:
2.      Phylotyper currently only has a scheme for Escherichia coli, which limits its use, and the interest resides in the possibility to build schemes for other bacteria. Therefore, it would be useful for the reader to have more information regarding how to build a new scheme in the contribution, and for example which existing resources of typing schemes could be used for main pathogens.

3.      Please provide more details regarding the “assumptions of the Phylotyper approach” and how/why they would be violated when building new subtype schemes.

Results:
4.      Whiteside et al compare here their tool with another BLAST based approach that they have developed for the comparison, and not methods published by other groups. The description of the method used for both the BLAST method and the cross validation is imprecise and would benefit from more being more detailed within the main article rather than in the supplementary material. Moreover, a comparison - at least theoretical - with other existing tools would strengthen the uniqueness of Phylotyper highlighted in the discussion.

Discussion:
5.      There is no data to support the claim that Phylotyper easily identifies novel subtype. Did the author try to remove all alleles of one subtype and compute the predictions to see if/how phylotyper creates a consistent subtype? Also, when a new subtype is identified, does Phylotyper includes it in his typing scheme automatically? What would be the recommended tests/further analysis recommended to researchers using Phylotyper in the case a new subtype is identified?
6.      “Subtypes are mainly used [..] the main uses of subtype information”. Please revise this paragraph, to allow a better logical flow. Also, Debroy et al have shown the discrepancy between between serotype-based typing and the genetic similarity of O antigen gene clusters, suggesting to replace serotyping by molecular typing. They did not particularly emphasize discrepancies between existing molecular typing scheme and whole genome data as suggested by Whiteside et al.

Minor comments:
1.      The abstract mentions a python package only whereas the text in the implementation mentions python and R. Please revise where necessary.
2.      “The main input into Phylotyper is assembled genome sequences”. Please mention the format type (apparently fasta).
3.      “… in addition to Stx include intimin and serotyping” Please revise the sentence.
4.      Replace aslo by also
5.      Revise the sentence “This type of approach, there is…”.

Figure 1.
6.      Please check the label that is cut on the side of the figure. Also, what does the letter at the end of each label correspond to - the expected subtype whereas the predicted subtype is color coded? Please better describe in the legend. Or remove if this is redundant information.
7.      The colors used by Phylotyper could be improved to increase readability of the output for the allele labels. There are several good packages to select colors within R or python.

Sup. Figure 1.
8.      Please plot the legend so that it does not overall the labels
9.      Labels are too close and therefore difficult to read if there were several query allele as an input.

Supplementary table S2:
1.      Four or five E. coli schemes?

Reviewer: 3

Comments to the Author
It's good to see a typing tool that is based on phylogenetic analysis. As public health practitioners still need to compare WGS data to old typing results, this tool will be useful to them.  Moreover, by putting typing into the framework of phylogenetic analysis, this tool can provide an interesting learning/visualization tool for epidemiologists and other public health practitioners to bridge the gap between subtypes (categorical data) and a tree (continuous data). The tool appears to be well implemented.
Some Specific Comments that the authors should address:
Major:
1) Please comment on what happens when an incomplete loci is found in the WGS data. Does the software warn the users of an partial alignment due to incomplete gene?
2) Please clarify how novel subtypes/alleles are identified and reported.
3) As there is a trend moving to core-genome or whole-genome MLST typing schemes, please comment if Phylotyper could be scaled up to handle that. How would horizontal gene transfer or recombination affect the Phylotyper results?

*Response*:
The Phylotyper approach is not likely not applicable to core-genome or whole-genome MLST schemes. The value that Phylotyper provides is that it can take as input gene sequences and return predictions for subtypes based on phenotype or subtypes that cannot be directly inferred from the gene sequences alone. For example, PCR or restriction digestion-based subtyping schemes are an abstraction of the underlying gene sequences and in most cases, cannot be directly inferred from an in silico genome sequence. We have demonstrated the ability of Phylotyper to accurately predict a PCR-based subtype (Stx1 and Stx2) as well as a phenotype-based scheme (O-type:H-type serotype).

Subtypes that are directly identified from the genome sequence and do not involve any inference do not require Phylotyper's additional prediction step. In MLST, the combination each unique allele sequence defines the type.  Potentially, and this is perhaps what the reviewer is refering to, a specific region of the genome that is smaller than the sequence regions used in an MLST, exists that can be used as a proxy to predict MLST subtype. The benefit in this case, is that the entire genome would not need to be sequenced (or multiple alleles interogated) in order to predict MLST type. There is precedent for this type of approach.  The spa and coa genes in Staphylococcus aureus are used as gene markers to infer phylogenetic and population variation.  The spa and coa types have noted concordance with MLST typing. While this is in interesting potential new application for Phylotyper, it is outside the scope of what we originally designed Phylotyper for.

Minor:
1) Typo in line 49: also not aslo
2) Line 51 mentioned assumptions for building new subtype pipeline.  It'll be good to explain that or if space is an issue, point to the portion of the manual that explains these assumptions
----------------------------------------

### 2017-04-13 manuscript/phylotyper_revised.tex
Forgot the new files

--- 2017-02-22 manuscript/phylotyper.tex
@@
 
-\title[short Title]{Phylotyper: {\it In silico} predictor of molecular subtypes from gene sequences}
+\title[short Title]{Phylotyper: {\it In silico} predictor of subtypes from gene subtypes}
 \author[Whiteside \textit{et~al}.]{Matthew D. Whiteside\,$^{\text{\sfb 1,}*}$, Chad R. Laing\,$^{\text{\sfb 1}}$ and Victor P.J. Gannon\,$^{\text{\sfb 1,}*}$}
@@
 \abstract{\textbf{Summary:} Whole genome sequencing (WGS) is being adopted in public health for improved surveillance and outbreak analysis.
-Molecular subtyping has been used in public health to infer phenotypes and flag high-risk bacterial strain groups.
-\textit{In silico} tools that predict molecular subtypes from gene sequences are needed to transition historical data to WGS-based protocols.
-Phylotyper is a novel solution for \textit{in silico} molecular subtype prediction from gene sequences.
+Subtyping; a  is used in public health to infer phenotypes and distinguish bacterial strain groups.
+\textit{In silico} tools that predict molecular subtypes from sequences data are needed to transition historical data to WGS-based protocols.
+Phylotyper is a novel solution for \textit{in silico} subtype prediction from gene sequences.
 Designed for incorporation into WGS pipelines, it is a general prediction tool that can be applied to most molecular subtype schemes.

### 2017-04-13 phylotyper/R/performance_testing_roc_tbh.R
Added files

@@ -1,0 +1,129 @@
+#!/usr/bin/env Rscript
+
+##########################################################
+## File: performance_tests_roc_tbh.R
+##
+##  Summarize Top-BLAST hit approach performance results
+##
+##  R script uses outputs from:
+##    1. python performance/cv.py kfcv ...
+##    2. python performance/cv.py lco ...
+##  
+##  Needs to be run from phylotyper/R directory
+##
+## author: Matt Whiteside
+## copyright: Copyright Government of Canada 2012-2015. Funded by the Government of Canada Genomics Research and Development Initiative
+## license: ASL
+## version: 2.0
+## maintainer: Matt Whiteside
+## email: mdwhitesi@gmail.com
+##
+##########################################################
+
+## MAIN
+
+# Check work directory
+wd = normalizePath(getwd())
+if(!grepl("phylotyper/R/?$", wd, perl=TRUE)) stop(c("Incorrect work directory: ", wd))
+
+# Load libraries and source
+source('phylotyper.R')
+suppressMessages(phylotyper$loadInstallLibraries())
+
+
+# Process command-line args
+option_list = list(
+	make_option(c("-o", "--out"), type="character", default="../output/", 
+    	help="output directory [default: %default]", metavar="character_string")
+); 
+ 
+argparser = OptionParser(usage="%prog [options] lco_results.tsv kfcv_results.tsv", option_list=option_list);
+arguments = parse_args(argparser, positional_arguments=2);
+
+args = arguments$args
+opts = arguments$options
+lcofile = args[1]
+kfcvfile = args[2]
+output_dir = opts$out
+
+
+# Check arguments
+if( file.access(treefile) == -1) {
+	stop(sprintf("Specified tree file ( %s ) does not exist", treefile))
+}
+
+if( file.access(subtypefile) == -1) {
+	stop(sprintf("Specified subtype file ( %s ) does not exist", treefile))
+}
+
+# Load data
+
+# Read the max classifier results
+fp = read.table(lcofile, sep='\t', as.is=TRUE, row.names=1, col.names=c('subtype','fp','n'), header=FALSE)
+
+# Read % identity vs postives
+con  <- file(kfcvfile, open = "r")
+line <- readLines(con, n = 1, warn = FALSE)
[…]

### 2017-04-13 phylotyper/data/download/download_stx_genomes.py

@@ -1,0 +1,301 @@
+#!/usr/bin/env python
+
+"""Download PMIDs for references related to pubmed journals
+
+
+Example:
+        $ python download_stx_genomes.py input output
+
+"""
+
+from Bio import Entrez
+from Bio.Blast import NCBIWWW, NCBIXML
+import argparse
+import logging
+import re
+import os
+import time
+
+__author__ = "Matthew Whiteside"
+__copyright__ = "Copyright Government of Canada 2012-2015. Funded by the Government of Canada Genomics Research and Development Initiative"
+__license__ = "APL"
+__version__ = "2.0"
+__maintainer__ = "Matthew Whiteside"
+__email__ = "mdwhitesi@gmail.com"
+
+
+logging.basicConfig(
+    #filename='download_.log',
+    level=logging.DEBUG,
+    format='%(asctime)s %(message)s',
+    datefmt='%m/%d/%Y %I:%M:%S %p',
+    filemode='w')
+logger = logging.getLogger(__name__)
+
+Entrez.email = 'mdwhitesi@gmail.com'
+
+
+def retrieve(idlist, output_directory):
+
+    logger.debug("trying to retrieve {}".format(idlist))
+
+    genome = None
+    attempt = 1
+    while attempt <= 3:
+        try:
+            print()
+            fetch = Entrez.efetch(db='nuccore', id=idlist, retmode='text', rettype='fasta')
+
+            header = fetch.readline()
+            match = re.search(r'^>(\S\S\S\S+)', header)
+            if not match:
+                raise Exception('Invalid fasta header: {}'.format(header))
+
+            genome = match.group(1)
+            filename = os.path.join(output_directory, genome.replace('|','_') + '.fasta')
+
+            with open(filename, 'w') as out:
+                out.write(header)
+                for line in fetch:
+                    out.write(line)
+
+            fetch.close()
+
+            attempt = 4
+            
+        except Exception as e:
+            logger.warning("Received error from server {}".format(e))
+            logger.warning("Attempt %i of 3".format(attempt))
+            attempt += 1
+            time.sleep(15)
+
+    logger.debug("{} retrieve returned {}".format(idlist, genome))
+
+    return(genome)
+
+
+
+if __name__ == "__main__":
+    """Run
+
+    """
+
+    # Parse command-line args
[…]

### 2017-05-01 manuscript/response_to_reviewers.txt
Updated response to reviewers

@@
 
-*Response*:
+*Response:
 We agree with the reviewer's suggestion that a working example provided for new users will enhance the adoption and usability of Phylotyper.
@@
 *Response*:
-The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype scheme.
-We recognized this dependency and incorporated a series of checks into the new subtype build pipeline.
-To verify that the phylogeny of the provided sequence is predictive of the subtype, .
-To further
+The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor.
+To mitigate poor prediction from unsuitable or inaccurate inputs, we have included in the build pipeline for the local user-submitted subtype feature a series of checks to automatically test the performance of any submitted subtype scheme.
+These tests 1) verify that the phylogeny of the provided sequences places the same subtypes into phylogenetic groups and 2) that the subtype can be accurately predicted with the given input sequences by running a leave-one-out cross-valiation.
+The phylogenetic test empirically determines the distribution of inter-patristic distance between the sequences in the tree with the same subtypes.
+It uses the distribution to identify subtree groupings that place different subtypes in the same clade that are expected to be the same.
+Users are notified if any unexpected subtype assignments are detected in the input.
+In our experience, this test was effective at identifying a limited number of anomylous subtype assignments, for example, incorrectly annotated sequences in the larger subtype scheme.
+In the cross-validation test, each input reference sequence is withheld from the training set and used to test the predictor.
+Users are notified if the subtype scheme has a F1-score below 0.9 from the cross-validation test.
+We found that this test identified subtypes schemes that have poor predictive power due to the subtype being weakly phylogenetically correlated with the input sequences or subtypes being underrepresented in the training set.
+While these checks identify major breakdowns in the requirements for an accurate subtype predictor, we note there is no check to identify missing data, specifically, missing subtypes or key subtypes gene sequences that define the range of the subtype sequence space.
+For any current general subtype predictor (i.e.
+not predictors that replicate PCR in silico, for instance), performance is dependent on the reference input set.
+In dealing with missing data, we demonstrate that Phylotyper is significantly more robust than a direct sequence similarity-based predictor.
+It is less likely to generate a type-I error by when encountering a subtype not found in the training set.
+The Phylotyper method estimates the uncertainty associated with a subtype prediction.
+When test sequences that are phylogenetically distant from known reference sequences or in a region of the phylogenetic tree that is rapidly evolving, the Phylotyper score will reflect the degree of uncertaintity.
+Most often the sequence for a novel or missing subtype will be relatively phylogenetically distant from the other subtype sequences in the tree and its Phylotyper uncertaintity score will reflect this.
+Only in cases where the novel/missing subbtype is closely phylogenetically related to an existing subtype in the training set, will Phylotyper generate a false positive prediction.
+This property of Phylotyper makes it more robust when working with incomplete training data.
+
+To enhance the users understanding of the checks that are provided in the build feature and the potential blindspots in creating their own subtype predictors with Phylotyper, we have greatly expanded the methods section and the README provided with the software.
+The added content describes the two checks, how they work and what they detect.
+In the README we also offer some guidelines about the inputs needed for an effective subtype predictor.
+We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the trainging set and used as test input.
+This analysis and results have been added to the Results and Discussions sections.
+
+The reviewer suggests having a means for users to submit subtype for evaluation and posting to a shared repository.
+We agree that this would be a helpful system.
+It will allow microbiologist specialists to submit and improve subtype schemes and increase the
+We have added a section to the GitHub page, subtype scheme will be checked by us and added to the repository or as a separate download.
 
@@
 1) Please comment on what happens when an incomplete loci is found in the WGS data. Does the software warn the users of an partial alignment due to incomplete gene?
+
+*Response:
+Complete or nearly complete loci sequences are needed for correctly placing the loci in the phylogenetic tree.
+To prevent incomplete loci being used, we set coverage cutoffs of 90% in the BLAST detection step.
+No alleles with less than this coverage to one of the reference set sequences will pass the detection step.
+Users are notified when no loci I have added a description of the requirement for complete loci sequences and the checks we have built into Phylotyper to the Discussion.
+
 2) Please clarify how novel subtypes/alleles are identified and reported.
@@
 
-*Response*:
+*Response:
 The Phylotyper approach is not likely not applicable to core-genome or whole-genome MLST schemes.
@@
 
-Subtypes that are directly identified from the genome sequence and do not involve any inference do not require Phylotyper's additional prediction step.
+Subtypes that are directly identified from the genome sequence do not require Phylotyper's additional prediction step.
 In MLST, the combination each unique allele sequence defines the type.

### 2017-05-01 manuscript/phylotyper_revised.tex
Updated response to reviewers

@@
 Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (Supplementary Table~1).
-However, the Phylotyper software aslo has the capability to add new subtype schemes. 
+However, the Phylotyper software also has the capability to add new subtype schemes. 
 Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
+Performance of the Phylotyper method is dependent on the quality of the subtype scheme.
+Region phylogeny is predictive of the subtype
+Linkage disequalibrium
+homologous genes
+Correct alignment and construction of phylogeny
 Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.
@@
 Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
-By estimating the phylogenetic distribution of each subtype, Phylotyper can more easily identify a novel subtype or accurately classify a novel sequence allele.
-The phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting novel alleles.
+By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make type 1 error when encountering a novel subtype. Our empircal testing confirmed this behavior. 
+The rate of false positive classifications when we withheld an entire subtype from the training set and used as it the test set was significantly lower than with a straight-forward
+
+ as one of the subtypes in the database. To validate
+
+or accurately classify a novel sequence allele.
+The phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting results.
 In comparison, the performance of a sequence similarity approach is highly dependent on the reference database.

### 2017-05-01 manuscript/phylotyper_suppl_info_revised.tex
Forgot the new files
Updated response to reviewers

--- 2017-02-21 manuscript/phylotyper_suppl_info.tex
@@
-\documentclass[10pt,letterpaper]{article}
+\documentclass[11pt,letterpaper]{article}
 %\usepackage[T1]{fontenc}
 %\usepackage[utf8]{inputenc}
+\usepackage[margin=1in]{geometry}
 \usepackage{ragged2e,booktabs,multirow}
@@
 
+~
+
 \begin{minipage}{\linewidth}
+\centering
 \captionof{table}{Subtype Schemes in Phylotyper}
@@
 \end{tabular}\par
-\bigskip
 
@@
 
-\bigskip
+\clearpage
 
@@
 
+%\captionsetup{justification=RaggedRight}
+
 \begin{minipage}{\linewidth}
+\centering
 \setlength{\tabcolsep}{4pt}
-\captionof{table}{Leave-One-Out Cross Validation Results}
+\captionof{table}[justification=RaggedRight]{Leave-One-Out Cross Validation Results}
 \medskip
@@
 {\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 16 & 0.96 & 0.85 & 0.90\\
-{\it E. coli} O-serotype & 1.00 & 0.61 & 0.75 & 278 & 1.00 & 0.36 & 0.53\\\bottomrule
+{\it E. coli} O-serotype & 1.00 & 0.61 & 0.75 & 67 & 1.00 & 0.36 & 0.53\\\bottomrule
 \end{tabular}\par
 \bigskip
+\raggedright
 Formula:
@@
 \bigskip
-$^{1}$ Run-time is for an {\it Escherichia coli} genome containing a non-identical subtype gene which triggers the full phylotyper pipeline.
+$^{1}$ Run-time is for an {\it Escherichia coli} genome containing a unique gene which triggers the full phylotyper pipeline.
+
+\end{minipage}
+
+\begin{minipage}{\linewidth}
+\centering
+\captionof{table}{V-typer Performance}
+\medskip
+\begin{tabular}{@{}lll@{}}\toprule
+NCBI Accession & Stx Subtype & V-typer Prediciton$^{1}$ \\\midrule
+CP015020.1 & 2d & 2d & N \\
+stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
+eae & Intimin subtype & {\it E. coli} & 1\\
+flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
+wz & O-serotype based on the wzy and wzx genes & {\it E. coli} & 2\\\bottomrule
+\end{tabular}\par
+\bigskip
+\raggedright
+$^{1}$ N=Negative (no results returned)
 

### 2017-05-04 phylotyper/R/performance_testing_roc.R
Fixed library loading order

@@ -74,7 +74,7 @@ plot(fpr, avg='vertical', lwd=1, col='red', spread.estimate="stderror",
 plot(tpr, avg='vertical', lwd=1, col='blue', spread.estimate="stderror", add=TRUE)
 legend(0.6,0.6,c('FPR','TPR'),col=c('red','blue'),lwd=1)
 
-# Compute PPV / TPR for max scheme
+# Plot PPV / TPR
 mx.pred = prediction(res$max$prediction, res$max$label)
 plot(performance(mx.pred, 'ppv'), col='red', ylab='Precision', xlab='Phylotyper posterior probability cutoff')
 plot(performance(mx.pred, 'tpr'), col='red', ylab='Recall', xlab='Phylotyper posterior probability cutoff')

### 2017-05-04 phylotyper/R/phylotyper.R

@@ -231,8 +231,6 @@ phylotyper$loadSubtype <- function(treefile, stfile=NULL, do.root=TRUE, resolve.
 	return(res)
 }
 
-phylotyper$getPalette <- colorRampPalette(brewer.pal(8, "Set1"))
-
 phylotyper$mypalette <- function(subtypes) {
 	# Generate a set of colors representing subtypes
 	#
@@ -250,7 +248,8 @@ phylotyper$mypalette <- function(subtypes) {
 	
 	states = levels(subtypes)
 	n = length(states)
-	cols = phylotyper$getPalette(n)
+	getPallete <- colorRampPalette(brewer.pal(8, "Set1"))
+	cols = getPalette(n)
 
 	if(n > length(cols)) {
 		stop("Number of subtypes exceeds available colors in palette. Please defined your own color palette.")

### 2017-05-04 phylotyper/R/performance_testing_roc.R
Added tryCatch to prevent early exit

@@ -62,28 +62,34 @@ rs = loadSubtype(treefile,subtypefile)
 tree = rs$tree; subtypes = rs$subtypes
 
 # Run cross-validation
-res = kfcv(tree, subtypes, scheme=5, model=modelname)
+res = tryCatch(kfcv(tree, subtypes, scheme=5, model=modelname), error=function(cond) { message(cond); return(NA) })
 
-# Compute FPR / TPR vs cutoff
-pred = prediction(res$all$prediction, res$all$label)
-fpr = performance(pred, 'fpr')
-tpr = performance(pred, 'tpr')
+if(!is.na(res)) {
+	# Compute FPR / TPR vs cutoff
+	pred = prediction(res$all$prediction, res$all$label)
+	fpr = performance(pred, 'fpr')
+	tpr = performance(pred, 'tpr')
 
-plot(fpr, avg='vertical', lwd=1, col='red', spread.estimate="stderror", 
-	ylab='Average rate across subtypes', xlab='Phylotyper posterior probability cutoff')
-plot(tpr, avg='vertical', lwd=1, col='blue', spread.estimate="stderror", add=TRUE)
-legend(0.6,0.6,c('FPR','TPR'),col=c('red','blue'),lwd=1)
+	plot(fpr, avg='vertical', lwd=1, col='red', spread.estimate="stderror", 
+		ylab='Average rate across subtypes', xlab='Phylotyper posterior probability cutoff')
+	plot(tpr, avg='vertical', lwd=1, col='blue', spread.estimate="stderror", add=TRUE)
+	legend(0.6,0.6,c('FPR','TPR'),col=c('red','blue'),lwd=1)
 
-# Plot PPV / TPR
-mx.pred = prediction(res$max$prediction, res$max$label)
-plot(performance(mx.pred, 'ppv'), col='red', ylab='Precision', xlab='Phylotyper posterior probability cutoff')
-plot(performance(mx.pred, 'tpr'), col='red', ylab='Recall', xlab='Phylotyper posterior probability cutoff')
-plot(performance(mx.pred, 'f'), col='red', ylab='F1-Statistic', xlab='Phylotyper posterior probability cutoff')
-	
+	# Plot PPV / TPR
+	mx.pred = prediction(res$max$prediction, res$max$label)
+	plot(performance(mx.pred, 'ppv'), col='red', ylab='Precision', xlab='Phylotyper posterior probability cutoff')
[…]

### 2017-05-10 manuscript/response_to_reviewers.txt
UPDATE: new figure colors, ADD: supplementary table S3, ADD: Results section

@@
 In dealing with missing data, we demonstrate that Phylotyper is significantly more robust than a direct sequence similarity-based predictor.
-It is less likely to generate a type-I error by when encountering a subtype not found in the training set.
+It is less likely to generate a type-I error when encountering a subtype not found in the training set.
 The Phylotyper method estimates the uncertainty associated with a subtype prediction.
@@
 Most often the sequence for a novel or missing subtype will be relatively phylogenetically distant from the other subtype sequences in the tree and its Phylotyper uncertaintity score will reflect this.
-Only in cases where the novel/missing subbtype is closely phylogenetically related to an existing subtype in the training set, will Phylotyper generate a false positive prediction.
+Only in cases where the novel/missing subtype is closely phylogenetically related to an existing subtype in the training set, will Phylotyper generate a false positive prediction.
 This property of Phylotyper makes it more robust when working with incomplete training data.
@@
 2) Please clarify how novel subtypes/alleles are identified and reported.
+
+*Response
+Phylotyper is significantly more robust than a direct sequence similarity-based predictor in handling novel subtypes or alleles.
+We do not mean that Phylotyper performs unsupervised construction of new subtype schemes (although the underlying method could be adapted for this task), rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e.
+novel subtype).
+In the case of a novel subtype, Phylotyper would not assign the allele any subtype.
+Compare this result with a sequence similarity approach, which, in most cases assign the novel subtype the closest subtype based on sequence.
+
+To clarify this distinction, we have added and modified text.
+We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set to simulate performance when encountering a novel subtype.
+This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach.
+
 3) As there is a trend moving to core-genome or whole-genome MLST typing schemes, please comment if Phylotyper could be scaled up to handle that.

### 2017-05-10 manuscript/phylotyper_revised.tex
UPDATE: new figure colors, ADD: supplementary table S3, ADD: Results section

@@
 
-Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (Supplementary Table~1).
+Currently, subtype schemes for \emph{Escherichia coli} (\textit{E.
+coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotype O- and H-types  (Supplementary Table~1).
 However, the Phylotyper software also has the capability to add new subtype schemes. 
@@
 Performance of the Phylotyper method is dependent on the quality of the subtype scheme.
-Region phylogeny is predictive of the subtype
+Region / phylogeny is predictive of the subtype
 Linkage disequalibrium
@@
 
-To compare the performance of Phylotyper to a sequence similarity-based approach, we ran a leave-one-out cross validation analysis.
-For this assessment, we developed a mock sequence-similarity based tool that assigns putative subtypes using BLAST.
-The leave-one-out analysis examined the four subtype schemes available in Phylotyper.
-The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than a top-BLAST hit approach (Supplementary Table S2).
+Phylotyper marks a progression from the sequence-similarity approach widely used in \it{in silico} subtyping strategies.
+It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that not in the training data or reference database.
+To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
+The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a training set which excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
+This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
+The second validation examined how the predictors respond when presented with a gene that has a subtype not found in the reference set.
+In this validation, each subtype was iterated over and all genes that are assigned the subtype were removed from the training set.
+In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
+The correct response for the predictors was to return a negative result, since the subtype does not exist in the training set.
+For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
+Details for the BLAST tool and how the assessment was conducted are available in supplementary methods section.
+The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
+When tasked with assigning a novel gene sequence (i.e. one that is not in the training set) in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST hit approach.
+The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST hit approach(result details are available in Supplementary Table S2).
+In order to achieve this level of precision, the BLAST approach often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper
+Similarily, when entire subtype classes were withheld from the training set, Phylotyper had consistenly lower false positive rates for all subtypes schemes tested; the average precision in this challenging test case was 0.11.
+In the BLAST approach, the average precision was 0.30.
+
+A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set.
+The test Stx genes had sufficent surrounding DNA sequence to support \it{in silico} PCR.
+In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
+Phylotyper can correctly predict the subtype for all these genes.
+Based on this level of recall, it appears the conditions in the Stx subtype setting are challenging for simulated PCR.
 
@@
 
+Early developers of a serotype predictor for E. coli. This tool uses BLAST.
 Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
+The goal of this tool is allele matching, and dependence on reference training set is critical to performance.
+To implement allele matching is relatively straight-forward, although some subtyping tools operate with input data such as short reads that require careful analysis.
+EcOH uses SRST tool to align and match short reads from ... to known annotated sequences. Again the focus is on matching rather than inference.
+
 Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.

### 2017-05-10 manuscript/phylotyper_suppl_info.tex
Updated run-time for O-types (original test genome had 4 copies of O-antigen cluster resulting in slower run-time)
UPDATE: new figure colors, ADD: supplementary table S3, ADD: Results section

@@
 
+\section{Supplementary Methods}
+
+\subsection{Performance Analysis}
+
+\subsubsection{BLAST-based method}
+
+BLAST
+
+\subsubsection{Selecting Percent Identity Cutoff for BLAST Approach}
+
+Following \citep{Jenkins2015}, percent identity in the BLAST alignments is used as a cutoff.
+Subtype classifications are transfered from the top BLAST match provided the alignment percent identity is above the cutoff.
+The cutoff impacts performance in one of two ways: too low and non-equivalent subtype alleles may be matched with the query (e.g.
+type I error), too high and valid subtype allele BLAST hits are filtered out (e.g.
+type II error).
+To select the ideal cutoff that balances these competing sources of error, a cross-validation simulation was run to estimate false discovery and true positive rates across varying percent identity cutoffs.
+A 10-fold cross-validation was performed; 10\% of the training set was randomly selected.
+The remaining 90\% was used to build the reference BLAST database.
+The 10\% test set was then searched against this database using BLAST and positive and negative results were recorded.
+This step was repeated 100 times and the F1-score, which is a weighted average of the precision and recall, was calculated.
+Precision is defined as:
+
+$Precision = TP / (TP + FP)$
+
+and recall as:
+
+$Recall = TP / (TP + FN)$
+
+These values are then combined to produce an F-score as follows:
+
+$2 * ( precision * recall ) / ( precision + recall )$
+
+The F-score; a binary classifer metric, was recorded individually for each subtype and then averaged to create a single performance metric for this multi-class predictor.
+F-score was computed for increasing percent identities, and the percent identity corresponding to the maximum F-score was selected as the cutoff.
+
+\subsubsection{Performance Assessment}
+
+Two separate analyses
+
 \section{Supplementary Figures}

### 2017-05-10 manuscript/phylotyper_suppl_info_revised.tex
UPDATE: new figure colors, ADD: supplementary table S3, ADD: Results section

@@
 \clearpage
+
+\section{Supplementary Methods}
+
+\subsection{Performance Analysis}
+
+\subsubsection{BLAST-based method}
+
+BLAST
+
+\subsubsection{Selecting Percent Identity Cutoff for BLAST Approach}
+
+Following \citep{Jenkins2015}, percent identity in the BLAST alignments is used as a cutoff.
+Subtype classifications are transfered from the top BLAST match provided the alignment percent identity is above the cutoff.
+The cutoff impacts performance in one of two ways: too low and non-equivalent subtype alleles may be matched with the query (e.g.
+type I error), too high and valid subtype allele BLAST hits are filtered out (e.g.
+type II error).
+To select the ideal cutoff that balances these competing sources of error, a cross-validation simulation was run to estimate false discovery and true positive rates across varying percent identity cutoffs.
+A 10-fold cross-validation was performed; 10\% of the training set was randomly selected.
+The remaining 90\% was used to build the reference BLAST database.
+The 10\% test set was then searched against this database using BLAST and positive and negative results were recorded.
+This step was repeated 100 times and the F1-score, which is a weighted average of the precision and recall, was calculated.
+Precision is defined as:
+
+$Precision = TP / (TP + FP)$
+
+and recall as:
+
+$Recall = TP / (TP + FN)$
+
+These values are then combined to produce an F-score as follows:
+
+$2 * ( precision * recall ) / ( precision + recall )$
+
+The F-score; a binary classifer metric, was recorded individually for each subtype and then averaged to create a single performance metric for this multi-class predictor.
+F-score was computed for increasing percent identities, and the percent identity corresponding to the maximum F-score was selected as the cutoff.
+
+\subsubsection{Performance Assessment}
+
+Two separate cross-validation tests were performed to compare the performance of Phylotyper with the BLAST method.
+The comparison to vtyper is a separate analysis, as vtyper does not use a training set and a cross-validation test would not work in this situation.
+A leave-one-out cross-validation analysis was performed where each sequence in the training set is withheld and the remaining sequences are used for form the phylogenetic tree for Phylotyper or are used to build the BLAST database for the BLAST-based approach.
+The predicted result for the test sequence is then compared to the true value.
+
+\subsubsection{V-typer Assessment}
+
+The V-typer tool predicts Stx1 or Stx2 subtypes by simulating PCR (i.e.
+mapping primers).
+Due to the unique method in V-typer,  which does not use a training dataset, the cross-validation assessment was adapted to test V-typer.
+Each experimentally-verified Stx gene sequence that makes up Phylotyper training dataset was searched for against the NCBI nt database to retrieve surrounding sequence that may be involved in the PCR detection by V-typer.
+Specifically, regions that have at least 5 kB up- and downstream from the Stx gene were selected for the assessment.
+V-typer was run on each sequence and the results recorded in supplmentary table XXXX.
 
@@
 \medskip
-\begin{tabular}{@{}lll@{}}\toprule
-NCBI Accession & Stx Subtype & V-typer Prediciton$^{1}$ \\\midrule
-CP015020.1 & 2d & 2d & N \\
-stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
-eae & Intimin subtype & {\it E. coli} & 1\\
-flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
-wz & O-serotype based on the wzy and wzx genes & {\it E. coli} & 2\\\bottomrule
+\begin{tabular}{@{}llll@{}}\toprule
+NCBI Accession & Stx Subtype & V-typer Prediciton$^{1}$ & Phylotyper Prediction\\\midrule
+CP015020 & 2d & N & 2d \\
+CP015229 & 2g & 2g & 2g \\
+CP009106 & 2a & N & 2a \\
+CP007133 & 2a & 2a & 2a \\
+CP006262 & 2a & 2a & 2a \\
+AP010960 & 2a & 2a & 2a \\
+CP013663 & 2a & N & 2a \\
+CP015228 & 2a & N & 2a \\
+CP011331 & 2a & N & 2a \\
+HF572917 & 2a & N & 2a \\
+LN554923 & 2a & N & 2a \\
+LM997071 & 2a & N & 2a \\
+CP018250 & 2c & N & 2c \\
+CP018252 & 2c & N & 2c \\
+CP018247 & 2c & N & 2c \\
+CP018245 & 2c & N & 2c \\
+CP018243 & 2c & N & 2c \\
+CP017446 & 2c & N & 2c \\
+CP017442 & 2c & N & 2c \\
+CP017438 & 2c & N & 2c \\
+CP014314 & 2c & N & 2c \\
+LM997036 & 2a & 2a & 2a \\
+LM996489 & 1c & 1c & 1c \\
+LM997036 & 1c & 1c & 1c \\
+Total & 24 & 7 & 24 \\
+\bottomrule
 \end{tabular}\par

### 2017-05-12 phylotyper/R/correlation.R
ADD: subtype randomizer script ADD: leave-subtype-out cross-validation script for phylotyper & blast

@@ -1,0 +1,179 @@
+#!/usr/bin/env Rscript
+
+##########################################################
+## File: correlation.R
+##
+##  Examine how the leave-one-out F-score changes when
+##  the subtype labels in tree are shuffled.
+##
+##  Needs to be run from phylotyper/R directory
+## 
+##
+## author: Matt Whiteside
+## copyright: Copyright Government of Canada 2012-2015. Funded by the Government of Canada Genomics Research and Development Initiative
+## license: ASL
+## version: 2.0
+## maintainer: Matt Whiteside
+## email: mdwhitesi@gmail.com
+##
+##########################################################
+
+## MAIN
+
+# Check work directory
+wd = normalizePath(getwd())
+if(!grepl("phylotyper/R/?$", wd, perl=TRUE)) stop(c("Incorrect work directory: ", wd))
+
+# Load libraries and source
+source('phylotyper.R')
+suppressMessages(phylotyper$loadInstallLibraries())
+source('test_functions.R')
+source('subtype_properties.R')
+x='dplyr'
+if (!require(x,character.only = TRUE)) {
+	install.packages(x,dep=TRUE)
+	library(x)
+	if(!require(x,character.only = TRUE)) stop(paste("Package not loaded: ",x))
+}
+
+
+# Process command-line args
+option_list = list(
+	make_option(c("-o", "--out"), type="character", default="../output/", 
+    	help="output directory [default: %default]", metavar="character_string")
+	make_option(c("-m", "--model"), type="character", default="ER", 
+    	help="transition rate model [default: %default]", metavar="character_string")
+); 
+ 
+argparser = OptionParser(usage="%prog [options] tree_file subtype_file modelname", option_list=option_list);
+arguments = parse_args(argparser, positional_arguments=3);
+
+args = arguments$args
+opts = arguments$options
+treefile = args[1]
+subtypefile = args[2]
+modelname = opts$model
+output_dir = opts$out
+
+
+# Check arguments
+if( file.access(treefile) == -1) {
+	stop(sprintf("Specified tree file ( %s ) does not exist", treefile))
+}
+if( file.access(subtypefile) == -1) {
+	stop(sprintf("Specified subtype file ( %s ) does not exist", treefile))
+}
[…]

### 2017-05-12 phylotyper/R/performance_testing_roc.R

@@ -57,37 +57,70 @@ if( file.access(subtypefile) == -1) {
 	stop(sprintf("Specified subtype file ( %s ) does not exist", treefile))
 }
 
+if(! modelname %in% c('ER','SYM','GROUP','ITER')) {
+	stop(sprintf("Invalid model ( %s )", modelname))
+}
+
 # Load tree
 rs = loadSubtype(treefile,subtypefile)
 tree = rs$tree; subtypes = rs$subtypes
 
-# Run cross-validation
-res = tryCatch(kfcv(tree, subtypes, scheme=5, model=modelname), error=function(cond) { message(cond); return(NA) })
+model <- NULL
+if(modelname == 'GROUP') {
+	rs = transition.rate.parameters(tree, subtypes)
+	model = rs$model
+} else if(modelname == 'ITER') {
+	rs = transition.rate.parameters2(tree, subtypes)
+	model = rs$model
+} else {
+	model = modelname
+}
 
-if(!is.na(res)) {
+print(model)
+
+# Run cross-validation
+res = tryCatch(kfcv(tree, subtypes, scheme=5, model=model), error=function(cond) { message(cond); return(NA) })
+
+if(length(res) > 1 && !is.na(res)) {
 	# Compute FPR / TPR vs cutoff
 	pred = prediction(res$all$prediction, res$all$label)
 	fpr = performance(pred, 'fpr')
 	tpr = performance(pred, 'tpr')
 
+	png(file=file.path(output_dir, 'roc.png'),
+		width = 720,
+		height = 720
+	)
 	plot(fpr, avg='vertical', lwd=1, col='red', spread.estimate="stderror", 
 		ylab='Average rate across subtypes', xlab='Phylotyper posterior probability cutoff')
 	plot(tpr, avg='vertical', lwd=1, col='blue', spread.estimate="stderror", add=TRUE)
 	legend(0.6,0.6,c('FPR','TPR'),col=c('red','blue'),lwd=1)
+	dev.off()
 
 	# Plot PPV / TPR
+	png(file=file.path(output_dir, 'ppv.png'),
+		width=6,
+		height=12,
+		units='in',
+		res=144
+	)
+	par(mfrow=c(3,1))
+	par(mar=rep(3,4))
 	mx.pred = prediction(res$max$prediction, res$max$label)
[…]

### 2017-05-12 phylotyper/R/performance_testing_roc_tbh.R

@@ -48,65 +48,94 @@ output_dir = opts$out
 
 
 # Check arguments
-if( file.access(treefile) == -1) {
-	stop(sprintf("Specified tree file ( %s ) does not exist", treefile))
+if( file.access(lcofile) == -1) {
+	stop(sprintf("Specified Leave-class-out results file ( %s ) does not exist", lcofile))
 }
 
-if( file.access(subtypefile) == -1) {
-	stop(sprintf("Specified subtype file ( %s ) does not exist", treefile))
+if( kfcvfile != 'NONE') {
+	if( file.access(kfcvfile) == -1) {
+		stop(sprintf("Specified K-folds cross-validation results file ( %s ) does not exist", kfcvfile))
+	}
 }
 
-# Load data
+
+# Read % identity vs postives
+if(kfcvfile != 'NONE') {
+	con  <- file(kfcvfile, open = "r")
+	line <- readLines(con, n = 1, warn = FALSE)
+	mxpredictions <- as.numeric(strsplit(line, "\t")[[1]])
+	line <- readLines(con, n = 1, warn = FALSE)
+	mxlabels <- as.numeric(strsplit(line, "\t")[[1]])
+
+	results <- list(predictions=NULL, labels=NULL)
+
+	while (length(line <- readLines(con, n = 1, warn = FALSE)) > 0) {
+
+		subtype <- line
+		line2 <- readLines(con, n = 1, warn = FALSE)
+		predictions <- (strsplit(line2, "\t"))
+	    predictions <- c(as.numeric(predictions[[1]]), 0) # Add a superfluous TN to prevent NaN in FPR calculations
+	                                                      # i.e. something with 0% percent identity should get a negative assignment
+	    line3 <- readLines(con, n = 1, warn = FALSE)
+		labels <- (strsplit(line3, "\t"))
+	    labels <- c(as.numeric(labels[[1]]),0)
+
+	    results$predictions[[subtype]] = predictions
+	    results$labels[[subtype]] = labels
+
+	} 
+
+	close(con)
+
+	# Compute FPR / TPR vs cutoff
+	pred = prediction(results$predictions, results$labels)
+	fpr = performance(pred, 'fpr')
+	tpr = performance(pred, 'tpr')
+
+	# png(file=file.path(output_dir, 'roc_tbh.png'),
+	# 	width = 720,
+	# 	height = 720
+	# )
+
[…]

### 2017-05-12 phylotyper/R/performance_tests.R

@@ -67,24 +67,27 @@ cat("Computing posterior probabilites for all internal nodes\n")
 
 priorR = phylotyper$makePriors(tree, subtypes)
 priorM = priorR$prior.matrix
-fit <- rerootingMethod(tree, priorM, model='ER', tips=FALSE)
+fit <- rerootingMethod(tree, priorM, model='SYM', tips=FALSE)
 
-# source('phylotyper.R')
-# file = 'posterior_probability_tree'
-# #dim = phylotyper$plotDim(tree)
-# r = 300
-# w = 3.38 * r
-# graphics.off()
-# # png(filename=file.path(output_dir, paste(file, '.png', sep='')),
-# #     width=dim[['x']],height=dim[['y']],res=dim[['res']]
-# # )
-# setEPS()
-# postscript(file.path(output_dir, paste(file, '.eps', sep='')),
-# 	width=3.38,height=4,pointsize=6)
-# # tiff(filename=file.path(output_dir, paste(file, '.tiff', sep='')),
-# # 	width=w, height=w*1.25, res=r, pointsize=6, units='px', compression='zip')
-# phylotyper$plot.anc(tree,fit,subtypes)
-# graphics.off()
+
+
+source('phylotyper.R')
+file = 'posterior_probability_tree_small'
+#file = 'posterior_probability_tree_large'
+dim = phylotyper$plotDim(tree)
+r = 300
+w = 3.38 * r
+graphics.off()
+# png(filename=file.path(output_dir, paste(file, '.png', sep='')),
+#     width=dim[['x']],height=dim[['y']],res=dim[['res']]
+# )
+setEPS()
+postscript(file.path(output_dir, paste(file, '.eps', sep='')),
+	width=3.38,height=4,pointsize=6)
+# tiff(filename=file.path(output_dir, paste(file, '.tiff', sep='')),
+# 	width=w, height=w*1.25, res=r, pointsize=6, units='px', compression='zip')
+phylotyper$plot.anc(tree,fit,subtypes)
+graphics.off()
 
 
 
@@ -99,7 +102,7 @@ for(validation in cvs) {
 	pp = do.call(validation, list(tree=tree, subtypes=subtypes, scheme=est.scheme))
 
 	# Summarize performance
-	results = simulationSummary(subtypes, pp, threshold=.75)
+	results = simulationSummary(subtypes, pp, threshold=.85)
 
 	# Write performance metrics to file
 	file = 'performance_metrics'

### 2017-05-12 phylotyper/R/phylotyper.R

@@ -248,9 +248,8 @@ phylotyper$mypalette <- function(subtypes) {
 	
 	states = levels(subtypes)
 	n = length(states)
-	getPallete <- colorRampPalette(brewer.pal(8, "Set1"))
-	cols = getPalette(n)
-
+	cols <- colorRampPalette(brewer.pal(8, "Set1"))(n)
+	
 	if(n > length(cols)) {
 		stop("Number of subtypes exceeds available colors in palette. Please defined your own color palette.")
 	}
@@ -373,7 +372,7 @@ phylotyper$plot.anc <- function(tree, fit, subtypes) {
 
 	cols = phylotyper$mypalette(subtypes)
 
-	plot(tree,label.offset=0.001,cex=0.8,type='fan',align.tip.label=TRUE,tip.col=cols[subtypes[tree$tip.label]],
+	plot(tree,label.offset=0.001,cex=0.8,type='phylo',align.tip.label=TRUE,tip.col=cols[subtypes[tree$tip.label]],
 		no.margin=TRUE)
 
 	nodelabels(pie=phylotyper$piecolors(fit$marginal.anc),

### 2017-05-12 phylotyper/R/test_functions.R

@@ -132,12 +132,20 @@ lsocv = function(tree, subtypes, scheme=5, model='ER', threshold=.7) {
 		
 		stopifnot(all(tipnames %in% tree$tip.label))
 
-		testfit = phylotyper$runSubtypeProcedure(tree, testprior, scheme, tips=tipnames, model=model, fixedQ=Q)
+		# Hide columns/rows if model is matrix
+		if(!is.character(model)) {
+			testmodel = model[-which(rownames(model) %in% state),]
+			testmodel = testmodel[,-which(colnames(model) %in% state)]
+		} else{
+			testmodel = model
+		}
+
+		testfit = phylotyper$runSubtypeProcedure(tree, testprior, scheme, tips=tipnames, model=testmodel, fixedQ=Q)
 		pp[state,'fp'] = sum(testfit$tip.pp[tipnames, ] >= threshold)
 		pp[state, 'n'] = length(testers)
 
 	}
-
+	
 	return(pp)
 }
 
@@ -163,7 +171,7 @@ kfcv = function(tree, subtypes, scheme=5, model='ER') {
 	# Compute k-fold size
 	n = length(subtypes)
 	iter = 100
-	k = 10
+	k = 5
 	ksize = floor(n/k)
 	cat("Cross-validation test set size: ",ksize,"\n")
 
@@ -261,7 +269,7 @@ updateClassificationSums = function(true_class, predicted_class, count_matrix) {
 }
 
 
-simulationSummary = function(subtypes, pp, threshold=.7) {
+simulationSummary = function(subtypes, pp, threshold=.85) {
 	# Summarize performance of validation
 	# 
 	# 

### 2017-05-12 phylotyper/performance/cv.py

@@ -238,7 +238,7 @@ def kfcv(config, options, k=10):
 
     
 
-def lco(config, options, pident_cutoff=90.0):
+def lco(config, options, pident_cutoff=97.0):
     """Leave-Class-Out Cross Validation
 
     Args:

### 2017-05-12 phylotyper/phylotyper.py

@@ -285,7 +285,7 @@ class Phylotyper(object):
             pp = loocv(tree=tree, subtypes=subtypes, scheme=est.scheme, model=mod)
 
             # Summarize performance
-            results = simulationSummary(subtypes, pp, threshold=0.9)
+            results = simulationSummary(subtypes, pp, threshold=0.85)
 
             # Save a copy of the fit
             fit <- fitMk(tree, subtypes, model=mod)

### 2017-05-12 phylotyper/tree/seq.py

@@ -462,8 +462,8 @@ class LociConcat(object):
 
                 # Record allele for this genome
                 sequences[genome].add(i, str(record.seq), str(record.description))
-                print('Added: {} in position {}'.format(str(record.description), i))
-
+                
+                
             for name in uniq:
                 if uniq[name] != i:
                     if missing == 'raise':
@@ -616,9 +616,6 @@ class TypingSequence(object):
         newallele = Allele(alleleid, sequence, header)
         self.alleles.append(newallele)
 
-        print(self.typing_sequences)
-        print(self.combinations)
-
         if not isinstance(loci, str):
             loci = str(loci)
 
@@ -636,8 +633,7 @@ class TypingSequence(object):
                 self.combinations.append([alleleid])
 
             self.nloci += 1
-            print('extending ts')
-
+            
         else:
             # Add allele to existing loci
             pos = self.positions[loci]
@@ -645,14 +641,11 @@ class TypingSequence(object):
 
             # Generate all combinations
             for c in xrange(len(self.combinations)):
-                print(c)
-                print(newseqs)
+                
                 if c == pos:
                     if newseqs:
                         for s in newseqs:
                             copy = self._copy(s)
-                            print(copy)
-                            print(c)
                             copy[c] = alleleid
                             newseqs.append(copy)
                     else:
@@ -668,21 +661,10 @@ class TypingSequence(object):
                         else:
                             newseqs.append([a])
                             
-            print('Adding variations:')
-            print(newseqs)
-            # for ts in self.typing_sequences:
-            #     # make copy and change allele
-            #     copy = self._copy(ts)
-                
-            #     newseqs.append(copy)
-
             # Save new typing sequences with alternate
             # allele
             self.typing_sequences.extend(newseqs)
-            print(self.typing_sequences)
-
-
-        print(len(self.typing_sequences))
+            
 
     def _copy(self, ts):
         # Make a copy of a typing sequence

### 2017-05-15 manuscript/phylotyper_revised.tex
UPDATE: Discussion section ADD: Randomized subtype test

@@
 The steps in the Phylotyper pipeline are: 
-(1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}.
+(1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}. Hits that do not align over 95\% or have under 90\% sequence identity with reference sequence are discarded.
 (2) Align input genes against a pre-aligned set of reference genes using MAFFT's \texttt{--add} feature \citep{Katoh2013}.
@@
 Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
-Performance of the Phylotyper method is dependent on the quality of the subtype scheme.
-Region / phylogeny is predictive of the subtype
-Linkage disequalibrium
-homologous genes
-Correct alignment and construction of phylogeny
-Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.
+To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
+Phylotyper assumes that the provided training sequence are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
+Where possible, checks are built-in to the \texttt{new} pipeline to ensure the assumptions of the Phylotyper approach are not violated.
+Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline.
+Results are reported to the user.
+The first test checks the the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
+This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
+A second check computes the accuracy measure; F-score, through a leave-one-out cross-validation.
+This validation uses each sequence in the training set as a test input (the test sequence is removed from the tree and then used as input).
+Results are recorded. 
+Metrics like precision, recall and F-score rapidly erode as the phylogenetic grouping is decoupled from the subtypes.
+These checks will verify that the sequences phylogenetic grouping is predictive of the subtype subtype.
+There is no check that can verify the training set covers all subtypes; this is left up to the user.
+However, as we show in the Results section, Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set.
+
+To grow the Phylotyper resource, we encourage users to contact us with their new subtype schemes or updates to schemes (\url{https://github.com/superphy/insilico-subtyping}). 
+We will continue to expand the Phylotyper resource by adding and updating high-quality subtype schemes in the Phylotyper repository.
 
@@
 In order to achieve this level of precision, the BLAST approach often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper
-Similarily, when entire subtype classes were withheld from the training set, Phylotyper had consistenly lower false positive rates for all subtypes schemes tested; the average precision in this challenging test case was 0.11.
-In the BLAST approach, the average precision was 0.30.
-
+Similarily, when entire subtype classes were withheld from the training set, Phylotyper had consistenly lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
 A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set.
@@
 Phylotyper can correctly predict the subtype for all these genes.
-Based on this level of recall, it appears the conditions in the Stx subtype setting are challenging for simulated PCR.
+Based on this level of recall, it appears conditions in the Stx subtype setting are challenging for simulated PCR.
+
+All new or updated subtype schemes added in Phylotyper are subject to a leave-one-out cross-validation test as described earlier.  
+The test is part of the \texttt{add} pipeline and is used to estimate the F-score of the subtype scheme. 
+The F-score reflects that predictive capability of the subtype scheme. 
+If the associated phylogeny for training set gene sequences are not correlated with the subtype distribution, this will be reflected in the F-score. 
+To demonstrate this property, we randomly assigned subtypes for increasing proportions of the genes in the training set and computed the F-score with the leave-one-out validation for each proportion level.
+The F-score and recall rapidly decrease as the proportion of altered subtypes increased (Supplementary Figure S2).
 
@@
 
-Early developers of a serotype predictor for E. coli. This tool uses BLAST.
+One of the strategies for predicting subtypes is to use sequence similarity to annotated loci alleles.
+Query genomes or genes are matched to alleles that are attributed the subtype phenotype or are correlated with the distribution of the subtype.
+For example SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textid{in silico} O-typing, and the flagellin genes for H-typing \textit{E.
+coli} genes.
+O-type and H-type are transfered from the top matches to the queries provided they are above coverage and percent identity thresholds.
+This general strategy of allele matching is also applied in EcOH database, however, the EcOH can directly use unassembled sequence reads as input.
+The EcOH tool aligns reads to alleles linked to \textit{E.
+coli} O-types and H-types, and identifies the top candidates that have an alignment score above pre-defined thresholds.
 Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
-The goal of this tool is allele matching, and dependence on reference training set is critical to performance.
-To implement allele matching is relatively straight-forward, although some subtyping tools operate with input data such as short reads that require careful analysis.
-EcOH uses SRST tool to align and match short reads from ... to known annotated sequences. Again the focus is on matching rather than inference.
-
-Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
-By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make type 1 error when encountering a novel subtype. Our empircal testing confirmed this behavior. 
-The rate of false positive classifications when we withheld an entire subtype from the training set and used as it the test set was significantly lower than with a straight-forward
-
- as one of the subtypes in the database. To validate
-
-or accurately classify a novel sequence allele.
-The phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting results.
-In comparison, the performance of a sequence similarity approach is highly dependent on the reference database.
-This type of approach, there is no inherient mechanism to identify novel alleles.
+With a strict allele matching stategy, the reference or training set make-up can be critical to performance.
+When alleles or even subtypes are missing, the sequence similarity approach more frequently generates type-I errors.
+Alternatively, Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
+By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make type-I error when encountering a novel subtype or allele. 
+The empircal testing we performed demonstrated this behavior; 
+the rate of false positive classifications was significantly lower than in a sequence-similarity approach in validations where we withheld an allele or an entire subtype from the training set and used it as a test input.
+V-typer takes a distinct approach; it directly simulates the \textit{in vitro} wet-lab PCR procedure used to perform Stx subtyping. 
+V-typer's direct replication of a wet-lab method \textit{in silico} means it can only be applied to subtypes schemes that use PCR. 
+Additionally, we found in our evaluation of Stx subtypes that it failed to generate predictions for most test cases.
+From a broader methodology viewpoint, one of the benefits of the phylogenetic framework in Phylotyper, is that it provides a statistical likelihood for interpreting results.
+In comparison, there is no built-in mechanism in the sequence similarity approach to quantify the level of confidence in assigning alleles a subtype.
 
 Subtypes are mainly used as a proxy for evolutionarily-related bacterial strain groups or to infer phenotypes.
-Recent analysis of serotype data revealed several inconsistencies between molecular subtype and genomic data \citep{DebRoy2016}.
-Subtypes predicted using a phylogenetic framework is more consistent with the main uses of subtype information, so 
-Phylotyper is uniquely capable to transition historical subtype data to new WGS systems.\vspace*{-10pt}
+A recent analysis of O-antigen serotypes and their associated O-antigen gene sequences in \textit{E.
+coli} found that the sequence data indicated several changes to the nomenclature for O-groups \citep{DebRoy2016}.
+There are potentially many other subtype schemes that would not be straightforward to adapt to a WGS approach.
+A tool that can evaluate the ability of a genotype to predict a subtype would be better equipped for developing the new subtype schemes or updating current schemes for WGS workflows.
+Phylotyper's \texttt{add} pipeline tests subtype schemes for their predictive accuracy by returning an F-score based on a cross-validation assessment.
+The validation verifies that the phylogeny generated by the training set sequences can be used to predict the subtypes with a high level of accuracy.
+We showed in empirical tests that the farther a subtype is dissociated from the genotype, the lower the F-score computed in the Phylotyper \texttt{add} pipeline.
+In addition to this subtype-level verification, Phylotyper also computes confidence scores for each individual prediction that reflect the rate of subtype change occuring at the input gene's phylogenetic locale.
+If an input sequence falls in region in the phylogenetic tree where the subtype is highly fluid due to evolution or poor subtype-genotype correlation, users would be notified in the condifidence score and in phylogenetic tree visualization output by Phylotyper.
+Phylotyper ability to inform users about the level of agreement between the subtype assignments and genotype makes it uniquely capable to transition historical subtype data to new whole genome sequence-based systems.\vspace*{-10pt}
 

### 2017-05-15 manuscript/phylotyper_suppl_info_revised.tex
UPDATE: Discussion section ADD: Randomized subtype test

@@
 
-\subsection{Performance Analysis}
+\subsection{BLAST-based method}
 
-\subsubsection{BLAST-based method}
-
-BLAST
-
-\subsubsection{Selecting Percent Identity Cutoff for BLAST Approach}
-
-Following \citep{Jenkins2015}, percent identity in the BLAST alignments is used as a cutoff.
-Subtype classifications are transfered from the top BLAST match provided the alignment percent identity is above the cutoff.
+For comparison purposes, we developed a generalized sequence-similarity predictor of subtypes using BLAST.
+Following \citep{Jenkins2015}, a training set of sequences with annotated subtypes is used to build a BLAST database (the training sets are from the the Phylotyper repository).
+To predict a subtype, query sequences are searched against the training BLAST database and the top hit is recorded.
+If the top hit's alignment with the query has at least 60\% coverage with the reference sequence and is above the percent identity cutoff, then the subtype of the top hit is returned as the predicted subtype of the query sequence.
 The cutoff impacts performance in one of two ways: too low and non-equivalent subtype alleles may be matched with the query (e.g.
@@
 The F-score; a binary classifer metric, was recorded individually for each subtype and then averaged to create a single performance metric for this multi-class predictor.
-F-score was computed for increasing percent identities, and the percent identity corresponding to the maximum F-score was selected as the cutoff.
+F-score was computed for increasing percent identities, and the lowest percent identity corresponding to the maximum F-score was selected as the cutoff.
+The percent identity cutoff was selected as 0.97.
 
-\subsubsection{Performance Assessment}
+\subsection{Performance Assessment}
 
@@
 The predicted result for the test sequence is then compared to the true value.
+
+The second validation iterated over each subtype and removed all genes assigned that subtype from the training set.
+The removed genes or test set was then used as input to identify the level of false positive predictions that occur when the tools are presented with novel subtypes not in the training set.
+
+Precision, recall and the F-score as defined earlier for a multi-classifier test, were computed for the validations.
 
@@
 \caption{Stx2 Phylogenetic tree showing the Phylotyper marginal likelihoods as pie charts. This output is provided to the user.}
+\end{figure}
+
+\clearpage
+
+\subsection{Supplementary Figure S2}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[scale=0.5]{sfig02.png}
+\caption{The effect of randomizing a specified proportion of subtype labels in the Eae phylogenetic tree on the empirically estimated performance metrics: F$_1$-score and recall.}
 \end{figure}

### 2017-05-16 manuscript/response_to_reviewers.txt
UPDATE: intro UPDATE: implementation UPDATE: suppl info UPDATE: response to reviewers

@@
 
+Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations were linkage disequlibrium can occur.
+If the sequences used to build the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E.
+coli being used to predict O-group serotypes, then the prediction would directly be linked to underlying genome.
+If however a proxy sequence is used, HGT could impact.
+We have recommended to users building subtype schemes they focus on
+
 Minor:

### 2017-05-16 manuscript/phylotyper_revised.tex
UPDATE: intro UPDATE: implementation UPDATE: suppl info UPDATE: response to reviewers

@@
 Adoption of WGS in public health, however, requires transitioning of historical data with the new methods \citep{Jenkins2015}.
-One of the workhorse methods in public health is molecular subtyping (such as serotyping).
+One of the workhorse methods in public health is subtyping.
+Subtyping methods can broadly categorized as phenotype-based or DNA-based.
+Phenotype-based subtypes are, for example, interrogated by biochemical tests (biotyping), detection of surface antigens (serotyping) or susceptibility to bacteriophage (phagetyping) \citep{Wiedmann2002}.
+Alternatively DNA-based subtyping examines and classifies bacteria based on genetic content.
+DNA-based subtypes use a variety of methods from PCR to Pulse Field Gel Electrophoresis to DNA sequencing to assign groups to bacteria \citep{Wiedmann2002}.
 As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
@@
 Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
+To predict from WGS data, these tools target subtypes (Phenotype-based or DNA-based) that can be linked to a specific region or gene in the genome.
+These tools have not been applied to subtyping methods that already directly use DNA sequence data to assign subtype such as Multi-Locus Sequnce Typing. 
+No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
+An example of a subtype system that has been adpated for WGS is serotyping in \textit{Escherichia coli}.
+The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O group serotype in \citep{Joensen2015,Ingle2016}.
+Another example, is the Shiga toxin subtype (Stx); a DNA-based subtying scheme generated using PCR.
+The tool in \citep{CARRILLO2016} predicts Stx subtype by simulating PCR in silico.
 
 Phylotyper is a novel \textit{in silico} predictor of subtypes from sequence data.
+Similar to \citep{Joensen2015,Ingle2016,CARRILLO2016}, it also works on subtypes that can be predicted from specific, pre-selected gene or genomic region sequences.
 Phylotyper is unique in that it builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences to help inform subtype prediction.
-Using phylogenetic ancestral state reconstruction to assign the likelihood of each subtype to the tree branch points, Phylotyper assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
+Using phylogenetic ancestral state reconstruction to assign the likelihood of each subtype to the tree branch points, Phylotyper then assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.
 
@@
 The steps in the Phylotyper pipeline are: 
-(1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}. Hits that do not align over 95\% or have under 90\% sequence identity with reference sequence are discarded.
+(1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}.
+Hits that do not align over 95\% or have under 90\% sequence identity with reference sequence are discarded.
+Users are notified if no loci are found in genome.
 (2) Align input genes against a pre-aligned set of reference genes using MAFFT's \texttt{--add} feature \citep{Katoh2013}.
@@
 Users are also provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g. Figure~1\vphantom{\ref{fig:01}}).
+
+Detailed descriptions on how to run Phylotyper and the underlying methods are provided here: \url{https://github.com/superphy/insilico-subtyping}.
 
@@
 To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
-Phylotyper assumes that the provided training sequence are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
+Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
 Where possible, checks are built-in to the \texttt{new} pipeline to ensure the assumptions of the Phylotyper approach are not violated.
-Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline.
-Results are reported to the user.
-The first test checks the the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
+Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
+The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
 This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
-A second check computes the accuracy measure; F-score, through a leave-one-out cross-validation.
-This validation uses each sequence in the training set as a test input (the test sequence is removed from the tree and then used as input).
-Results are recorded. 
-Metrics like precision, recall and F-score rapidly erode as the phylogenetic grouping is decoupled from the subtypes.
-These checks will verify that the sequences phylogenetic grouping is predictive of the subtype subtype.
-There is no check that can verify the training set covers all subtypes; this is left up to the user.
-However, as we show in the Results section, Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set.
+A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
+The performance metrics: recall and F-score rapidly decrease as the correspondence of training set phylogeny and subtype distribution decreases (Supplementary Figure)
+These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
+There is no check that can confirm the training set covers all subtypes in a particular scheme; it is important that the user monitor and update schemes as gaps are identified.
+Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
+Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).
 
@@
 
-All new or updated subtype schemes added in Phylotyper are subject to a leave-one-out cross-validation test as described earlier.  
+All new or updated subtype schemes added in Phylotyper are subject to a leave-one-out cross-validation test.
 The test is part of the \texttt{add} pipeline and is used to estimate the F-score of the subtype scheme. 

### 2017-05-16 manuscript/phylotyper_suppl_info_revised.tex
UPDATE: intro UPDATE: implementation UPDATE: suppl info UPDATE: response to reviewers

@@
 \section{Supplementary Methods}
+
+\subsection{Phylotyper Verifications Performed on New Subtype Schemes}
+
+Phylotyper conducts two tests for submitted new subtype schemes.
+The purpose of the first check is to identify isolated cases of distinct subtypes that are clustered together in the phylogenetic tree.
+In large-scale cases, descendent clades in the phylogenetic tree with multiple distinct subtypes may indicate a subtype that is not correlated with the phylogeny.
+In limited cases, it might indicate annotation errors.
+Phylotyper computes an inter-patritristic distances for all leaves in the subtype phylogenetic tree for the given subtype scheme.
+It then computes a threshold is equivalent to 0.4 probability that inter-patristic distances equal to or greater then threshold belong to the same-subtype distribution.
+To estimate the distributions, all inter-patristic distances for nodes with the same subtype and all distances for nodes with different subtypes are fit with R's Mclust package.
+Subtrees with a max inter-patristic distance less than this threshold that have distinct subtypes are flagged.
+Users will be notified and an updated subtype scheme will be generated with proposed corrections when this situation is detected.
+
+All new subtype schemes are subject to a leave-one-out cross-validation analysis. You will be notified if the F1-score (a equally weighted average of precision and recall) is below 0.9.
 

### 2017-05-26 manuscript/response_to_reviewers.txt
UPDATE: response to reviewers

@@
 response_to_reviewers.txt
+
+Firstly, we would like to thank the reviewers for their thoughtful comments.
+We have made subtantual changes to the results and discussion sections as well as to the Phylotyper code that we hope address your suggestions for the Phylotyper manuscript.
+We have responded to each individual comment below:
 
@@
 *Response:
-We agree with the reviewer's suggestion that a working example provided for new users will enhance the adoption and usability of Phylotyper.
-We have added an "Example" section to the README.md file (which is also used as the landing page for Phylotyper's GitHub repository: https://github.com/superphy/insilico-subtyping/) along with example data that demonstrates how to 1) predict subtypes for schemes already packaged in Phylotyper and 2) how to add and run user-defined subtype schemes.
+We agree with the reviewer's suggestion that a working example for new users will enhance the adoption and usability of Phylotyper.
+We have added an "Example" section to the README.md file (which is also the landing page for Phylotyper's GitHub repository: https://github.com/superphy/insilico-subtyping/#examples) along with example data that demonstrates how to 1) predict subtypes for schemes already packaged in Phylotyper and 2) how to add and run user-defined subtype schemes.
 
@@
 The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor.
-To mitigate poor prediction from unsuitable or inaccurate inputs, we have included in the build pipeline for the local user-submitted subtype feature a series of checks to automatically test the performance of any submitted subtype scheme.
+To mitigate poor prediction from unsuitable or inaccurate inputs, we included in the build pipeline for the local user-submitted subtype feature a series of checks to automatically test the performance of any submitted subtype scheme.
 These tests 1) verify that the phylogeny of the provided sequences places the same subtypes into phylogenetic groups and 2) that the subtype can be accurately predicted with the given input sequences by running a leave-one-out cross-valiation.
@@
 
-The reviewer suggests having a means for users to submit subtype for evaluation and posting to a shared repository.
+The reviewer suggests having a means for users to submit subtypes for evaluation and posting to a shared repository.
 We agree that this would be a helpful system.
@@
 
+*Response*:
+We added a high-level description of mechanisms currently in use to perform subtyping, namely phenotype- and DNA-based methods, with specific examples in each category.
+We feel this will give the readers an understanding of the breadth of subtyping methods.
+We also describe the characteristics of the subtypes that have been adapted for in silico prediction using WGS data.
+Our intention was to convey the wide range of subtyping approaches, and how certain subtyping  can be considated and recapitulated with WGS data.
+
 Implementation:
@@
 
+*Response*:
+A detailed step-by-step account of how to run Phylotyper's automated build pipeline is provided in the Phylotyper package instructions.
+We added a reference and link to these instructions in the Implementation section.
+We also added to the Implmentation section a description of the requirements in Phylotyper for new subtype schemes as well as the checks that the pipeline automatically performs to validate subtypes schemes created by users.
+In the Discussion section, we elaborate on the ideal sequence regions that should be targeted by users when developing a WGS-based predictor of subtypes with Phylotyper.
+Also in the Discussion seciton, we name some current examples of subtyping schemes in other pathogens outside of E.
+coli that can be predicted with Phylotyper.
+We did not systematically go through all main pathogens to identify the subtyping schemes that can predicted with Phylotyper.
+As we state in our revised Discussion, the subtype schemes that can be linked biologically to a specific genome sequence region are the best candidates for WGS-based predictors.
+To compile a complete list of these type of subtypes requires domain knowledge in the pathogen (Our lab works with E.
+coli, hence the the initial focus in Phylotyper's resources on E.
+coli).
+We did identify and mention select examples in diverse pathogens to show that Phylotyper is not limited to E.
+coli.
+Phylotyper can work with any suitable subtype; identifying a suitable subtype, however, requires knowledge of the subtype and its underlying biological mechanism.
+To help expand Phylotyper to other pathogens, we added a request in the manuscript and on the Phylotyper website for Microbiologists to send us either subtype schemes or additions/corrections to existing subtype schemes.
+Subtype schemes that we recieve we will be packaged into the Phylotyper repository for download and general use.
+
 3.      Please provide more details regarding the “assumptions of the Phylotyper approach” and how/why they would be violated when building new subtype schemes.
+
+*Response*: 
+The Phylotyper assumptions, and its mechanisms to detect possible violations, are now outlined in the Implementation section.
+The revised Implementat covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.
 
@@
 Moreover, a comparison - at least theoretical - with other existing tools would strengthen the uniqueness of Phylotyper highlighted in the discussion.
+
+*Response*:
+Our main goal in the performance assessment was to compare how the prediction methods generalize to an independent dataset.
+There may exists differences in the reference training sets that underpin the tools, and these differences may mask the true performance of the method.
+We wanted to evalute the predictive capability of the methods themselves.
+SeroTypeFinder is provided as a web service.
+The software for SeroTypeFinder has not been made publicly available.
+Without access to the software, we cannot manipulate the training and test datasets to perform cross-validation analysis.
+The EcOH tool works directly on unassembled small reads and not assembled genome sequences.
+A direct comparison of the performance with these tools with Phylotyper is not possible.
+The intention of using a general BLAST-based approach is demonstrate the benefit of Hidden State Prediction with a phylogenetic tree over approaches that directly match input sequences against a reference database.
+BLAST, as used in our assessment with a percent identity and coverage cutoff, is the basis of the SeroTypeFinder tool.
+We used a more stringent cutoff however.
+OUur simulations indicated that most subtype alleles had higher alignment metrics than the reported to be used in SeroTypeFinder.
+EcOH employs a more sophisticated algorithm for matching alleles in order to filter out spurious small reads that might align by chance.
+At its core, it is still matching alleles and does not develop a model of allele sequence evolution to define boundaries on subtype sequences.
+We feel that our comparison to the generalized BLAST-based method does make a valid case for HSP in subtype prediction.
+
+We added a direct comparison of Phylotyper to Vtyper for Stx genes.
+The performance results are summerized in the Results section and further details are provided in Supplementary Information.
+We were cautious in making the test set; all test genes were required to have sufficient upstream and downstream sequence to for the in silico PCR approach to work.
+Phylotyper has no such constraint.
 
@@
 1.      The abstract mentions a python package only whereas the text in the implementation mentions python and R. Please revise where necessary.
+
+*Response*:
+Fixed.
+
 2.      “The main input into Phylotyper is assembled genome sequences”. Please mention the format type (apparently fasta).
+
+*Response*:
+Fixed.
+
 3.      “… in addition to Stx include intimin and serotyping” Please revise the sentence.
+
 4.      Replace aslo by also
+
+*Response*:
+Fixed.
+
 5.      Revise the sentence “This type of approach, there is…”.
@@
 7.      The colors used by Phylotyper could be improved to increase readability of the output for the allele labels. There are several good packages to select colors within R or python.
+*Response*:
+Switched to discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper uses
 
@@
 8.      Please plot the legend so that it does not overall the labels
+*Response*:
+Change made
 9.      Labels are too close and therefore difficult to read if there were several query allele as an input.

### 2017-05-26 manuscript/phylotyper_revised.tex
UPDATE: response to reviewers

@@
 The phylogenic framework in Phylotyper improves accuracy, provides useful contextual feedback, and is more capable of identifying novel subtypes over approaches based solely on sequence similarity.\\
-\textbf{Availability and Implementation:} Phylotyper is a python package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
+\textbf{Availability and Implementation:} Phylotyper is a python and R package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
 \textbf{Contact:} \href{matthew.whiteside@phac-aspc.gc.ca}{matthew.whiteside@phac-aspc.gc.ca}\\
@@
 Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
-To predict from WGS data, these tools target subtypes (Phenotype-based or DNA-based) that can be linked to a specific region or gene in the genome.
-These tools have not been applied to subtyping methods that already directly use DNA sequence data to assign subtype such as Multi-Locus Sequnce Typing. 
+To predict from WGS data, these tools have targeted subtypes (either Phenotype-based or DNA-based systems) that can be casually linked to a specific region or gene in the genome.
+\textit{In silico} subtyping has not been applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequnce Typing (MLST). 
 No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
 An example of a subtype system that has been adpated for WGS is serotyping in \textit{Escherichia coli}.
-The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O group serotype in \citep{Joensen2015,Ingle2016}.
+The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O-antigen group serotype in \citep{Joensen2015,Ingle2016}.
 Another example, is the Shiga toxin subtype (Stx); a DNA-based subtying scheme generated using PCR.
-The tool in \citep{CARRILLO2016} predicts Stx subtype by simulating PCR in silico.
+The tool in \citep{CARRILLO2016} predicts Stx subtype by simulating PCR \textit{in silico}.
 
@@
 (1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}.
+Inputs are in fasta format.
 Hits that do not align over 95\% or have under 90\% sequence identity with reference sequence are discarded.
@@
 Phylotyper was designed to be incorporated into a WGS workflow.  
-The main input into Phylotyper is assembled genome sequences.  
+The main input into Phylotyper is assembled genome sequences (in fasta format).  
 Putative loci needed for the subtype scheme are identified in the input genomes using BLAST \citep{Camacho2009}.
@@
 coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotype O- and H-types  (Supplementary Table~1).
-However, the Phylotyper software also has the capability to add new subtype schemes. 
+However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
 Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
@@
 A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
-The performance metrics: recall and F-score rapidly decrease as the correspondence of training set phylogeny and subtype distribution decreases (Supplementary Figure)
+The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure)
 These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
@@
 Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).
-
-To grow the Phylotyper resource, we encourage users to contact us with their new subtype schemes or updates to schemes (\url{https://github.com/superphy/insilico-subtyping}). 
-We will continue to expand the Phylotyper resource by adding and updating high-quality subtype schemes in the Phylotyper repository.
 
@@
 Currently the Phylotyper software offers subtyping schemes for \textit{E. coli}.
-It can, however, be applied to most molecular subtype schemes and Phylotyper includes functionality to build new schemes.
+It can, however, be applied to other molecular subtype schemes and Phylotyper includes functionality to build new schemes.
+Phylotyper can produce predicted subtypes from any input sequence that is strongly correlated with the subtype distribution, however, input sequences with a direct biological casual link to the subtype will have fewer cavaets; A gene sequence that is casual cannot become dissassociated from the subtype through recombination or horizontal gene transfer as a gene sequence that is merely correlated with the subtype.
+Outside of \textit{E.
+coli}, the PCR-based capsular typing system for \textit{Haemophilus influenzae}, Neurotoxin serotyping in \textit{Clostridium botulinum} and the haemagglutinin (HA) and neuraminidase (NA) types in Influenza A virus are all examples of potential future subtype schemes that can be incorporated into Phylotyper.
+We plan on expanding the Phylotyper resource by adding and updating high-quality subtype schemes for other pathogens.
+We encourage users to contact us with their new subtype schemes or updates to schemes (\url{https://github.com/superphy/insilico-subtyping}).
 
-One of the strategies for predicting subtypes is to use sequence similarity to annotated loci alleles.
+The main strategy currently in use by other \textit{in silico} tools for predicting subtypes is to use sequence similarity to annotated loci alleles.
 Query genomes or genes are matched to alleles that are attributed the subtype phenotype or are correlated with the distribution of the subtype.
@@
 With a strict allele matching stategy, the reference or training set make-up can be critical to performance.
-When alleles or even subtypes are missing, the sequence similarity approach more frequently generates type-I errors.
+When alleles or even subtypes are missing in the reference database, the sequence similarity approach more frequently generates type-I errors.
 Alternatively, Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.

### 2017-05-29 manuscript/response_to_reviewers.txt
ADDED: responses to reviewer 2.

@@
 We also describe the characteristics of the subtypes that have been adapted for in silico prediction using WGS data.
-Our intention was to convey the wide range of subtyping approaches, and how certain subtyping  can be considated and recapitulated with WGS data.
+Our intention was to convey the wide range of subtyping approaches, and how certain subtyping approaches can be recapitulated from WGS data.
 
@@
 *Response*: 
-The Phylotyper assumptions, and its mechanisms to detect possible violations, are now outlined in the Implementation section.
-The revised Implementat covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.
+The Phylotyper assumptions are now outlined in the Implementation section.
+The revised Implementation section covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.
 
@@
 Our main goal in the performance assessment was to compare how the prediction methods generalize to an independent dataset.
-There may exists differences in the reference training sets that underpin the tools, and these differences may mask the true performance of the method.
-We wanted to evalute the predictive capability of the methods themselves.
+There may exists differences in the reference training sets that underpin the tools, and these differences may mask the true performance of the method and we wanted to evalute the predictive capability of the methods themselves.
 SeroTypeFinder is provided as a web service.
@@
 BLAST, as used in our assessment with a percent identity and coverage cutoff, is the basis of the SeroTypeFinder tool.
-We used a more stringent cutoff however.
-OUur simulations indicated that most subtype alleles had higher alignment metrics than the reported to be used in SeroTypeFinder.
 EcOH employs a more sophisticated algorithm for matching alleles in order to filter out spurious small reads that might align by chance.
 At its core, it is still matching alleles and does not develop a model of allele sequence evolution to define boundaries on subtype sequences.
-We feel that our comparison to the generalized BLAST-based method does make a valid case for HSP in subtype prediction.
+We feel that our performance comparison to the generalized BLAST-based method does make a valid case for using HSP in subtype prediction.
+We did add a direct comparison of Phylotyper to Vtyper for Stx genes.
+The performance results are summerized in the Results section and details on how we selected the test set are provided in Supplementary Information.
 
-We added a direct comparison of Phylotyper to Vtyper for Stx genes.
-The performance results are summerized in the Results section and further details are provided in Supplementary Information.
-We were cautious in making the test set; all test genes were required to have sufficient upstream and downstream sequence to for the in silico PCR approach to work.
-Phylotyper has no such constraint.
+To address the reviwers comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis.
+Our aim with the revised descriptions is to provide readers sufficient information give them a complete understanding of the mechanics of the methods.
+Implementation details, such as percent identity cutoff or number of iterations, remain in Supplementary Information.
+This was done to fit with page constraints of the article type.
+
+As suggested, we also added a theoretical comparison of the tools to the Discussion; the theoretical comparison looks at use cases, potential issues, and additional benefits outside of performance for the tools.
 
@@
 What would be the recommended tests/further analysis recommended to researchers using Phylotyper in the case a new subtype is identified?
+
+*Response*:
+
 6.
@@
 They did not particularly emphasize discrepancies between existing molecular typing scheme and whole genome data as suggested by Whiteside et al.
+
+*Response*:
+
+We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and genome region that is used for prediction.
 
@@
 
+*Response*:
+Change made.
+
 4.      Replace aslo by also
@@
 
+*Response*:
+Sentence replaced.
+
 Figure 1.
@@
 Or remove if this is redundant information.
+
+*Response*:
+Removed the subtype letters at the end of the label; this was redundant information that was also conveyed by the color.
+Added information in the figure description on the meaning of the colors.
+Fixed labels so they fit in the window.
+
 7.      The colors used by Phylotyper could be improved to increase readability of the output for the allele labels. There are several good packages to select colors within R or python.
+
 *Response*:
-Switched to discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper uses
+Switched to discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper uses RColorRamp and this palette to generate sufficient colors for all subtypes.
 
@@
 8.      Please plot the legend so that it does not overall the labels
+
 *Response*:
-Change made
+Change made.
+
 9.      Labels are too close and therefore difficult to read if there were several query allele as an input.
+
+*Response*:
+Spacing increased.
 
@@
 1.      Four or five E. coli schemes?
+
+*Response*:
+The E. coli serotypes available in Phylotyper comprise two schemes: O- and H-type (here incorrectly counted as one), resulting in five tested schemes. The reference to four has been removed.
 

### 2017-05-29 manuscript/phylotyper_revised.tex
ADDED: responses to reviewer 2.

@@
 The subtype marginal likelihood is displayed at each node as a pie chart.
+Subtype is indicated by color as shown in the legend.
 The full Stx2 tree is displayed in Supplementary Figure S1.}\label{fig:01}
@@
 
-Currently, subtype schemes for \emph{Escherichia coli} (\textit{E.
-coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotype O- and H-types  (Supplementary Table~1).
+Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli} (\textit{E. coli}): Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
 However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
@@
 Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
-Where possible, checks are built-in to the \texttt{new} pipeline to ensure the assumptions of the Phylotyper approach are not violated.
+Where possible, checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
 Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
@@
 
-Phylotyper marks a progression from the sequence-similarity approach widely used in \it{in silico} subtyping strategies.
-It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that not in the training data or reference database.
+Phylotyper marks a progression from the sequence-similarity approach that is the basis of multiple \it{in silico} subtyping strategies.
+It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that are not in the training data or reference database.
 To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
@@
 For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
+This generalized BLAST tool assigns a subtype when the top BLAST match for an unknown query gene against an annotated reference database is above a pre-selected percent identify and alignment coverage; the query is given the subtype from the top match in this case
 Details for the BLAST tool and how the assessment was conducted are available in supplementary methods section.
@@
 To demonstrate this property, we randomly assigned subtypes for increasing proportions of the genes in the training set and computed the F-score with the leave-one-out validation for each proportion level.
-The F-score and recall rapidly decrease as the proportion of altered subtypes increased (Supplementary Figure S2).
+The F-score and recall rapidly decrease as the proportion of randomly altered subtypes increases (Supplementary Figure S2).
 

### 2017-05-30 manuscript/response_to_reviewers.txt
EDIT: finished response to reviewers

@@
 The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor.
-To mitigate poor prediction from unsuitable or inaccurate inputs, we included in the build pipeline for the local user-submitted subtype feature a series of checks to automatically test the performance of any submitted subtype scheme.
-These tests 1) verify that the phylogeny of the provided sequences places the same subtypes into phylogenetic groups and 2) that the subtype can be accurately predicted with the given input sequences by running a leave-one-out cross-valiation.
-The phylogenetic test empirically determines the distribution of inter-patristic distance between the sequences in the tree with the same subtypes.
-It uses the distribution to identify subtree groupings that place different subtypes in the same clade that are expected to be the same.
-Users are notified if any unexpected subtype assignments are detected in the input.
-In our experience, this test was effective at identifying a limited number of anomylous subtype assignments, for example, incorrectly annotated sequences in the larger subtype scheme.
-In the cross-validation test, each input reference sequence is withheld from the training set and used to test the predictor.
-Users are notified if the subtype scheme has a F1-score below 0.9 from the cross-validation test.
-We found that this test identified subtypes schemes that have poor predictive power due to the subtype being weakly phylogenetically correlated with the input sequences or subtypes being underrepresented in the training set.
-While these checks identify major breakdowns in the requirements for an accurate subtype predictor, we note there is no check to identify missing data, specifically, missing subtypes or key subtypes gene sequences that define the range of the subtype sequence space.
+To mitigate poor prediction from unsuitable or inaccurate inputs, we included in the build pipeline for the local user-submitted subtypes, a series of checks to automatically test the performance of any submitted subtype scheme.
+These checks are now described in the Implementation section.
+
+While these checks identify major breakdowns in the training set used by Phylotyper, we note there is no check to identify missing data, specifically, missing subtypes or key subtypes gene sequences that define the range of the subtype sequence space.
 For any current general subtype predictor (i.e.
@@
 The added content describes the two checks, how they work and what they detect.
-In the README we also offer some guidelines about the inputs needed for an effective subtype predictor.
+In manuscript we also provide guidelines about the inputs needed for an effective subtype predictor.
 We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the trainging set and used as test input.
@@
 We agree that this would be a helpful system.
-It will allow microbiologist specialists to submit and improve subtype schemes and increase the
-We have added a section to the GitHub page, subtype scheme will be checked by us and added to the repository or as a separate download.
+It will allow microbiologist specialists to submit and improve subtype schemes and increase coverage.
+We mention in the manscript and on the README, that users can contact us and submit new or modified subtype schemes.
+We will check the submitted schemes and add them to our Github repository for general use.
 
@@
 EcOH employs a more sophisticated algorithm for matching alleles in order to filter out spurious small reads that might align by chance.
-At its core, it is still matching alleles and does not develop a model of allele sequence evolution to define boundaries on subtype sequences.
+At its core, it is still matching alleles and does not develop a model of subtype/allele sequence evolution to define boundaries on the subtypes.
 We feel that our performance comparison to the generalized BLAST-based method does make a valid case for using HSP in subtype prediction.
 We did add a direct comparison of Phylotyper to Vtyper for Stx genes.
-The performance results are summerized in the Results section and details on how we selected the test set are provided in Supplementary Information.
+The performance results are summerized in the Results section.
 
-To address the reviwers comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis.
+To address the reviwers comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis in the manuscript.
 Our aim with the revised descriptions is to provide readers sufficient information give them a complete understanding of the mechanics of the methods.
 Implementation details, such as percent identity cutoff or number of iterations, remain in Supplementary Information.
-This was done to fit with page constraints of the article type.
+This was done to fit with page constraints of the Application Note format.
 
@@
 *Response*:
+The point we were trying make was that more frequently, Phylotyper correctly handles novel subtypes than existing methods.
+In our emperical testing we found Phylotyper makes fewer Type-I errors when encountering a novel allele of an existing subtype or a allele with a entirely new subtype not found in reference training set.
+In former case, Phylotyper returns non-significant/undetermined.
+We have clarified this point throughout the manuscript, removing references to Phylotyper identifying novel subtypes and replacing it with the more nuanced description that Phylotyper more frequently properly classifies novel subtypes as a negative result or correctly classifies novel alleles with subtypes in the database.
+To strengthen this statement, we have as the reviewers suggest, remove entire subtypes from the training set and calculated the false positive rate.
+Phylotyper was found to have a much lower false positive rate when classifying genes with unknown subtypes.
+
+Phylotyper does not have the ability to create subtype schemes de novo, or to add new subtype classes on the fly.
+We have not explored this potential application of the Phylotyper method.
+It would be an interesting avenue to explore as using a phylogenetic model of subtype evolution to define boundaries for new or existing subtypes might make a nice complement for existing de novo stragies like MLST that assign every new allele a separate subtype.
+However, our goal in this manuscript is to present an alternative genome-based subtype prediction method.
+De novo subtype creation is outside our intended applications for Phylotyper.
+Again, we corrected all references to Phylotyper identifing novel subtypes.
 
@@
 We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and genome region that is used for prediction.
+We cite Debroy et al., as an example of how the associated genetic data and subtypes can have discrepancies.
 
@@
 We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set to simulate performance when encountering a novel subtype.
-This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach.
+This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach, and more frequently returns a negative result.
 
@@
 
-Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations were linkage disequlibrium can occur.
+Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations where the genotype is only correlated with subtype and not causative.
 If the sequences used to build the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E.
 coli being used to predict O-group serotypes, then the prediction would directly be linked to underlying genome.
-If however a proxy sequence is used, HGT could impact.
-We have recommended to users building subtype schemes they focus on
+If however unrelated, correlated sequence is used, HGT could disconnect a particular genotype from its subtype.
+We have added to the manuscript recommendations for users building subtype schemes to focus on genomic regions that have a direct biological role in determining the subtype to reduce possibility of impact by HGT.
 
@@
 1) Typo in line 49: also not aslo
+
+*Response*:
+Fixed
+
 2) Line 51 mentioned assumptions for building new subtype pipeline.  It'll be good to explain that or if space is an issue, point to the portion of the manual that explains these assumptions
+
+*Response*:
+We have added descriptions of these assumputions to the mansucript.
+
 ----------------------------------------

### 2017-05-30 manuscript/phylotyper_revised.tex
EDIT: finished response to reviewers

@@
 Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
-The phylogenic framework in Phylotyper improves accuracy, provides useful contextual feedback, and is more capable of identifying novel subtypes over approaches based solely on sequence similarity.\\
+The phylogenic framework in Phylotyper improves accuracy and provides useful contextual feedback, then approaches based solely on sequence similarity.\\
 \textbf{Availability and Implementation:} Phylotyper is a python and R package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
@@
 These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
-There is no check that can confirm the training set covers all subtypes in a particular scheme; it is important that the user monitor and update schemes as gaps are identified.
+There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are covered in the trainging set.
+It is important that the user monitor and update schemes as gaps are identified.
 Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.

### 2017-05-30 manuscript/phylotyper_revised.tex
EDIT: abstract

@@
 Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
-The phylogenic framework in Phylotyper improves accuracy and provides useful contextual feedback, then approaches based solely on sequence similarity.\\
+The phylogenic framework in Phylotyper improves accuracy and provides useful contextual feedback, compared to approaches based solely on sequence similarity.\\
 \textbf{Availability and Implementation:} Phylotyper is a python and R package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\
@@
 
-Phylotyper marks a progression from the sequence-similarity approach that is the basis of multiple \it{in silico} subtyping strategies.
+Phylotyper marks a progression from the sequence-similarity approach that is the basis of multiple \textit{in silico} subtyping strategies.
 It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that are not in the training data or reference database.
@@
 A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set.
-The test Stx genes had sufficent surrounding DNA sequence to support \it{in silico} PCR.
+The test Stx genes had sufficent surrounding DNA sequence to support \textit{in silico} PCR.
 In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
@@
 Query genomes or genes are matched to alleles that are attributed the subtype phenotype or are correlated with the distribution of the subtype.
-For example SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textid{in silico} O-typing, and the flagellin genes for H-typing \textit{E.
+For example SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textit{in silico} O-typing, and the flagellin genes for H-typing \textit{E.
 coli} genes.

### 2017-06-01 manuscript/phylotyper_revised.tex
EDIT: abstract updated

@@
 \abstract{\textbf{Summary:} Whole genome sequencing (WGS) is being adopted in public health for improved surveillance and outbreak analysis.
-Subtyping; a  is used in public health to infer phenotypes and distinguish bacterial strain groups.
+In public health, subtyping has been used to infer phenotypes and distinguish bacterial strain groups.
 \textit{In silico} tools that predict molecular subtypes from sequences data are needed to transition historical data to WGS-based protocols.
 Phylotyper is a novel solution for \textit{in silico} subtype prediction from gene sequences.
-Designed for incorporation into WGS pipelines, it is a general prediction tool that can be applied to most molecular subtype schemes.
+Designed for incorporation into WGS pipelines, it is a general prediction tool that can be applied to different subtype schemes.
 Phylotyper uses phylogeny to model the evolution of the subtype and infer subtypes for unannotated sequences.
-The phylogenic framework in Phylotyper improves accuracy and provides useful contextual feedback, compared to approaches based solely on sequence similarity.\\
+The phylogenic framework in Phylotyper improves accuracy over approaches based solely on sequence similarity and provides useful contextual feedback.\\
 \textbf{Availability and Implementation:} Phylotyper is a python and R package. It is available from: \url{https://github.com/superphy/insilico-subtyping}.\\

### 2017-06-01 manuscript/phylotyper_revised.tex
EDIT: discussion results updated.

@@
 In public health, subtyping has been used to infer phenotypes and distinguish bacterial strain groups.
-\textit{In silico} tools that predict molecular subtypes from sequences data are needed to transition historical data to WGS-based protocols.
+\textit{In silico} tools that predict subtypes from sequences data are needed to transition historical data to WGS-based protocols.
 Phylotyper is a novel solution for \textit{in silico} subtype prediction from gene sequences.
@@
 DNA-based subtypes use a variety of methods from PCR to Pulse Field Gel Electrophoresis to DNA sequencing to assign groups to bacteria \citep{Wiedmann2002}.
-As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
+As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes.
 A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain\citep{Jenkins2015}.
 Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
-To predict from WGS data, these tools have targeted subtypes (either Phenotype-based or DNA-based systems) that can be casually linked to a specific region or gene in the genome.
-\textit{In silico} subtyping has not been applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequnce Typing (MLST). 
+To predict from WGS data, these tools have targeted subtypes (either Phenotype-based or DNA-based systems) that can be predicted by sequence variation in specific region or gene in the genome.
+\textit{In silico} subtype prediction is not applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequnce Typing (MLST). 
 No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
@@
 Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
-Where possible, checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
-Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
-The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
+Checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
+Each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
+The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from different subtypes.
 This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
-A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
-The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure)
-These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
-There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are covered in the trainging set.
+A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (see Supplementary Methods for details).
+The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure S2).
+These checks verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
+There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are represented in the training set.
 It is important that the user monitor and update schemes as gaps are identified.
 Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
-Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).
 
@@
 
-Phylotyper marks a progression from the sequence-similarity approach that is the basis of multiple \textit{in silico} subtyping strategies.
-It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that are not in the training data or reference database.
+Phylotyper is a progression from the sequence-similarity approach that is the basis of current \textit{in silico} subtype prediction strategies.
 To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
-The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a training set which excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
+The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a reduced training set that excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
 This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
-The second validation examined how the predictors respond when presented with a gene that has a subtype not found in the reference set.
+The second validation examined how the predictors perform when tested with a gene that has a subtype, not in the reference set.
 In this validation, each subtype was iterated over and all genes that are assigned the subtype were removed from the training set.
 In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
-The correct response for the predictors was to return a negative result, since the subtype does not exist in the training set.
+The correct response for the predictors was to return a negative result since the subtype does not exist in the training set.
 For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
-This generalized BLAST tool assigns a subtype when the top BLAST match for an unknown query gene against an annotated reference database is above a pre-selected percent identify and alignment coverage; the query is given the subtype from the top match in this case
-Details for the BLAST tool and how the assessment was conducted are available in supplementary methods section.
+This generalized BLAST tool, based on the approach used in \citep{Joensen2015}, assigns a query sequence a subtype when the top BLAST match from an annotated reference database is above a pre-selected percent identity and alignment coverage cutoff.
+Details how the assessment was conducted are available in Supplementary methods.
 The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
-When tasked with assigning a novel gene sequence (i.e. one that is not in the training set) in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST hit approach.
-The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST hit approach(result details are available in Supplementary Table S2).
-In order to achieve this level of precision, the BLAST approach often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper
-Similarily, when entire subtype classes were withheld from the training set, Phylotyper had consistenly lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
-A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set.
-The test Stx genes had sufficent surrounding DNA sequence to support \textit{in silico} PCR.
+When tasked with assigning a novel gene sequence not in the training set in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST-hit approach.
+The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST-hit approach (Supplementary Table S2).
+The BLAST approach also often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper.
+Similarly, when entire subtype classes were withheld from the training set, Phylotyper had consistently lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
+A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set \citep{CARRILLO2016}.
+The test Stx genes had sufficient surrounding DNA sequence to support \textit{in silico} PCR.
 In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
-Phylotyper can correctly predict the subtype for all these genes.
-Based on this level of recall, it appears conditions in the Stx subtype setting are challenging for simulated PCR.
+Phylotyper correctly predicted the subtype for all these genes.
+Based on this level of recall, it appears conditions in the Stx subtype environment are challenging for simulated PCR.
 
@@
 
-From assembled WGS data, Phylotyper can assign unclassifed strains a molecular subtype.
-Currently the Phylotyper software offers subtyping schemes for \textit{E. coli}.
-It can, however, be applied to other molecular subtype schemes and Phylotyper includes functionality to build new schemes.
-Phylotyper can produce predicted subtypes from any input sequence that is strongly correlated with the subtype distribution, however, input sequences with a direct biological casual link to the subtype will have fewer cavaets; A gene sequence that is casual cannot become dissassociated from the subtype through recombination or horizontal gene transfer as a gene sequence that is merely correlated with the subtype.
+From assembled WGS data, Phylotyper can assign unclassified strains a subtype.
+Currently, the Phylotyper software offers subtyping schemes for \textit{E. coli}.
+It can, however, be applied to other subtype schemes and Phylotyper includes functionality to build new schemes.
+Phylotyper can produce predicted subtypes from any input sequence that is strongly correlated with the subtype distribution, however, input sequences with a direct biological causal link to the subtype will have fewer caveats; A gene sequence that is causal cannot become disassociated from the subtype through recombination or horizontal gene transfer.
 Outside of \textit{E.
-coli}, the PCR-based capsular typing system for \textit{Haemophilus influenzae}, Neurotoxin serotyping in \textit{Clostridium botulinum} and the haemagglutinin (HA) and neuraminidase (NA) types in Influenza A virus are all examples of potential future subtype schemes that can be incorporated into Phylotyper.
+coli}, the PCR-based capsular typing system for \textit{Haemophilus influenzae}, Neurotoxin serotyping in \textit{Clostridium botulinum} and the haemagglutinin and neuraminidase types in Influenza A virus are all examples of potential future subtype schemes that we are incorporating into Phylotyper.
 We plan on expanding the Phylotyper resource by adding and updating high-quality subtype schemes for other pathogens.
@@
 
-The main strategy currently in use by other \textit{in silico} tools for predicting subtypes is to use sequence similarity to annotated loci alleles.
+The main strategy currently in use by other \textit{in silico} tools for predicting subtypes is to use sequence similarity to annotated gene alleles.
 Query genomes or genes are matched to alleles that are attributed the subtype phenotype or are correlated with the distribution of the subtype.
-For example SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textit{in silico} O-typing, and the flagellin genes for H-typing \textit{E.
-coli} genes.
-O-type and H-type are transfered from the top matches to the queries provided they are above coverage and percent identity thresholds.
-This general strategy of allele matching is also applied in EcOH database, however, the EcOH can directly use unassembled sequence reads as input.
+For example, SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textit{in silico} O-typing and the flagellin genes for H-typing \textit{E.
+coli} genes \citep{Jenkins2015}.
+O-type and H-type are transferred from the top matches to the queries provided they are above coverage and percent identity thresholds.
+This general strategy of allele matching is also applied in the EcOH tool \citep{Ingle2016}, however, the EcOH tool can directly use unassembled sequence reads as input.
 The EcOH tool aligns reads to alleles linked to \textit{E.
 coli} O-types and H-types, and identifies the top candidates that have an alignment score above pre-defined thresholds.
-Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
-With a strict allele matching stategy, the reference or training set make-up can be critical to performance.
-When alleles or even subtypes are missing in the reference database, the sequence similarity approach more frequently generates type-I errors.
+Phylotyper is comparatively more robust as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set.
+With the allele matching strategy, the reference set make-up can have a greater impact on performance.
+When alleles or even subtypes are missing in the reference database, the sequence similarity approach more frequently generates false positive predictions.
 Alternatively, Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
-By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make type-I error when encountering a novel subtype or allele. 
-The empircal testing we performed demonstrated this behavior; 
+By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make a Type-I error when encountering a novel subtype or allele. 
+The empirical testing we performed demonstrated this behavior; 
 the rate of false positive classifications was significantly lower than in a sequence-similarity approach in validations where we withheld an allele or an entire subtype from the training set and used it as a test input.
-V-typer takes a distinct approach; it directly simulates the \textit{in vitro} wet-lab PCR procedure used to perform Stx subtyping. 
+V-typer takes a distinct approach; it directly simulates the \textit{in vitro} wet-lab PCR procedure used to perform Stx subtyping \citep{CARRILLO2016}. 
 V-typer's direct replication of a wet-lab method \textit{in silico} means it can only be applied to subtypes schemes that use PCR. 
 Additionally, we found in our evaluation of Stx subtypes that it failed to generate predictions for most test cases.
-From a broader methodology viewpoint, one of the benefits of the phylogenetic framework in Phylotyper, is that it provides a statistical likelihood for interpreting results.
+From a methodology standpoint, Phylotyper has an additional benefit over current methods; the phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting results.
 In comparison, there is no built-in mechanism in the sequence similarity approach to quantify the level of confidence in assigning alleles a subtype.
@@
 A recent analysis of O-antigen serotypes and their associated O-antigen gene sequences in \textit{E.
-coli} found that the sequence data indicated several changes to the nomenclature for O-groups \citep{DebRoy2016}.
-There are potentially many other subtype schemes that would not be straightforward to adapt to a WGS approach.
+coli} found that the sequence data indicated several changes to the organization of the O-groups \citep{DebRoy2016}.
+There are potentially other subtype schemes that would show discrepancies between genetic data and subtype grouping.
 A tool that can evaluate the ability of a genotype to predict a subtype would be better equipped for developing the new subtype schemes or updating current schemes for WGS workflows.
@@
 The validation verifies that the phylogeny generated by the training set sequences can be used to predict the subtypes with a high level of accuracy.
-We showed in empirical tests that the farther a subtype is dissociated from the genotype, the lower the F-score computed in the Phylotyper \texttt{add} pipeline.
-In addition to this subtype-level verification, Phylotyper also computes confidence scores for each individual prediction that reflect the rate of subtype change occuring at the input gene's phylogenetic locale.
-If an input sequence falls in region in the phylogenetic tree where the subtype is highly fluid due to evolution or poor subtype-genotype correlation, users would be notified in the condifidence score and in phylogenetic tree visualization output by Phylotyper.
-Phylotyper ability to inform users about the level of agreement between the subtype assignments and genotype makes it uniquely capable to transition historical subtype data to new whole genome sequence-based systems.\vspace*{-10pt}
+We showed in empirical tests that the farther a subtype is dissociated from the input gene's phylogeny, the lower the F-score computed in the Phylotyper \texttt{add} pipeline.
+In addition to this subtype-level verification, Phylotyper also computes confidence scores for each individual prediction that reflect the rate of subtype change occurring at the input gene's phylogenetic locale.
+If an input sequence falls in a region in the phylogenetic tree where the subtype is highly fluid due to evolution or poor subtype-genotype correlation, users would be notified in the confidence score and in the phylogenetic tree visualization output by Phylotyper.
+Phylotyper's ability to inform users about the level of agreement between the subtype assignments and genotype makes it uniquely capable of transitioning historical subtype data to new whole genome sequence-based systems.\vspace*{-10pt}
 

### 2017-06-03 manuscript/response_to_reviewers.txt
EDIT: revisions complete

@@
-response_to_reviewers.txt
+I am pleased to resubmit for publication the revised version of BIOINF-2017-0270 “Phylotyper: In silico predictor of subtypes from gene sequences”.
 
 Firstly, we would like to thank the reviewers for their thoughtful comments.
-We have made subtantual changes to the results and discussion sections as well as to the Phylotyper code that we hope address your suggestions for the Phylotyper manuscript.
-We have responded to each individual comment below:
-
-----------------------------------------
-
-Here are the comments of the reviewers:
+From the recommendations, we have made subtantual changes to the Implementation, Results and Discussion sections.
+We have clarified how Phylotyper generates fewer false positive predictions when encountering novel alleles or subtypes, while maintaing high recall rates (Note: some of our statements were interpreted as Phylotyper being able to create new subtype classes de novo).
+We have expanded the Implementation section to include a description of the assumptions of Phylotyper, and we added additional analyses that test and compare Phylotyper's performance.
+I have addressed each reviewer's comments individually below:
 
@@
 
-*Response:
+Response:
 We agree with the reviewer's suggestion that a working example for new users will enhance the adoption and usability of Phylotyper.
@@
 
-*Response*:
+Response:
 The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor.
@@
 In dealing with missing data, we demonstrate that Phylotyper is significantly more robust than a direct sequence similarity-based predictor.
-It is less likely to generate a type-I error when encountering a subtype not found in the training set.
+It is less likely to generate a Type-I error when encountering a subtype not found in the training set.
 The Phylotyper method estimates the uncertainty associated with a subtype prediction.
@@
 Most often the sequence for a novel or missing subtype will be relatively phylogenetically distant from the other subtype sequences in the tree and its Phylotyper uncertaintity score will reflect this.
-Only in cases where the novel/missing subtype is closely phylogenetically related to an existing subtype in the training set, will Phylotyper generate a false positive prediction.
 This property of Phylotyper makes it more robust when working with incomplete training data.
 
-To enhance the users understanding of the checks that are provided in the build feature and the potential blindspots in creating their own subtype predictors with Phylotyper, we have greatly expanded the methods section and the README provided with the software.
-The added content describes the two checks, how they work and what they detect.
-In manuscript we also provide guidelines about the inputs needed for an effective subtype predictor.
+To enhance the users understanding of the checks that are provided in the build feature and the potential blindspots in creating their own subtype predictors with Phylotyper, we have greatly expanded the Implementation section and the README provided with the software.
+The added content describes the validations in Phylotyper, how they work and what they detect.
+In manuscript we also provide guidelines about selecting sequences for an effective subtype predictor.
 We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the trainging set and used as test input.
@@
 It will allow microbiologist specialists to submit and improve subtype schemes and increase coverage.
-We mention in the manscript and on the README, that users can contact us and submit new or modified subtype schemes.
+We now mention in the manscript and on the README, that users can contact us and submit new or modified subtype schemes.
 We will check the submitted schemes and add them to our Github repository for general use.
 
+----------------------------------------
 Reviewer: 2
@@
 
-*Response*:
+Response:
 We added a high-level description of mechanisms currently in use to perform subtyping, namely phenotype- and DNA-based methods, with specific examples in each category.
 We feel this will give the readers an understanding of the breadth of subtyping methods.
-We also describe the characteristics of the subtypes that have been adapted for in silico prediction using WGS data.
-Our intention was to convey the wide range of subtyping approaches, and how certain subtyping approaches can be recapitulated from WGS data.
+We also describe the characteristics of the subtypes that have been adapted for in silico prediction with WGS data.
+Our intention was to convey the wide range of subtyping approaches, and how certain subtyping approaches have been targeted for prediction with WGS data.
 
@@
 
-*Response*:
+Response:
 A detailed step-by-step account of how to run Phylotyper's automated build pipeline is provided in the Phylotyper package instructions.
 We added a reference and link to these instructions in the Implementation section.
-We also added to the Implmentation section a description of the requirements in Phylotyper for new subtype schemes as well as the checks that the pipeline automatically performs to validate subtypes schemes created by users.
+We also added to the Implementation section a description of the requirements in Phylotyper for new subtype schemes as well as the checks that the pipeline automatically performs to validate subtypes schemes created by users.
 In the Discussion section, we elaborate on the ideal sequence regions that should be targeted by users when developing a WGS-based predictor of subtypes with Phylotyper.
@@
 coli.
-Phylotyper can work with any suitable subtype; identifying a suitable subtype, however, requires knowledge of the subtype and its underlying biological mechanism.
+Phylotyper can work with any suitable subtype.
+Identifying a suitable subtype, however, requires knowledge of the subtype and its underlying biological mechanism.
 To help expand Phylotyper to other pathogens, we added a request in the manuscript and on the Phylotyper website for Microbiologists to send us either subtype schemes or additions/corrections to existing subtype schemes.
@@
 
-*Response*: 
+Response: 
 The Phylotyper assumptions are now outlined in the Implementation section.
@@
 
-*Response*:
+Response:
 Our main goal in the performance assessment was to compare how the prediction methods generalize to an independent dataset.
-There may exists differences in the reference training sets that underpin the tools, and these differences may mask the true performance of the method and we wanted to evalute the predictive capability of the methods themselves.
+There may exists differences in the reference training sets that underpin the tools and these differences may mask the true performance of the method.
+We wanted to evalute the predictive capability of the methods themselves.
 SeroTypeFinder is provided as a web service.
@@
 The intention of using a general BLAST-based approach is demonstrate the benefit of Hidden State Prediction with a phylogenetic tree over approaches that directly match input sequences against a reference database.
-BLAST, as used in our assessment with a percent identity and coverage cutoff, is the basis of the SeroTypeFinder tool.
+BLAST, as used in our assessment with a percent identity and coverage cutoff, is the same method used in the SeroTypeFinder tool.
 EcOH employs a more sophisticated algorithm for matching alleles in order to filter out spurious small reads that might align by chance.
-At its core, it is still matching alleles and does not develop a model of subtype/allele sequence evolution to define boundaries on the subtypes.
+At its core, it is still matching alleles and does not develop a model of subtype allele evolution to define boundaries on the subtypes.
 We feel that our performance comparison to the generalized BLAST-based method does make a valid case for using HSP in subtype prediction.
 We did add a direct comparison of Phylotyper to Vtyper for Stx genes.
-The performance results are summerized in the Results section.
+The performance results are summerized in the Results section and Supplementary Information.
 
@@
 
-*Response*:
+Response:
 The point we were trying make was that more frequently, Phylotyper correctly handles novel subtypes than existing methods.
-In our emperical testing we found Phylotyper makes fewer Type-I errors when encountering a novel allele of an existing subtype or a allele with a entirely new subtype not found in reference training set.
+In our empirical testing we found Phylotyper makes fewer Type-I errors when encountering a novel allele of an existing subtype or a allele with a entirely new subtype not found in reference training set.
 In former case, Phylotyper returns non-significant/undetermined.
 We have clarified this point throughout the manuscript, removing references to Phylotyper identifying novel subtypes and replacing it with the more nuanced description that Phylotyper more frequently properly classifies novel subtypes as a negative result or correctly classifies novel alleles with subtypes in the database.
-To strengthen this statement, we have as the reviewers suggest, remove entire subtypes from the training set and calculated the false positive rate.
+To strengthen this statement, we have as the reviewers suggest, added an analysis where we remove entire subtypes from the training set and calculated the false positive rate.
 Phylotyper was found to have a much lower false positive rate when classifying genes with unknown subtypes.
@@
 However, our goal in this manuscript is to present an alternative genome-based subtype prediction method.
-De novo subtype creation is outside our intended applications for Phylotyper.
+De novo subtype creation is outside our original intended applications for Phylotyper.
 Again, we corrected all references to Phylotyper identifing novel subtypes.
@@
 
-*Response*:
+Response:
 
 We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and genome region that is used for prediction.
-We cite Debroy et al., as an example of how the associated genetic data and subtypes can have discrepancies.
+We cite Debroy et al., as an example of how the associated genetic data and subtype schemes can have discrepancies.
 
@@
 
-*Response*:
-Fixed.
+Response:
+Corrected.
 
@@
 
-*Response*:
+Response:
 Change made.
@@
 
-*Response*:
+Response:
 Sentence replaced.
@@
 
-*Response*:
-Removed the subtype letters at the end of the label; this was redundant information that was also conveyed by the color.
-Added information in the figure description on the meaning of the colors.
-Fixed labels so they fit in the window.
+Response:
+We removed the subtype letters at the end of the label; this was redundant information that was also conveyed by the color.
+Also we added a sentence in the figure description explaining the meaning of the colors.
+Labels are fixed so they fit in the window.
 
@@
 
-*Response*:
-Switched to discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper uses RColorRamp and this palette to generate sufficient colors for all subtypes.
+Response:
+We switched to the discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper uses RColorRamp and this palette to generate sufficient colors for all subtypes.
 
@@
 
-*Response*:
+Response:
 Spacing increased.
@@
 
-*Response*:
-The E. coli serotypes available in Phylotyper comprise two schemes: O- and H-type (here incorrectly counted as one), resulting in five tested schemes. The reference to four has been removed.
+Response:
+The E. coli serotypes available in Phylotyper comprise two schemes: O- and H-type (prevously, incorrectly counted as one), resulting in five tested schemes. The reference to four has been removed.
 
+----------------------------------------
 Reviewer: 3
@@
 
-*Response:
+Response:
 Complete or nearly complete loci sequences are needed for correctly placing the loci in the phylogenetic tree.
@@
 No alleles with less than this coverage to one of the reference set sequences will pass the detection step.
-Users are notified when no loci I have added a description of the requirement for complete loci sequences and the checks we have built into Phylotyper to the Discussion.
+Users are notified when no loci are found.
+I have added a description of the requirement for complete loci sequences to the Implementation section.
 
@@
 
-*Response
+Response
 Phylotyper is significantly more robust than a direct sequence similarity-based predictor in handling novel subtypes or alleles.
-We do not mean that Phylotyper performs unsupervised construction of new subtype schemes (although the underlying method could be adapted for this task), rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e.
+We did not mean that Phylotyper performs unsupervised construction of new subtype schemes (although the underlying method could be adapted for this task), rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e.
 novel subtype).
@@
 
-To clarify this distinction, we have added and modified text.
+To clarify this distinction, we have modified the text and removed all references of "Phylotyper identifying novel alleles", replacing it with the more specific statement.
 We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set to simulate performance when encountering a novel subtype.
-This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach, and more frequently returns a negative result.
+This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach, as it more frequently returns a negative result and not an incorrect subtype prediction.
 
@@
 
-*Response:
+Response:
 The Phylotyper approach is not likely not applicable to core-genome or whole-genome MLST schemes.
@@
 
-Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations where the genotype is only correlated with subtype and not causative.
-If the sequences used to build the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E.
-coli being used to predict O-group serotypes, then the prediction would directly be linked to underlying genome.
-If however unrelated, correlated sequence is used, HGT could disconnect a particular genotype from its subtype.
-We have added to the manuscript recommendations for users building subtype schemes to focus on genomic regions that have a direct biological role in determining the subtype to reduce possibility of impact by HGT.
+Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations where the reference gene sequences are only correlated with subtype and not causative.
+If the sequences used to predict the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E.
+coli being used to predict O-group serotypes, the Phylotyper prediction would directly correspond to the subtype.
+If however unrelated, correlated sequences are used, HGT could separate a particular genotype from its subtype.
+We have made the readers aware of this potential issue and added to the manuscript recommendations for users building subtype schemes to focus on genomic regions that have a direct biological role in determining the subtype to reduce the possibility of impact by HGT.
 
@@
 
-*Response*:
-Fixed
+Response:
+Change made.
 
@@
 
-*Response*:
-We have added descriptions of these assumputions to the mansucript.
+Response:
+We have added descriptions of these assumputions to the Implementaion section.
 

### 2017-06-03 manuscript/phylotyper_revised.tex
EDIT: revisions complete

@@
 
-\title[short Title]{Phylotyper: {\it In silico} predictor of subtypes from gene subtypes}
+\title[short Title]{Phylotyper: {\it In silico} predictor of subtypes from gene sequences}
 \author[Whiteside \textit{et~al}.]{Matthew D. Whiteside\,$^{\text{\sfb 1,}*}$, Chad R. Laing\,$^{\text{\sfb 1}}$ and Victor P.J. Gannon\,$^{\text{\sfb 1,}*}$}
@@
 
-Whole-genome sequencing (WGS) is transforming the public health field by providing an efficient method for surveying bacterial populations.
+WWhole-genome sequencing (WGS) is transforming the public health field by providing an efficient method for surveying bacterial populations.
 The speed, discriminatory power and broad utility of WGS can improve surveillance and outbreak analysis.
@@
 One of the workhorse methods in public health is subtyping.
-Subtyping methods can broadly categorized as phenotype-based or DNA-based.
+Subtyping methods can broadly be categorized as phenotype-based or DNA-based.
 Phenotype-based subtypes are, for example, interrogated by biochemical tests (biotyping), detection of surface antigens (serotyping) or susceptibility to bacteriophage (phagetyping) \citep{Wiedmann2002}.
-Alternatively DNA-based subtyping examines and classifies bacteria based on genetic content.
+Alternatively, DNA-based subtyping examines and classifies bacteria based on genetic content.
 DNA-based subtypes use a variety of methods from PCR to Pulse Field Gel Electrophoresis to DNA sequencing to assign groups to bacteria \citep{Wiedmann2002}.
-As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes.
+As a surveillance tool, subtypes provide a clear cut designation that is typically used to distinguish taxonomic groups and infer phenotypes.
 A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain\citep{Jenkins2015}.
 Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
-To predict from WGS data, these tools have targeted subtypes (either Phenotype-based or DNA-based systems) that can be predicted by sequence variation in specific region or gene in the genome.
-\textit{In silico} subtype prediction is not applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequnce Typing (MLST). 
+To predict from WGS data, these tools have targeted subtypes (either Phenotype-based or DNA-based systems) that can be predicted by sequence variation in a specific region or gene in the genome.
+\textit{In silico} subtype prediction is not applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequence Typing (MLST). 
 No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
-An example of a subtype system that has been adpated for WGS is serotyping in \textit{Escherichia coli}.
+An example of a subtyping system that has been adapted for WGS is serotyping in \textit{Escherichia coli}.
 The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O-antigen group serotype in \citep{Joensen2015,Ingle2016}.
-Another example, is the Shiga toxin subtype (Stx); a DNA-based subtying scheme generated using PCR.
+Another example is the Shiga toxin subtype (Stx); a DNA-based subtyping scheme generated using PCR.
 The tool in \citep{CARRILLO2016} predicts Stx subtype by simulating PCR \textit{in silico}.
@@
 Inputs are in fasta format.
-Hits that do not align over 95\% or have under 90\% sequence identity with reference sequence are discarded.
+Hits that do not align over 95\% or have under 90\% sequence identity with a reference sequence are discarded.
 Users are notified if no loci are found in genome.
@@
 
-Detailed descriptions on how to run Phylotyper and the underlying methods are provided here: \url{https://github.com/superphy/insilico-subtyping}.
+Detailed descriptions on how to run Phylotyper are provided here: \url{https://github.com/superphy/insilico-subtyping}.
 
@@
 
-Phylotyper was designed to be incorporated into a WGS workflow.  
+PPhylotyper was designed to be incorporated into a WGS workflow.  
 The main input into Phylotyper is assembled genome sequences (in fasta format).  
@@
 
-Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli} (\textit{E. coli}): Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
+Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli}: Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
 However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
@@
 Checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
-Each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
-The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from different subtypes.
+Each new subtype is subject to two tests and results are reported to the user.
+The first test checks that the distribution of inter-patristic phylogenetic distances between instances of the same subtype is both smaller and distinct from subtypes that are different.
 This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
@@
 There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are represented in the training set.
-It is important that the user monitor and update schemes as gaps are identified.
-Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
+It is important that the user monitor schemes and update it as gaps are identified.
+Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering an unknown subtype that has no representative in the training set.
 
@@
 The test is part of the \texttt{add} pipeline and is used to estimate the F-score of the subtype scheme. 
-The F-score reflects that predictive capability of the subtype scheme. 
-If the associated phylogeny for training set gene sequences are not correlated with the subtype distribution, this will be reflected in the F-score. 
+The F-score reflects the predictive capability of the subtype scheme. 
+If the associated phylogeny for the training set gene sequences is not correlated with the subtype distribution, this will be reflected in the F-score. 
 To demonstrate this property, we randomly assigned subtypes for increasing proportions of the genes in the training set and computed the F-score with the leave-one-out validation for each proportion level.

### 2017-06-03 manuscript/phylotyper_suppl_info_revised.tex
UPDATE: response to reviewers
ADDED: responses to reviewer 2.
EDIT: revisions complete

--- 2017-05-15 manuscript/phylotyper_suppl_info_revised.tex
@@
 For comparison purposes, we developed a generalized sequence-similarity predictor of subtypes using BLAST.
-Following \citep{Jenkins2015}, a training set of sequences with annotated subtypes is used to build a BLAST database (the training sets are from the the Phylotyper repository).
+A training set of sequences with annotated subtypes is used to build a BLAST database (the training sets are from the the Phylotyper repository).
 To predict a subtype, query sequences are searched against the training BLAST database and the top hit is recorded.
 If the top hit's alignment with the query has at least 60\% coverage with the reference sequence and is above the percent identity cutoff, then the subtype of the top hit is returned as the predicted subtype of the query sequence.
-The cutoff impacts performance in one of two ways: too low and non-equivalent subtype alleles may be matched with the query (e.g.
-type I error), too high and valid subtype allele BLAST hits are filtered out (e.g.
-type II error).
+The cutoff impacts performance in one of two ways: too low and non-equivalent subtype alleles may be matched with the query (i.e.
+Type-I error), too high and valid subtype allele BLAST hits are filtered out (i.e.
+Type-II error).
 To select the ideal cutoff that balances these competing sources of error, a cross-validation simulation was run to estimate false discovery and true positive rates across varying percent identity cutoffs.
@@
 
-The F-score; a binary classifer metric, was recorded individually for each subtype and then averaged to create a single performance metric for this multi-class predictor.
+$TP = True~Positive,~~FP = False~Positive,~~FN = False~Negative$
+
+The F-score; a binary classifier metric, was recorded individually for each subtype and then averaged to create a single performance metric for this multi-class predictor.
 F-score was computed for increasing percent identities, and the lowest percent identity corresponding to the maximum F-score was selected as the cutoff.
-The percent identity cutoff was selected as 0.97.
+The percent identity cutoff was selected as 0.96.
 
@@
 Two separate cross-validation tests were performed to compare the performance of Phylotyper with the BLAST method.
-The comparison to vtyper is a separate analysis, as vtyper does not use a training set and a cross-validation test would not work in this situation.
-A leave-one-out cross-validation analysis was performed where each sequence in the training set is withheld and the remaining sequences are used for form the phylogenetic tree for Phylotyper or are used to build the BLAST database for the BLAST-based approach.
-The predicted result for the test sequence is then compared to the true value.
-
+The comparison to V-typer is a separate analysis, as V-typer does not use a training set and a cross-validation test would not work in this situation.
+A leave-one-out cross-validation analysis was performed.
+For testing Phylotyper, the subtype label of each sequence in the training set was hidden, one at a time and the remaining training set sequences were used to predict the subtype state of the test gene.
+For the BLAST-based subtype predictor, each gene was removed from the BLAST database and used as a query input.
+Positive and negative results are recorded for each cross-validation iteration.
 The second validation iterated over each subtype and removed all genes assigned that subtype from the training set.
 The removed genes or test set was then used as input to identify the level of false positive predictions that occur when the tools are presented with novel subtypes not in the training set.
-
 Precision, recall and the F-score as defined earlier for a multi-classifier test, were computed for the validations.
@@
 The V-typer tool predicts Stx1 or Stx2 subtypes by simulating PCR (i.e.
-mapping primers).
-Due to the unique method in V-typer,  which does not use a training dataset, the cross-validation assessment was adapted to test V-typer.
-Each experimentally-verified Stx gene sequence that makes up Phylotyper training dataset was searched for against the NCBI nt database to retrieve surrounding sequence that may be involved in the PCR detection by V-typer.
-Specifically, regions that have at least 5 kB up- and downstream from the Stx gene were selected for the assessment.
-V-typer was run on each sequence and the results recorded in supplmentary table XXXX.
+mapping primers to the genome).
+Due to the unique method in V-typer, which does not use a training dataset, the cross-validation assessment was not used.
+Instead, a special test set of Stx1 and Stx2 genes was collected and used to evaluate the performance of V-typer relative to Phylotyper.
+Each experimentally-verified Stx gene sequence that makes up the Phylotyper training dataset was searched against the NCBI nt database to retrieve surrounding sequence that may be involved in PCR primer mapping by V-typer.
+Specifically, sequences that have at least 5 kB up- and downstream from the Stx gene were selected for the assessment.
+V-typer was run on each sequence and the results recorded in Supplementary Table S3.
+
+\subsection{Randomizing Subtype Labels}
+
+The Phylotyper method assumes the reference sequence phylogeny aligns with the subtype distribution.
+To validate this assumption, an empirically computed F-score is generated for each new subtype scheme (the F-score is computed through a leave-one-out cross-validation test as described in the Performance Assessment section).
+The F-score indicates the ability of the reference sequence phylogeny to predict the subtype.
+To test how the F-score is impacted by the level of correlation between the subtype distribution and the subtype sequence phylogeny, we ran a simulation where we randomly assigned subtypes to a select proportion of the reference set sequences.
+In the simulation, we randomly selected proportions ranging from 10 to 50 percent of the reference set and then randomly assigned the selected sequence a new subtype (so selected subtypes did not match original subtype label).
+With the modified dataset, F-score was computed through Phylotyper's leave-one-out cross-validation test.
+For each proportion, 100 iterations were performed and metrics averaged.
 
@@
 
+\clearpage
+
+\subsection{Supplementary Table S3}
+
 \begin{minipage}{\linewidth}

### 2017-06-13 phylotyper/R/phylotyper.R
MODIFY: moved legend to margin in tree image

@@ -448,6 +448,10 @@ phylotyper$plotTPP <- function(fit, tree, subtypes) {
 	#   nothing
 	#
 
+	# Setup device
+	margin = 2
+	par(mai=c(1.02,margin,0.82,0.42),xpd=TRUE)
+
 	tree = fit$rerootedTree
 
 	# Save original node assignments in the node.label field
@@ -475,8 +479,11 @@ phylotyper$plotTPP <- function(fit, tree, subtypes) {
 		piecol=cols,
 		cex=0.4)
 	
-	add.simmap.legend(colors=cols,x=0.1*par()$usr[2],
-		y=0.9*par()$usr[4],prompt=FALSE)
+	user.range <- par("usr")[c(2,4)] - par("usr")[c(1,3)]
+	region.pin <- par("pin")[c(2,4)] - par("pin")[c(1,3)]
+	pin.per.xy = region.pin / user.range
+	phylotyper$add.simmap.legend.wide(colors=cols,x=-margin/pin.per.xy[1],
+		y=0.95*par()$usr[4],prompt=FALSE,fsize=1)
 
 }
 
@@ -495,6 +502,7 @@ phylotyper$plotDim <- function(tree, type='fan') {
 	y_scale = 16
 	x_max = 1728
 	y_max = 1728
+	margin = 100
 	res = 72
 
 	# X
@@ -505,7 +513,7 @@ phylotyper$plotDim <- function(tree, type='fan') {
 	if(type != 'fan') {
 		# Y
 		l = length(tree$tip.label)
-		y = l * y_scale
+		y = l * y_scale + margin
 		y = ifelse(y > y_max, y_max, y)
 	} else {
 		y <- x
@@ -514,8 +522,50 @@ phylotyper$plotDim <- function(tree, type='fan') {
 	return(list(x=x,y=y,res=res))
 }
 
+phylotyper$add.simmap.legend.wide<-function(leg=NULL,colors,prompt=TRUE,vertical=TRUE,...){
+	# Adapted from phytools add.simmap.legend written by Liam J. Revell
+	#
+	# Increases size of squares
+
+
+    if(hasArg(shape)) shape<-list(...)$shape
+    else shape<-"square"
+    if(prompt){
+        cat("Click where you want to draw the legend\n")
+        x<-unlist(locator(1))
+        y<-x[2]
+        x<-x[1]
+    } else {
+        if(hasArg(x)) x<-list(...)$x
+        else x<-0
+        if(hasArg(y)) y<-list(...)$y
[…]

### 2017-06-13 phylotyper/phylotyper.py

@@ -131,7 +131,7 @@ class Phylotyper(object):
         # Plot pp on the tree
         if plot_name:
             rcode = '''
-            dim = phylotyper$plotDim(tree)
+            dim = phylotyper$plotDim(tree, 'phylogram')
             graphics.off()
             png(filename="%s", width=dim[['x']],height=dim[['y']],res=dim[['res']])
             do.call(result$plot.function, list(tree=tree, fit=result$result, subtypes=subtypes))
@@ -339,13 +339,3 @@ class Phylotyper(object):
         robjects.r('setwd("%s")' % self.cwd)
 
 
-
-
-        
-
-
-       
-
-
-  
-	

### 2017-06-13 manuscript/response_to_reviewers.txt
MODIFY: Chad's corrections added to document MODIFY: edits to response letter

@@
 The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor.
-To mitigate poor prediction from unsuitable or inaccurate inputs, we included in the build pipeline for the local user-submitted subtypes, a series of checks to automatically test the performance of any submitted subtype scheme.
+To mitigate poor prediction from unsuitable or inaccurate inputs, we included in the build pipeline, checks to automatically test the performance of any submitted subtype scheme.
 These checks are now described in the Implementation section.
 
-While these checks identify major breakdowns in the training set used by Phylotyper, we note there is no check to identify missing data, specifically, missing subtypes or key subtypes gene sequences that define the range of the subtype sequence space.
+While these checks identify major breakdowns in the training set used by Phylotyper, we note there are no checks to identify missing data, specifically, missing subtypes or key gene sequences that define the range of the subtype sequence space.
 For any current general subtype predictor (i.e.
@@
 The Phylotyper method estimates the uncertainty associated with a subtype prediction.
-When test sequences that are phylogenetically distant from known reference sequences or in a region of the phylogenetic tree that is rapidly evolving, the Phylotyper score will reflect the degree of uncertaintity.
-Most often the sequence for a novel or missing subtype will be relatively phylogenetically distant from the other subtype sequences in the tree and its Phylotyper uncertaintity score will reflect this.
-This property of Phylotyper makes it more robust when working with incomplete training data.
 
@@
 The added content describes the validations in Phylotyper, how they work and what they detect.
-In manuscript we also provide guidelines about selecting sequences for an effective subtype predictor.
-We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the trainging set and used as test input.
+In the Discussion section, we also provide guidelines for selecting sequences for an effective subtype predictor.
+We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the training set and used as test input.
 This analysis and results have been added to the Results and Discussions sections.
@@
 We agree that this would be a helpful system.
-It will allow microbiologist specialists to submit and improve subtype schemes and increase coverage.
-We now mention in the manscript and on the README, that users can contact us and submit new or modified subtype schemes.
+We now mention in the manuscript and on the README, that users can contact us and submit new or modified subtype schemes.
 We will check the submitted schemes and add them to our Github repository for general use.
@@
 Response:
-We added a high-level description of mechanisms currently in use to perform subtyping, namely phenotype- and DNA-based methods, with specific examples in each category.
+We added a high-level description of subtyping methods to the Introduction, which relates how subtyping can be categorized as phenotype- or DNA-based, with specific examples in each category.
 We feel this will give the readers an understanding of the breadth of subtyping methods.
@@
 A detailed step-by-step account of how to run Phylotyper's automated build pipeline is provided in the Phylotyper package instructions.
-We added a reference and link to these instructions in the Implementation section.
+We reference and link to these instructions in the Implementation section.
 We also added to the Implementation section a description of the requirements in Phylotyper for new subtype schemes as well as the checks that the pipeline automatically performs to validate subtypes schemes created by users.
-In the Discussion section, we elaborate on the ideal sequence regions that should be targeted by users when developing a WGS-based predictor of subtypes with Phylotyper.
-Also in the Discussion seciton, we name some current examples of subtyping schemes in other pathogens outside of E.
+In the Discussion section, we elaborate on the ideal sequence regions that should be targeted by users when developing a new WGS-based predictor of subtypes with Phylotyper.
+To address the recommendation for providing existing typing schemes, we added in the Discussion section, some current examples of subtyping schemes in other pathogens outside of E.
 coli that can be predicted with Phylotyper.
-We did not systematically go through all main pathogens to identify the subtyping schemes that can predicted with Phylotyper.
+We did not systematically go through all main pathogens to identify the subtyping schemes that can be predicted with Phylotyper.
 As we state in our revised Discussion, the subtype schemes that can be linked biologically to a specific genome sequence region are the best candidates for WGS-based predictors.
-To compile a complete list of these type of subtypes requires domain knowledge in the pathogen (Our lab works with E.
-coli, hence the the initial focus in Phylotyper's resources on E.
+To compile a complete list of these type of subtypes requires domain knowledge in the pathogen (our lab works with E.
+coli, hence the initial focus in Phylotyper's resources on E.
 coli).
@@
 To help expand Phylotyper to other pathogens, we added a request in the manuscript and on the Phylotyper website for Microbiologists to send us either subtype schemes or additions/corrections to existing subtype schemes.
-Subtype schemes that we recieve we will be packaged into the Phylotyper repository for download and general use.
+Subtype schemes that we receive will be packaged into the Phylotyper repository for download and general use.
 
@@
 Our main goal in the performance assessment was to compare how the prediction methods generalize to an independent dataset.
-There may exists differences in the reference training sets that underpin the tools and these differences may mask the true performance of the method.
-We wanted to evalute the predictive capability of the methods themselves.
+There may exist differences in the reference training sets that underpin the tools and these differences may mask the true performance of the method.
+We wanted to evaluate the predictive capability of the methods themselves.
 SeroTypeFinder is provided as a web service.
@@
 A direct comparison of the performance with these tools with Phylotyper is not possible.
-The intention of using a general BLAST-based approach is demonstrate the benefit of Hidden State Prediction with a phylogenetic tree over approaches that directly match input sequences against a reference database.
+The intention of using a general BLAST-based approach is to demonstrate the benefit of Hidden State Prediction with a phylogenetic tree over approaches that directly match input sequences against a reference database.
 BLAST, as used in our assessment with a percent identity and coverage cutoff, is the same method used in the SeroTypeFinder tool.
@@
 We feel that our performance comparison to the generalized BLAST-based method does make a valid case for using HSP in subtype prediction.
-We did add a direct comparison of Phylotyper to Vtyper for Stx genes.
-The performance results are summerized in the Results section and Supplementary Information.
+We did add a direct comparison of Phylotyper to V-typer for Stx genes.
+The V-typer comparison results are summarized in the Results section and provided fully in the Supplementary Information.
 
-To address the reviwers comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis in the manuscript.
-Our aim with the revised descriptions is to provide readers sufficient information give them a complete understanding of the mechanics of the methods.
-Implementation details, such as percent identity cutoff or number of iterations, remain in Supplementary Information.
-This was done to fit with page constraints of the Application Note format.
+To address the comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis in the Results section.
+Our aim with the revised descriptions is to provide readers sufficient information to give them a complete understanding of the mechanics of the methods.
+Implementation details, such as percent identity cutoff or the number of iterations, remain in Supplementary Information.
+This was done to fit with the page constraints of the Application Note format.
 
-As suggested, we also added a theoretical comparison of the tools to the Discussion; the theoretical comparison looks at use cases, potential issues, and additional benefits outside of performance for the tools.
+As suggested, we also added a theoretical comparison of the tools to the Discussion; the theoretical comparison looks at use cases (assembled vs unassembled WGS), potential issues (false positive rates), and additional benefits outside of performance for the tools (prediction confidence values).
 
@@
 Response:
-The point we were trying make was that more frequently, Phylotyper correctly handles novel subtypes than existing methods.
-In our empirical testing we found Phylotyper makes fewer Type-I errors when encountering a novel allele of an existing subtype or a allele with a entirely new subtype not found in reference training set.
-In former case, Phylotyper returns non-significant/undetermined.
-We have clarified this point throughout the manuscript, removing references to Phylotyper identifying novel subtypes and replacing it with the more nuanced description that Phylotyper more frequently properly classifies novel subtypes as a negative result or correctly classifies novel alleles with subtypes in the database.
+The point we were trying to convey was that more frequently, Phylotyper correctly handles novel subtypes than existing methods.
+In our empirical testing, we found Phylotyper makes fewer Type-I errors when encountering a novel allele of an existing subtype or an allele with an entirely new subtype not found in the reference training set.
+In the former case, Phylotyper returns non-significant/undetermined.
+We have clarified this point throughout the manuscript, removing references to Phylotyper identifying novel subtypes and replacing it with the more nuanced description that Phylotyper more frequently correctly classifies novel subtypes as a negative result or correctly classifies novel alleles with subtypes in the database.
 To strengthen this statement, we have as the reviewers suggest, added an analysis where we remove entire subtypes from the training set and calculated the false positive rate.
@@
 
-Phylotyper does not have the ability to create subtype schemes de novo, or to add new subtype classes on the fly.
+Phylotyper does not have the ability to create subtype schemes de novo or to add new subtype classes on the fly.
 We have not explored this potential application of the Phylotyper method.
-It would be an interesting avenue to explore as using a phylogenetic model of subtype evolution to define boundaries for new or existing subtypes might make a nice complement for existing de novo stragies like MLST that assign every new allele a separate subtype.
+It would be an interesting avenue to explore as using a phylogenetic model of subtype evolution to define boundaries for new or existing subtypes might make a nice complement to existing de novo strategies like MLST that assign every new allele a separate subtype.
 However, our goal in this manuscript is to present an alternative genome-based subtype prediction method.
 De novo subtype creation is outside our original intended applications for Phylotyper.
-Again, we corrected all references to Phylotyper identifing novel subtypes.
 
@@
 
-We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and genome region that is used for prediction.
+We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and the genome regions that are used for prediction.
 We cite Debroy et al., as an example of how the associated genetic data and subtype schemes can have discrepancies.
@@
 Response:
-We switched to the discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper uses RColorRamp and this palette to generate sufficient colors for all subtypes.
+We switched to the discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper now uses RColorRamp and this palette to generate sufficient colors for all subtypes.
 
@@
 Response:
-Change made.
+Legend placement has been moved to the margin.
 
@@
 Response:
-Spacing increased.
+Spacing has been increased.
 
@@
 Response:
-Complete or nearly complete loci sequences are needed for correctly placing the loci in the phylogenetic tree.
-To prevent incomplete loci being used, we set coverage cutoffs of 90% in the BLAST detection step.
+Complete or nearly complete loci sequences are needed for correctly placing the allele in the phylogenetic tree.
+To prevent incomplete sequences being used, we set coverage cutoffs of 90% in the BLAST detection step.
 No alleles with less than this coverage to one of the reference set sequences will pass the detection step.
-Users are notified when no loci are found.
-I have added a description of the requirement for complete loci sequences to the Implementation section.
+Users are notified when no gene copies are found.
+I have added a description of the requirement for complete gene sequences to the Implementation section.
 
@@
 Phylotyper is significantly more robust than a direct sequence similarity-based predictor in handling novel subtypes or alleles.
-We did not mean that Phylotyper performs unsupervised construction of new subtype schemes (although the underlying method could be adapted for this task), rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e.
+We did not mean that Phylotyper performs unsupervised construction of new subtype schemes, rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele that is not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e.
 novel subtype).
@@
 To clarify this distinction, we have modified the text and removed all references of "Phylotyper identifying novel alleles", replacing it with the more specific statement.
-We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set to simulate performance when encountering a novel subtype.
-This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach, as it more frequently returns a negative result and not an incorrect subtype prediction.
+We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set.
+This analysis examines the performance of the tools when encountering a novel subtype.
+It showed that Phylotyper is a marked improvement over the sequence-similarity approach, as Phylotyper more frequently returns a true negative result and not an incorrect subtype prediction.
 
@@
 Response:
-The Phylotyper approach is not likely not applicable to core-genome or whole-genome MLST schemes.
-The value that Phylotyper provides is that it can take as input gene sequences and return predictions for subtypes based on phenotype or subtypes that cannot be directly inferred from the gene sequences alone.
+The Phylotyper approach is not designed for core-genome or whole-genome MLST schemes.
+The value that Phylotyper provides is that it can take input gene sequences and return predictions for subtypes that cannot be directly inferred from the gene sequences alone.
 For example, PCR or restriction digestion-based subtyping schemes are an abstraction of the underlying gene sequences and in most cases, cannot be directly inferred from an in silico genome sequence.
@@
 
-Subtypes that are directly identified from the genome sequence do not require Phylotyper's additional prediction step.
-In MLST, the combination each unique allele sequence defines the type.
-Potentially, and this is perhaps what the reviewer is refering to, a specific region of the genome that is smaller than the sequence regions used in an MLST, exists that can be used as a proxy to predict MLST subtype.
-The benefit in this case, is that the entire genome would not need to be sequenced (or multiple alleles interogated) in order to predict MLST type.
-There is precedent for this type of approach.
-The spa and coa genes in Staphylococcus aureus are used as gene markers to infer phylogenetic and population variation.
+Subtypes that are directly identified from the genome sequence do not require imputation from a software tool.
+In MLST, the combination of each unique allele sequence defines the type.
+Potentially, and this is perhaps what the reviewer is referring to, a specific region of the genome that is smaller than the sequence regions used in an MLST, exists that can be used as a proxy to predict MLST subtype.
+The benefit, in this case, is that the entire genome would not need to be sequenced (or multiple alleles interrogated) in order to predict MLST type.
+There is a precedent for this type of approach.
+The spa and coa genes in Staphylococcus aureus are used as gene markers to infer population variation.
 The spa and coa types have noted concordance with MLST typing.
-While this is in interesting potential new application for Phylotyper, it is outside the scope of what we originally designed Phylotyper for.
+While this is an interesting potential new application for Phylotyper, it is outside the scope of what we originally designed Phylotyper for.
 
-Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations where the reference gene sequences are only correlated with subtype and not causative.
+Horizontal gene transfer or recombination could impact the result of Phylotyper in situations where the reference gene sequences used to predict subtype are only correlated with the subtype and not causative of the subtype.
 If the sequences used to predict the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E.
-coli being used to predict O-group serotypes, the Phylotyper prediction would directly correspond to the subtype.
-If however unrelated, correlated sequences are used, HGT could separate a particular genotype from its subtype.
-We have made the readers aware of this potential issue and added to the manuscript recommendations for users building subtype schemes to focus on genomic regions that have a direct biological role in determining the subtype to reduce the possibility of impact by HGT.
+coli being used to predict O-group serotypes, the Phylotyper prediction should directly correspond to the subtype.
+In contrast, if sequences that are correlated but have no biological link are used, HGT could recombine a particular gene allele with a new subtype.
+We have made the readers aware of this potential issue and added to the Discussion section recommendations for users building subtype schemes to focus on genomic regions that have a direct biological role.
 
@@
 Response:
-We have added descriptions of these assumputions to the Implementaion section.
+We have added descriptions of the Phylotyper method assumptions to the Implementation section.
+The revised Implementation section covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.
 

### 2017-06-13 manuscript/phylotyper_revised.tex
MODIFY: Chad's corrections added to document MODIFY: edits to response letter

@@
 Alternatively, DNA-based subtyping examines and classifies bacteria based on genetic content.
-DNA-based subtypes use a variety of methods from PCR to Pulse Field Gel Electrophoresis to DNA sequencing to assign groups to bacteria \citep{Wiedmann2002}.
+DNA-based subtypes use a variety of methods from PCR and Pulse Field Gel Electrophoresis, to DNA sequencing to assign subtype designations to bacterial isolates \citep{Wiedmann2002}.
 As a surveillance tool, subtypes provide a clear cut designation that is typically used to distinguish taxonomic groups and infer phenotypes.
@@
 Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
-To predict from WGS data, these tools have targeted subtypes (either Phenotype-based or DNA-based systems) that can be predicted by sequence variation in a specific region or gene in the genome.
+To predict from WGS data, these tools target subtypes (either Phenotype-based or DNA-based systems) that can be predicted by sequence variation in a specific region or gene in the genome.
 \textit{In silico} subtype prediction is not applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequence Typing (MLST). 
 No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
-An example of a subtyping system that has been adapted for WGS is serotyping in \textit{Escherichia coli}.
+An example of a subtyping system that has been adapted for WGS is serotyping of \textit{Escherichia coli}.
 The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O-antigen group serotype in \citep{Joensen2015,Ingle2016}.
@@
 The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure S2).
-These checks verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
+These checks verify that the phylogenetic grouping provided by the training set is predictive of the subtype.
 There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are represented in the training set.
@@
 This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
-The second validation examined how the predictors perform when tested with a gene that has a subtype, not in the reference set.
-In this validation, each subtype was iterated over and all genes that are assigned the subtype were removed from the training set.
+The second validation examined how the predictors perform when tested with a gene that has a subtype not in the reference set.
+In this validation, each subtype was iterated over and all genes that were assigned the subtype were removed from the training set.
 In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
@@
 This generalized BLAST tool, based on the approach used in \citep{Joensen2015}, assigns a query sequence a subtype when the top BLAST match from an annotated reference database is above a pre-selected percent identity and alignment coverage cutoff.
-Details how the assessment was conducted are available in Supplementary methods.
+Details how the assessment was conducted are available in Supplementary Methods.
 The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
@@
 The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST-hit approach (Supplementary Table S2).
-The BLAST approach also often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper.
+The BLAST approach also had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper.
 Similarly, when entire subtype classes were withheld from the training set, Phylotyper had consistently lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
@@
 The main strategy currently in use by other \textit{in silico} tools for predicting subtypes is to use sequence similarity to annotated gene alleles.
-Query genomes or genes are matched to alleles that are attributed the subtype phenotype or are correlated with the distribution of the subtype.
+Query genomes or genes are matched to alleles that attribute the subtype, or that are correlated with the subtype.
 For example, SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textit{in silico} O-typing and the flagellin genes for H-typing \textit{E.
@@
 When alleles or even subtypes are missing in the reference database, the sequence similarity approach more frequently generates false positive predictions.
-Alternatively, Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
+In contrast, Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
 By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make a Type-I error when encountering a novel subtype or allele. 
@@
 A tool that can evaluate the ability of a genotype to predict a subtype would be better equipped for developing the new subtype schemes or updating current schemes for WGS workflows.
-Phylotyper's \texttt{add} pipeline tests subtype schemes for their predictive accuracy by returning an F-score based on a cross-validation assessment.
+The \texttt{add} pipeline in Phylotyper tests subtype schemes for their predictive accuracy by returning an F-score based on a cross-validation assessment.
 The validation verifies that the phylogeny generated by the training set sequences can be used to predict the subtypes with a high level of accuracy.
@@
 If an input sequence falls in a region in the phylogenetic tree where the subtype is highly fluid due to evolution or poor subtype-genotype correlation, users would be notified in the confidence score and in the phylogenetic tree visualization output by Phylotyper.
-Phylotyper's ability to inform users about the level of agreement between the subtype assignments and genotype makes it uniquely capable of transitioning historical subtype data to new whole genome sequence-based systems.\vspace*{-10pt}
+The ability of Phylotyper to inform users about the level of agreement between the subtype assignments and genotype makes it uniquely capable of transitioning historical subtype data to new whole genome sequence-based systems.\vspace*{-10pt}
 

### 2017-06-16 phylotyper/test/test_seq.py
FIX: allele combinations now working

@@ -18,8 +18,6 @@ class SequenceManipulationTests(unittest.TestCase):
             inputs.append(os.path.join(self.data_dir, i))
         sequences = concat.load(inputs)[0]
 
-	print sequences
-
         correctlen = [ len(ts) == 3 for ts in sequences['genome1'] ]
        
         self.assertTrue(all([
@@ -46,8 +44,8 @@ class SequenceManipulationTests(unittest.TestCase):
 
         self.assertTrue(all([
             testseq1.seq() == 'ATGCTACGAA',
-            testseq2.seq() == 'ATGGTACGAA',
-            testseq3.seq() == 'ATGCTACGAT',
+            testseq2.seq() == 'ATGCTACGAT',
+            testseq3.seq() == 'ATGGTACGAA',
             testseq4.seq() == 'ATGGTACGAT',
         ]))
 

### 2017-06-16 phylotyper/tree/seq.py

@@ -594,6 +594,104 @@ class TypingSequence(object):
         return len(self.typing_sequences)
 
 
+    # def add2(self, loci, sequence, header):
+    #     """Add allele for given loci
+
+    #     Either appends new loci to typing sequence,
+    #     or if this loci is already in sequence, splits
+    #     typing sequence into multiple instances for each
+    #     allele.
+
+    #     Args:
+    #         loci(str): Loci identifier
+    #         sequence(str): bio sequence string
+    #         header(str): Fasta header associated with this loci copy
+
+    #     Returns:
+    #         None
+
+    #     """
+
+    #     alleleid = len(self.alleles)
+    #     newallele = Allele(alleleid, sequence, header)
+    #     self.alleles.append(newallele)
+
+    #     if not isinstance(loci, str):
+    #         loci = str(loci)
+
+    #     print("adding l{} matching {}".format(loci,header))
+    #     print(self.positions)
+    #     if not loci in self.positions:
+    #         # Adding new loci
+
+    #         self.positions[loci] = self.nloci
+
+    #         for ts in self.typing_sequences:
+    #             ts.append(alleleid)
+
+    #         if self.nloci < len(self.combinations):
+    #             self.combinations[self.nloci].append(alleleid)
+    #         else:
+    #             self.combinations.append([alleleid])
+
+    #         self.nloci += 1
+            
+    #     else:
+    #         # Add allele to existing loci
+    #         pos = self.positions[loci]
+    #         newseqs = []
+
+    #         print("pos: {}".format(pos))
+    #         print("this is the combos we work with: {}".format(self.combinations))
+
+    #         # Generate all combinations
+    #         for c in xrange(len(self.combinations)):
+    #             print("combination array: {}".format(c))
+    #             if c == pos:
+
+    #                 if newseqs:
+    #                     for s in newseqs:
+    #                         if c < len(newseqs):
+    #                             copy = self._copy(s)
+    #                             copy[c] = alleleid
+    #                             newseqs.append(copy)
+    #                         else:
+    #                             s.append(alleleid)
+    #                 else: 
+                    
+
+    #             else:
+    #                 for a in self.combinations[c]:
+    #                     print("combination array entry: {}".format(a))
+
+    #                     if 
+    #                     for s in newseqs:
+    #                         if c < len(newseqs):
+    #                             copy = self._copy(s)
+    #                             copy[c] = a
+    #                             newseqs.append(copy)
+    #                         else:
+    #                             s.append(a)
+
+    #                     if c < len(newseqs):
[…]
